### 2025-01-20 index.qmd
Test commit

---
title: "Cox Proportional Hazards Model"
subtitle: "Temp"
author: "Jayme Reed (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

::: callout-important
**Remember:** Your goal is to make your audience understand and care
about your findings. By crafting a compelling story, you can effectively
communicate the value of your data science project.

Carefully read this template since it has instructions and tips to
writing!
:::

## Introduction

The introduction should:

-   Develop a storyline that captures attention and maintains interest.

-   Your audience is your peers

-   Clearly state the problem or question you're addressing.

<!-- -->

-   Introduce why it is relevant needs.

-   Provide an overview of your approach.

Example of writing including citing references:

*This is an introduction to ..... regression, which is a non-parametric
estimator that estimates the conditional expectation of two variables
which is random. The goal of a kernel regression is to discover the
non-linear relationship between two random variables. To discover the
non-linear relationship, kernel estimator or kernel smoothing is the
main method to estimate the curve for non-parametric statistics. In
kernel estimator, weight function is known as kernel function
[@efr2008]. Cite this paper [@bro2014principal]. The GEE [@wang2014].
The PCA [@daffertshofer2004pca]*

*This is my work and I want to add more work...*

## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total))

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References

### 2025-01-20 articleSummaries.qmd
fixed article summaries (forget self-contained)

---
title: "Article Summaries"
format: html
editor: visual
self-contained: true
execute: 
  warning: false
  message: false
bibliography: references.bib
---

[@ziadi2023]This article focused on the length of productive life (LPL in the article) for female Floridian goats. The article discusses the failures of linear models and random regression linear models when it comes to modeling the LPL. As such, Cox proportional hazards model was chosen because it can handle censored records, time-dependent variables, and skewed distribution of characteristics. The does were classified as censored if they were living and uncensored if they were deceased. A preliminary analysis using likelihood ratio tests on a Weibull hazard survival model. This showed that the effects selected all had a significant effect on LPL. The Cox model is then provided for this specific problem. The results were that the age of the first kidding did significant influence LPL. The article provides tables of their results.

[@zapletal2021] The goal of this paper was to examine how explanatory variables influence on an event and if the estimates of the effects of covariates will differ based on the approach that was selected. This paper in particular focused on the Cox proportional hazard and the competing risks approach. The Cox proportional hazards model provides an unspecified baseline hazard function which is beneficial when dealing with a single outcome of interest or one that does not have a specific cause (equations are provided in article). The article describes which competing risks model is the one that will be used in the modeling section. The predictors used by the author were gender, age when the contract began, and region where the person lived. The particular data used in this article was from a Czech Republic insurance company who provide policies for certain illness events until the event occurs. The insured will receive money if the event occurs, but their contract is then complete. Cox models have an assumption of proportional hazards which means the ratios of the hazards need to be independent of time to use the model. The test used was the zph test using Schoenfield residuals (paper is linked in article) and the hypothesis was rejected for age, but not for gender and region. This resulted in the data being filtered to only individuals 18 or older. Two Cox models were developed for each cause of occurrence (models are provided). R packages were survival and cmprsk. The results of this study were that the Cox model showed that if the cause of the event was not considered, the risk based on gender did not vary. The competing risks model showed differently.

### 2025-01-20 slides.qmd
fixed article summaries (forget self-contained)

--- 2025-01-20 index.qmd
@@
 format:
-  html:
-    code-fold: true
+  revealjs
 course: Capstone Projects in Data Science
@@
 
-Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
-to edit)
-
 ::: callout-important
@@
 writing!
+
+More information about `revealjs`:
+<https://quarto.org/docs/reference/formats/presentations/revealjs.html>
 :::
 
-## Introduction
-
-The introduction should:
+## Introduction  {.smaller}
 
@@
 
-<!-- -->
-
 -   Introduce why it is relevant needs.
@@
 
-Example of writing including citing references:
-
-*This is an introduction to ..... regression, which is a non-parametric
-estimator that estimates the conditional expectation of two variables
-which is random. The goal of a kernel regression is to discover the
-non-linear relationship between two random variables. To discover the
-non-linear relationship, kernel estimator or kernel smoothing is the
-main method to estimate the curve for non-parametric statistics. In
-kernel estimator, weight function is known as kernel function
+In kernel estimator, weight function is known as kernel function
 [@efr2008]. Cite this paper [@bro2014principal]. The GEE [@wang2014].
-The PCA [@daffertshofer2004pca]*
-
-*This is my work and I want to add more work...*
+The PCA [@daffertshofer2004pca]\*
 
@@
 
-*The common non-parametric regression model is*
-$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
-of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
-unknown and* $\varepsilon_i$ *some errors. With the help of this
-definition, we can create the estimation for local averaging i.e.*
-$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
-*is near to* $x$*. In other words, this means that we are discovering
-the line through the data points with the help of surrounding data
-points. The estimation formula is printed below [@R-base]:*
-
-$$
-M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
-$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
-Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*
-
-*Another equation:*
-
-$$
-y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
-$$
-
-## Analysis and Results
-
-### Data Exploration and Visualization
+## Data Exploration and Visualization
 
@@
 
+## Data Exploration and Visualization {.smaller}
+
 A study was conducted to determine how...
 
-```{r, warning=FALSE, echo=T, message=FALSE}
+```{r, warning=FALSE, echo=F, message=FALSE}
 # loading packages 
@@
 
-```{r, warning=FALSE, echo=TRUE}
+```{r, warning=FALSE, echo=F}
 # Load Data
-kable(head(murders))
+#kable(head(murders))
 
@@
 
-### Modeling and Results
+## Modeling and Results
 
@@
 
-### Conclusion
+## Conclusion
 

### 2025-01-28 articleSummaries.qmd
Updated article summaries

@@
 The competing risks model showed differently.
+
+[@asghar2024] The goal of the paper is to examine the survival and risk of death associated with Covid-19 in the hospitals of a Pakistani city.
+Specifically, the paper looks at asymptomatic vs symptomatic individuals and focuses on the effects of gender, age, and if the person was symptomatic to assist in determining an individual's survival if they had the virus.
+This paper and research is important as Covid-19 demonstrated that a virus can and will still spread massively throughout the world.
+Determining the chances of survival with the virus would have been beneficial in the height of the pandemic, but is also beneficial now as we are not going to stop living with the virus anytime soon.
+The data the paper uses is from the Directorate General Health Services in KP, Pakistan and contains the hospital attendance records from the first wave of the pandemic, specifically March 2020 to June 2020.
+Individuals age, gender, date of illness, dates of admission and discharge/deceased, and if the individual had symptoms or not when diagnosed with Covid-19.
+The study focuses on individuals who are younger than 100 and did not have missing values.
+They first examined the Kaplan-Meier curve for the survival probability from their data.
+This allowed them to determine that the conventional Cox proportional hazards model may not be best since patients would recover and continue to survive, but the Cox model assumes survival probability will eventually go to zero.
+As such, they also determined a second model using logisitic regression and the Weibull regression model.
+They used a mixture cure model and a non-mixture model for the Weibull regression model.
+Using the Cox proportional hazards model, with three covariates (age, gender, being asymptomatic), age and being asymptomatic were considered to be significantly associated with survival of Covid-19 patients.
+Patients who are older and were symptomatic had a higher risk of death.
+However, the model showed that gender was insignificant.
+For the Weibull mixture cure (which allows for a patients to be cured), older and symptomatic patients were again at an increased risk of death if the person was uncured.
+However, those patients did have a longer survival rate than the uncured asymptomatic patients.
+For the Weibull non-mixture cure model, age and being asymptomatic indicated an increased risk of death in the long-term survival, with the symptomatic indicating an increased risk in the short-term survival.
+Gender for both Weibull models was not considered significant.
+
+[@jiang2024] The goal of the paper is to critique another paper on chest pain incidence and mortality in the US and provide the limits of the model used in that paper.
+Heart disease is the leading cause of death for the US according to the CDC.
+As such, research into the topic, especially attempting to predict the event of death should be thorough and well thought out.
+This paper address concerns for a research topic done.
+The study the paper focuses on found that there is a significant relationship between Triglyceride-Glucose and mortality.
+This was regardless of chest pain occurring within the patients.
+However, there were several limitations using the Cox proportional hazards model as the study did.
+The first limitation brought up is that the Cox model assumes constant hazard ratios over time, but that the effect of Triglyceride-Glucose could vary over time which would violate the assumption of proportional hazards ratios.
+That violation could lead to biased estimates and inaccurate conclusions.
+The Cox model also struggles with time-dependent variables.
+The study took into account body weight and blood glucose levels, both of which can (and will) change throughout time which could impact the true effect of Triglyceride-Glucose.
+The paper provides several other weaknesses of the original study.
+The paper ens by suggesting a better alternative approach would have been to use time-varying covariate models.
+This paper was helpful as it provided data driven examples of why Cox proportional hazards model may not be the best model.
+This will be helpful in guiding the research into datasets to model.

### 2025-02-04 articleSummaries.qmd
Updated articles

@@
 This will be helpful in guiding the research into datasets to model.
+
+[@wang2025] The goal of the paper is to try and find a connection between the salt intake of an individual and if they will develop anxiety or depression later in life.
+Depression and anxiety are extremely prevalent in our society today and we are prone to eating salt with every meal, especially in the United States.
+Determining an association between salt and mental health will allow professionals an area to ask their patients to assist them with their mental health.
+In addition, salt is known to be harmful in large quantities to our health overall so this would be another area to encourage a decrease in salt intake.
+The group took almost half a million participants who, at the start of the study, had never been diagnosed with depression or anxiety and had no missing data.
+The participants were asked if they added salt to their foods and to note how often they do and major changes to their diet, if any.
+They first found the mean with SD and used the chi-square tests for categorical variables and general linear models for continuous variables to check the proportions between their participants and their salt intake level.
+They then used the Cox proportional hazards for the association between an individual's salt intake and the risk of developing depression or anxiety.
+They tested the assumption using the Kaplan-Meier method and Schoenfeld residuals and found no violation.
+They looked at the analyses with hazard ratios and did multiplicative interaction assessments in the Cox models.
+They also conducted several sensitivity tests to make adjustments.
+The results showed that hazard of depression did increase with the increased intake of salt.
+A similar result was found for anxiety and increased salt.
+The limitations of this particular study included the self-reporting of a person's salt intake, their sociodemographic factors, and their lifestyle factors.
+Primarily, the limitations for this paper were based on possible bias in their data.
+
+[@zhang2025] The goal of the paper is to determine an association with postpartum depression and having two or more chronic diseases (called multimorbidity).
+This is important because many women suffer from postpartum depression and this can have lasting affects on their lives.
+In addition, understanding the long term implications of postpartum depression could help in preparing women for what might happen with childbirth and the mental impact it can have.
+The data was provided by self reporting.
+Women noted if they had a depressive episode in the months after giving birth or if they were told they had postpartum depression.
+They selected 36 chronic diseases to focus on and used the prevalence of multimorbidity, the incidence of multimorbidity, the number of new-onset chronic diseases, the cumulative rate of chronic diseases, and the multimorbidity patterns as the specific outcomes they were measuring.
+They provided the list of covariates within the paper.
+They first used binomial logistic regression models and quasi-Poisson mixed effects models to assist in gaining the initial findings of the study and create a baseline.
+They used Cox proportional hazard models to evaluate the associations of postpartum depression with having developed one or more of their chronic diseases.
+They checked the proportional hazard assumption using Schoenfeld's residual methods and no violations were found.
+They then used multinomial logistic regression models to study the association between postpartum depression and the five multimorbidity patterns they were able to determine.
+Patients who had a history of postpartum depression were shown to have a higher rate prevalence and incidence of one or more of the chronic diseases.
+They also discovered that those who had multimorbidity as their baseline prior to the study had a higher rate of prevalence and incidence as well.
+The limitations of this study are similar to those in the salt and anxiety study as their data came from the same place in both studies.
+As such, the data was open to bias and issues with self-reporting from the patients.
+The data also limited the representation of women due to various factors.
+
+-   What is the goal of the paper?
+
+-   Why is it important?
+
+-   How is it solved? - methods
+
+-   Results/limitations, if any.

### 2025-02-10 index.qmd
fixed article summaries (forget self-contained)
Add intro draft to index.qmd

@@
 subtitle: "Temp"
-author: "Jayme Reed (Advisor: Dr. Cohen)"
+author: "Jayme Reed & Brad Paton (Advisor: Dr. Cohen)"
 date: '`r Sys.Date()`'
@@
 to edit)
-
-::: callout-important
-**Remember:** Your goal is to make your audience understand and care
-about your findings. By crafting a compelling story, you can effectively
-communicate the value of your data science project.
-
-Carefully read this template since it has instructions and tips to
-writing!
-:::
 
@@
 kernel estimator, weight function is known as kernel function
-[@efr2008]. Cite this paper [@bro2014principal]. The GEE [@wang2014].
-The PCA [@daffertshofer2004pca]*
+[@efr2008]. Cite this paper [@bro2014principal]. The goat study*
+[@ziadi2023]*. The insurance[@zapletal2021]*
 
 *This is my work and I want to add more work...*
+
+## Introduction (draft)
+
+Analyzing survival data is challenging for traditional machine learning models because they struggle to capture the time-to-event nature of the data.Survival data typically contains challenges difficult to account for because the datasets may be “censored” meaning that the event hasn’t had the opportunity to occur yet by the end of the study period.
+Also, traditional methods aren’t able to effectively model time-to-event data because they treat time as fixed values opposed to treating the values as time leading to less accurate results.
+
+Cox proportional hazards models are commonly used to interpret, analyze, and examine survival data.
+These models specialize in time-to-event predictions and can account for multiple variables and their effects on survival.
+Cox proportional hazards models are commonly used in the health field while predicting patient survivability based on symptoms and medical history, the financial field to predict bank failure, and mechanical studies to predict failure of machines.
+Cox proportional hazards models are popular due to their ability to incorporate multiple covariates and they have the advantage of being semi-parametric which allows more flexibility as compared to other models.
+
+In this paper, we will be examining the prediction of remaining useful life (RUL)  for NASA turbofan jet engines.
+The lifespan of a jet engine is important as they cost thousands of dollars to maintain and produce.
+The ability to predict the remaining useful lifespan of these engines will allow for better judgments on the finances needed to maintain or repair them as well as alert teams that an engine may not have the RUL for a mission they would like to use it for.
+
+We will first generate the summary statistics for this dataset as well as run linear and logistic regression tests.
+We will then test our assumptions which include the proportional hazards assumption, the linearity between covariates and the log hazard, that each data is independent of each other and there is no omitted data.
+A model will be generated using the survival package from R.
+We will then test the model and make adjustments as needed before generating our predictions.
 

### 2025-02-10 articleSummaries.qmd
Updated Intro with html

@@
 ---
-title: "Article Summaries"
+title: "Literature Review - Reed"
 format: html

### 2025-02-15 index.qmd
Updated Intro with html
Added link to article summaries
Updated Intro Take 2

@@
 ---
-title: "Cox Proportional Hazards Model"
-subtitle: "Temp"
+title: "Predicting Remaining Useful Life of NASA Aircraft Engines "
+subtitle: "Using Cox Proportional Hazards Model"
 author: "Jayme Reed & Brad Paton (Advisor: Dr. Cohen)"
@@
 
-## Introduction
+[Literature](articleSummaries.html)
 
-The introduction should:
+## Introduction (take 2)
+
+As the COVID-19 pandemic demonstrated, the ability to predict the
+survival of a patient based on their symptoms and medical history is a
+major benefit as society deals with health crisis like pandemics. Those
+predictions would assist medical practitioners in triage and allow
+researchers the ability to generate predictions on the individuals who
+are at the greatest risk of losing their lives due to an illness. While
+the medical world is the most obvious use of survival predictions, the
+same methods can be applied to the predicting bank failure, the
+remaining useful life of machines, machinery maintenance timelines, and
+insurance likelihood of payouts. There are various survival methods that
+can be examined, but this paper will focus on Cox proportional hazards
+model. (*don't like)*
+
+Cox proportional hazards (CPH) model is a statistical regression model
+for survival modeling that specializes in time-to-event predictions
+using using single or multiple covariates and the effects they have on
+survival. Traditional regression models struggle to capture the
+time-to-event nature of data and treat the time variable as a fixed
+value instead of continuous variable. CPH models are able to overcome
+those obstacles and are capable of dealing with survival data that
+contains censored (events that have not occurred yet in the study
+period) data as well. The models also have the advantage of being
+semi-parametric which allows more flexibility as compared to other
+models.
+
+## Introduction (draft)
+
+Analyzing survival data is challenging for traditional machine learning
+models because they struggle to capture the time-to-event nature of the
+data. Survival data typically contains challenges difficult to account
+for because the datasets may be “censored” meaning that the event hasn’t
+had the opportunity to occur yet by the end of the study period. Also,
+traditional methods aren’t able to effectively model time-to-event data
+because they treat time as fixed values opposed to treating the values
+as time leading to less accurate results.
+
+Cox proportional hazards models are commonly used to interpret, analyze,
+and examine survival data. These models specialize in time-to-event
+predictions and can account for multiple variables and their effects on
+survival. Cox proportional hazards models are commonly used in the
+health field while predicting patient survivability based on symptoms
+and medical history, the financial field to predict bank failure, and
+mechanical studies to predict failure of machines. Cox proportional
+hazards models are popular due to their ability to incorporate multiple
+covariates and they have the advantage of being semi-parametric which
+allows more flexibility as compared to other models.
+
+In this paper, we will be examining the prediction of remaining useful
+life (RUL) for NASA turbofan jet engines. The lifespan of a jet engine
+is important as they cost thousands of dollars to maintain and produce.
+The ability to predict the remaining useful lifespan of these engines
+will allow for better judgments on the finances needed to maintain or
+repair them as well as alert teams that an engine may not have the RUL
+for a mission they would like to use it for.
+
+We will first generate the summary statistics for this dataset as well
+as run linear and logistic regression tests. We will then test our
+assumptions which include the proportional hazards assumption, the
+linearity between covariates and the log hazard, that each data is
+independent of each other and there is no omitted data. A model will be
+generated using the survival package from R. We will then test the model
+and make adjustments as needed before generating our predictions.
+
+*Intro guidelines:* The introduction should:
 
@@
 *This is my work and I want to add more work...*
-
-## Introduction (draft)
-
-Analyzing survival data is challenging for traditional machine learning models because they struggle to capture the time-to-event nature of the data.Survival data typically contains challenges difficult to account for because the datasets may be “censored” meaning that the event hasn’t had the opportunity to occur yet by the end of the study period.
-Also, traditional methods aren’t able to effectively model time-to-event data because they treat time as fixed values opposed to treating the values as time leading to less accurate results.
-
-Cox proportional hazards models are commonly used to interpret, analyze, and examine survival data.
-These models specialize in time-to-event predictions and can account for multiple variables and their effects on survival.
-Cox proportional hazards models are commonly used in the health field while predicting patient survivability based on symptoms and medical history, the financial field to predict bank failure, and mechanical studies to predict failure of machines.
-Cox proportional hazards models are popular due to their ability to incorporate multiple covariates and they have the advantage of being semi-parametric which allows more flexibility as compared to other models.
-
-In this paper, we will be examining the prediction of remaining useful life (RUL)  for NASA turbofan jet engines.
-The lifespan of a jet engine is important as they cost thousands of dollars to maintain and produce.
-The ability to predict the remaining useful lifespan of these engines will allow for better judgments on the finances needed to maintain or repair them as well as alert teams that an engine may not have the RUL for a mission they would like to use it for.
-
-We will first generate the summary statistics for this dataset as well as run linear and logistic regression tests.
-We will then test our assumptions which include the proportional hazards assumption, the linearity between covariates and the log hazard, that each data is independent of each other and there is no omitted data.
-A model will be generated using the survival package from R.
-We will then test the model and make adjustments as needed before generating our predictions.
 

### 2025-02-15 index.qmd
Update Intro and references

@@
 
-Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
-to edit)
+Slides: [slides.html](slides.html){target="_blank"}
 
-[Literature](articleSummaries.html)
+Literature Comments: [Literature.html](articleSummaries.html)
 
@@
 insurance likelihood of payouts. There are various survival methods that
-can be examined, but this paper will focus on Cox proportional hazards
-model. (*don't like)*
+can be examined, including Cox proportional hazards model which will be
+examined in this paper.
 
@@
 those obstacles and are capable of dealing with survival data that
-contains censored (events that have not occurred yet in the study
-period) data as well. The models also have the advantage of being
-semi-parametric which allows more flexibility as compared to other
-models.
+contains censored data as well. Censored data is when the information
+about an individual in a study is only known for a certain period of
+time, such as the length of the study [@klein2005, 1]. The models also
+have the advantage of being semi-parametric which allows more
+flexibility as compared to other models.
+
+As mentioned, CPH is used in various fields including the health
+industry, but the usage of the model does not come without challenges. A
+COVID-19 study done using data from a Pakistani hospital noted that the
+model was limited due to the recovery possibility of COVID-19 patients
+as CPH assumes as time goes on, the survival probability will approach
+zero with no survivors. Though CPH was not the best model for this
+study, the results of the model were that individuals who were
+asymptomatic and young had a higher chance of surviving the virus
+[@asghar2024]. In 2024, a study was done in the US on the relationship
+between chest pain occurring and mortality of the person experiencing
+chest pain using CPH. A response to that study was written that provided
+the limitations of CPH with the biggest concern being the assumption of
+constant hazards ratio that CPH relies on for the model. In the heart
+study, the effect of Triglyceride-Glucose was assumed to be a constant,
+but is actually a value that fluctuates over time along with an
+individual's body weight and blood glucose levels. As such, the model
+would struggle to predict correctly the effect of Triglyceride-Glucose
+given the violation of the constant hazards ratio [@jiang2024].
+
+An additional limitation of CPH comes from the data in the covariaties
+selected and how can they can become biased. Two studies done using UK
+health data demonstrated how bias can impact the values for the
+covariates when relying on patient submitted information. The first was
+a study on the connection between salt intake and the development of
+anxiety or depression after 14.5 years [@wang2025] and the second was a
+study on the association between postpartum depression and the
+development of two or more chronic diseases [@zhang2025]. Both of those
+studies relied on patient submitted health data and survey responses
+that were supplemented with death records and other accessible health
+records thus creating a possible bias in their data which could
+negatively influence their model as their covariates may not have
+accurately reflected the true data. These articles provide examples of
+the care that needs to be taken when working with CPH models.
+
+While health care is the dominate field in which CPH is used, a study in
+2021 provides an example of using CPH in insurance fields. An insurance
+company in the Czech Republic provides policies for certain illness
+events until the event occurs when the insured will receive money and
+their contract is complete. To assist the company with predicting if an
+individual will have one of the illness events occur, a model using CPH
+was generated using gender, age, and region as the
+covariates[@zapletal2021]. A study in 2023 focused on the length of
+productive life of female Floridian goats to determine the factors that
+would impact the longevity of the goats. CPH was used to determine that
+the age of first kidding and the specific herd impacts the length of
+productive life of the goats [@ziadi2023]. Both of these studies provide
+examples of the reaches of CPH models outside the health industry.
+
+Continuing with using CPH outside of health data, we will study the
+remaining useful life of a NASA aircraft engine using data provided by
+NASA for the study on propagation modeling [@saxena2008]. From the
+literature review, we have noted that we will need to be concerned with
+the constant hazards ratio within our data and determine how we will
+censor our data.
 
@@
 
-*Intro guidelines:* The introduction should:
-
--   Develop a storyline that captures attention and maintains interest.
-
--   Your audience is your peers
-
--   Clearly state the problem or question you're addressing.
-
 <!-- -->
-
--   Introduce why it is relevant needs.
-
--   Provide an overview of your approach.
-
-Example of writing including citing references:
-
-*This is an introduction to ..... regression, which is a non-parametric
-estimator that estimates the conditional expectation of two variables
-which is random. The goal of a kernel regression is to discover the
-non-linear relationship between two random variables. To discover the
-non-linear relationship, kernel estimator or kernel smoothing is the
-main method to estimate the curve for non-parametric statistics. In
-kernel estimator, weight function is known as kernel function
-[@efr2008]. Cite this paper [@bro2014principal]. The goat study*
-[@ziadi2023]*. The insurance[@zapletal2021]*
-
-*This is my work and I want to add more work...*
 

### 2025-02-17 articleSummaries.qmd
Updated Intro Take 2
Brad edits 2/16

@@
 
--   What is the goal of the paper?
+[@Ming-Chiang2014]
 
--   Why is it important?
+https://citeseerx.ist.psu.edu/document?repid=rep1&type=pdf&doi=8e45bb12b4bf296f002977caf11e969d5cc90a02
 
--   How is it solved? - methods
+This article uses Cox Proportional Hazards model to predict bankruptcies in companies in the Taiwanese stock market.
+The model used uses 12 variables from companies found between 2003 and 2009 to classify companies into the categories of distressed vs non-distressed to indicate if they are at risk of bankruptcy.
+Each variable had reasoning and/or sources to explain the reasoning behind using them in the model along with statistical justification.
+The resulting model had an accuracy of 88%.
 
--   Results/limitations, if any.
+[@Bustan2018]
+
+https://iopscience.iop.org/article/10.1088/1742-6596/1028/1/012230/pdf
+
+This article describes a Cox Proportional Hazards Model to predict the probability of survival in breast cancer patients.
+The model uses six variables to make predictions and provides an explanation of the risk factors that are the most heavily weighted when deciding a patient's outlook.
+The article includes a table showing the survival probabilities of patients based on the values of their corresponding covariates.
+The article concludes the most important factor in surviving breast cancer is to catch it in as early of a stage as possible and recommends more outreach programs to encourage individuals to get themselves examined by a doctor regularly.
+
+[@Kuitunen2021]
+
+This article explores the proportionality assumption of cox proportional hazard models on arthroplasty research datasets.
+It accounts for over 1100 studies.
+The article concludes that the proportionality assumption cannot be ignored, as without it the results of the models created are unreliable.
+It is critical for the proportional hazards assumption to be tested properly, or inaccurate predictions will be made,
+
+[@Smith2003]
+
+In this article, Cox Proportional Hazard modeling was used for modeling time until event while accounting for other variables.
+The author explains how Cox PH models work mathematically and the strenths of Cox PH models.
+Cox PH models are able to accomodate incomplete or "censored" data which is prevalent in survival data.
+In many clinical trials, a patient death may not be observed because the trial period will end not allowing enough time for the event to occur.
+Health datasets were analyzed with hospitalization and demographic data to predict hospitalization time after inhaling oil well fire smoke.
+The model creates a probability for each individual's chances of being hospitalized.
+
+[@Abeysekera2009]
+
+This article summarizes the use of Shoenfeld's global test for testing assumptions of Cox PH models.
+It mentions that other methods for checking assumptions have an aspect of subjectivity to them, but this method shines because it does not.
+The proportional hazards assumption is the most important aspect of a dataset when trying to analyze data with a Cox PH model, and the Schoenfeld's global test is an effective method to test the assumption.
+The article includes visualizations and ways to interpret the visuals.
+The example provided in the paper shows the conclusion of which covariates were shown to be in violation of the assumption in a better way than typical methods.
+The article recommends Shoenfeld's global test as the ideal method for testing assumptions.
+
+[@Seung2023]
+
+https://www.elifecycle.org/archive/view_article?pid=lc-3-0-8
+
+This article compares Kaplan-Meier survival analysis with Cox Proportional Hazards models.
+It summarizes how effective and necessary the models are in the health industry, and how important it is to pass the proportional hazards assumption to ensure an accurate model.
+It demonstrates the reason Cox models are so widely used over Kaplan-Meier because they're able to accommodate multiple covariates into predictions.
+The paper also explains how to handle datasets that break the assumptions of Cox PH models.
+It also goes over the time dependent regression models that can be used when the assumptions are broken. The author concludes that each model mentioned has its merits, but Cox PH models are the most popular models because of their flexibility and ability to include multiple covariates.

### 2025-02-17 index.qmd
Brad edits 2/16

@@
 ---
-title: "Predicting Remaining Useful Life of NASA Aircraft Engines "
+title: "Predicting Remaining Useful Life of NASA Aircraft Engines"
 subtitle: "Using Cox Proportional Hazards Model"
@@
 Literature Comments: [Literature.html](articleSummaries.html)
+
+## Introduction (take 3)
+
+As the COVID-19 pandemic demonstrated, the ability to predict the
+survival of a patient based on their symptoms and medical history is a
+major benefit as society deals with health crises such as pandemics. Those
+predictions would assist medical practitioners in triage and allow
+researchers the ability to generate predictions on the individuals who
+are at the greatest risk of losing their lives due to an illness. While
+the medical world is the most obvious use of survival predictions, the
+same methods can be applied to the predicting bank failure, the
+remaining useful life of machines, machinery maintenance timelines, and
+insurance likelihood of payouts. There are various survival methods that
+can be examined, including Cox proportional hazards model which will be
+examined in this paper.
+
+Cox proportional hazards (CPH) models are statistical regression models specializing in
+modeling time-to-event predictions with survival data. Survival data is data with a value
+for time and an event. For example, in a study investigating the survival likelihood of
+mechanical equipment time would begin when the equipment begins functioning and would
+conclude in the event the equipment stops functioning. [@Smith2003] CPH models can use
+single or multiple covariates as a means of creating predictions until failure and are
+capable of dealing with survival data that contains censored data as well.[@Seung2023]
+Censored data is when the information about an individual in a study is only known for a
+certain period of time, such as the length of the study [@klein2005, 1]. The models also
+have the advantage of being semi-parametric which allows more flexibility as compared to
+other models. [@Ming-Chiang2014]
+
+As mentioned, CPH is used in various fields including the health
+industry, but the usage of the model does not come without challenges. A
+COVID-19 study done using data from a Pakistani hospital noted that the
+model was limited due to the recovery possibility of COVID-19 patients
+as CPH assumes as time goes on, the survival probability will approach
+zero with no survivors. Though CPH was not the best model for this
+study, the results of the model were that individuals who were
+asymptomatic and young had a higher chance of surviving the virus
+[@asghar2024]. In 2024, a study was done in the US on the relationship
+between chest pain occurring and mortality of the person experiencing
+chest pain using CPH. A response to that study was written that provided
+the limitations of CPH with the biggest concern being the assumption of
+constant hazards ratio that CPH relies on for the model. In the heart
+study, the effect of Triglyceride-Glucose was assumed to be a constant,
+but is actually a value that fluctuates over time along with an
+individual's body weight and blood glucose levels. As such, the model
+would struggle to predict correctly the effect of Triglyceride-Glucose
+given the violation of the constant hazards ratio [@jiang2024].
+
+An additional limitation of CPH comes from the data in the covariates
+selected and how can they can become biased. Two studies done using UK
+health data demonstrated how bias can impact the values for the
+covariates when relying on patient submitted information. The first was
+a study on the connection between salt intake and the development of
+anxiety or depression after 14.5 years [@wang2025] and the second was a
+study on the association between postpartum depression and the
+development of two or more chronic diseases [@zhang2025]. Both of those
+studies relied on patient submitted health data and survey responses
+that were supplemented with death records and other accessible health
+records thus creating a possible bias in their data which could
+negatively influence their model as their covariates may not have
+accurately reflected the true data. These articles provide examples of
+the care that needs to be taken when working with CPH models.
+
+While health care is the dominate field in which CPH is used, a study in
+2021 provides an example of using CPH in insurance fields. An insurance
+company in the Czech Republic provides policies for certain illness
+events until the event occurs when the insured will receive money and
+their contract is complete. To assist the company with predicting if an
+individual will have one of the illness events occur, a model using CPH
+was generated using gender, age, and region as the
+covariates[@zapletal2021]. A study in 2023 focused on the length of
+productive life of female Floridian goats to determine the factors that
+would impact the longevity of the goats. CPH was used to determine that
+the age of first kidding and the specific herd impacts the length of
+productive life of the goats [@ziadi2023]. Both of these studies provide
+examples of the reaches of CPH models outside the health industry.
+
+Continuing with using CPH outside of health data, we will study the
+remaining useful life of a NASA aircraft engine using data provided by
+NASA for the study on propagation modeling [@saxena2008]. From the
+literature review, we have noted that we will need to be concerned with
+the constant hazards ratio within our data and determine how we will
+accommodate censored data when creating our model.
 

### 2025-02-17 index.qmd
2/17 updates

@@
   html:
+    theme: spacelab
+    toc: true
+    toc-depth: 3
     code-fold: true
@@
 modeling time-to-event predictions with survival data. Survival data is data with a value
-for time and an event. For example, in a study investigating the survival likelihood of
+for time and an event.[@Abeysekera2009] For example, in a study investigating the survival likelihood of
 mechanical equipment time would begin when the equipment begins functioning and would

### 2025-02-18 articleSummaries.qmd
2/17 updates
Intro Finalization

@@
 ---
-title: "Literature Review - Reed"
+title: "Literature Review"
 format: html
@@
 ---
+
+### Article Summaries (Reed)
 
@@
 
+### Article Summaries (Paton)
+
 [@Ming-Chiang2014]
@@
 
+https://link.springer.com/article/10.1186/s12891-021-04379-2
+
 This article explores the proportionality assumption of cox proportional hazard models on arthroplasty research datasets.
@@
 [@Abeysekera2009]
+
+https://www.researchgate.net/publication/238483310_Use_of_Schoenfeld's_global_test_to_test_the_proportional_hazards_assumption_in_the_Cox_proportional_hazards_model_An_application_to_a_clinical_study
 

### 2025-02-18 index.qmd
Intro Finalization

@@
 
-## Introduction (take 3)
+## Introduction
 
@@
 survival of a patient based on their symptoms and medical history is a
-major benefit as society deals with health crises such as pandemics. Those
-predictions would assist medical practitioners in triage and allow
+major benefit as society deals with health crises such as pandemics.
+Those predictions would assist medical practitioners in triage and allow
 researchers the ability to generate predictions on the individuals who
@@
 
-Cox proportional hazards (CPH) models are statistical regression models specializing in
-modeling time-to-event predictions with survival data. Survival data is data with a value
-for time and an event.[@Abeysekera2009] For example, in a study investigating the survival likelihood of
-mechanical equipment time would begin when the equipment begins functioning and would
-conclude in the event the equipment stops functioning. [@Smith2003] CPH models can use
-single or multiple covariates as a means of creating predictions until failure and are
-capable of dealing with survival data that contains censored data as well.[@Seung2023]
-Censored data is when the information about an individual in a study is only known for a
-certain period of time, such as the length of the study [@klein2005, 1]. The models also
-have the advantage of being semi-parametric which allows more flexibility as compared to
-other models. [@Ming-Chiang2014]
+Cox proportional hazards (CPH) models are statistical regression models
+specializing in modeling time-to-event predictions with survival data.
+Survival data is data with a value for time and an event
+[@Abeysekera2009]. For example, in a study investigating the survival
+likelihood of mechanical equipment, the time value would begin when the
+equipment begins functioning and would conclude in the event the
+equipment stops functioning [@Smith2003]. CPH models can use single or
+multiple covariates as a means of creating predictions until failure and
+are capable of dealing with survival data that contains censored data as
+well [@Seung2023]. Censored data is when the information about an
+individual in a study is only known for a certain period of time, such
+as the length of the study [@klein2005, 1]. The models also have the
+advantage of being semi-parametric which allows more flexibility as
+compared to other models [@Ming-Chiang2014].
 
-As mentioned, CPH is used in various fields including the health
+As mentioned, CPH is used in various fields, primarily the health
 industry, but the usage of the model does not come without challenges. A
@@
 accommodate censored data when creating our model.
-
-## Introduction (take 2)
-
-As the COVID-19 pandemic demonstrated, the ability to predict the
-survival of a patient based on their symptoms and medical history is a
-major benefit as society deals with health crisis like pandemics. Those
-predictions would assist medical practitioners in triage and allow
-researchers the ability to generate predictions on the individuals who
-are at the greatest risk of losing their lives due to an illness. While
-the medical world is the most obvious use of survival predictions, the
-same methods can be applied to the predicting bank failure, the
-remaining useful life of machines, machinery maintenance timelines, and
-insurance likelihood of payouts. There are various survival methods that
-can be examined, including Cox proportional hazards model which will be
-examined in this paper.
-
-Cox proportional hazards (CPH) model is a statistical regression model
-for survival modeling that specializes in time-to-event predictions
-using using single or multiple covariates and the effects they have on
-survival. Traditional regression models struggle to capture the
-time-to-event nature of data and treat the time variable as a fixed
-value instead of continuous variable. CPH models are able to overcome
-those obstacles and are capable of dealing with survival data that
-contains censored data as well. Censored data is when the information
-about an individual in a study is only known for a certain period of
-time, such as the length of the study [@klein2005, 1]. The models also
-have the advantage of being semi-parametric which allows more
-flexibility as compared to other models.
-
-As mentioned, CPH is used in various fields including the health
-industry, but the usage of the model does not come without challenges. A
-COVID-19 study done using data from a Pakistani hospital noted that the
-model was limited due to the recovery possibility of COVID-19 patients
-as CPH assumes as time goes on, the survival probability will approach
-zero with no survivors. Though CPH was not the best model for this
-study, the results of the model were that individuals who were
-asymptomatic and young had a higher chance of surviving the virus
-[@asghar2024]. In 2024, a study was done in the US on the relationship
-between chest pain occurring and mortality of the person experiencing
-chest pain using CPH. A response to that study was written that provided
-the limitations of CPH with the biggest concern being the assumption of
-constant hazards ratio that CPH relies on for the model. In the heart
-study, the effect of Triglyceride-Glucose was assumed to be a constant,
-but is actually a value that fluctuates over time along with an
-individual's body weight and blood glucose levels. As such, the model
-would struggle to predict correctly the effect of Triglyceride-Glucose
-given the violation of the constant hazards ratio [@jiang2024].
-
-An additional limitation of CPH comes from the data in the covariaties
-selected and how can they can become biased. Two studies done using UK
-health data demonstrated how bias can impact the values for the
-covariates when relying on patient submitted information. The first was
-a study on the connection between salt intake and the development of
-anxiety or depression after 14.5 years [@wang2025] and the second was a
-study on the association between postpartum depression and the
-development of two or more chronic diseases [@zhang2025]. Both of those
-studies relied on patient submitted health data and survey responses
-that were supplemented with death records and other accessible health
-records thus creating a possible bias in their data which could
-negatively influence their model as their covariates may not have
-accurately reflected the true data. These articles provide examples of
-the care that needs to be taken when working with CPH models.
-
-While health care is the dominate field in which CPH is used, a study in
-2021 provides an example of using CPH in insurance fields. An insurance
-company in the Czech Republic provides policies for certain illness
-events until the event occurs when the insured will receive money and
-their contract is complete. To assist the company with predicting if an
-individual will have one of the illness events occur, a model using CPH
-was generated using gender, age, and region as the
-covariates[@zapletal2021]. A study in 2023 focused on the length of
-productive life of female Floridian goats to determine the factors that
-would impact the longevity of the goats. CPH was used to determine that
-the age of first kidding and the specific herd impacts the length of
-productive life of the goats [@ziadi2023]. Both of these studies provide
-examples of the reaches of CPH models outside the health industry.
-
-Continuing with using CPH outside of health data, we will study the
-remaining useful life of a NASA aircraft engine using data provided by
-NASA for the study on propagation modeling [@saxena2008]. From the
-literature review, we have noted that we will need to be concerned with
-the constant hazards ratio within our data and determine how we will
-censor our data.
-
-## Introduction (draft)
-
-Analyzing survival data is challenging for traditional machine learning
-models because they struggle to capture the time-to-event nature of the
-data. Survival data typically contains challenges difficult to account
-for because the datasets may be “censored” meaning that the event hasn’t
-had the opportunity to occur yet by the end of the study period. Also,
-traditional methods aren’t able to effectively model time-to-event data
-because they treat time as fixed values opposed to treating the values
-as time leading to less accurate results.
-
-Cox proportional hazards models are commonly used to interpret, analyze,
-and examine survival data. These models specialize in time-to-event
-predictions and can account for multiple variables and their effects on
-survival. Cox proportional hazards models are commonly used in the
-health field while predicting patient survivability based on symptoms
-and medical history, the financial field to predict bank failure, and
-mechanical studies to predict failure of machines. Cox proportional
-hazards models are popular due to their ability to incorporate multiple
-covariates and they have the advantage of being semi-parametric which
-allows more flexibility as compared to other models.
-
-In this paper, we will be examining the prediction of remaining useful
-life (RUL) for NASA turbofan jet engines. The lifespan of a jet engine
-is important as they cost thousands of dollars to maintain and produce.
-The ability to predict the remaining useful lifespan of these engines
-will allow for better judgments on the finances needed to maintain or
-repair them as well as alert teams that an engine may not have the RUL
-for a mission they would like to use it for.
-
-We will first generate the summary statistics for this dataset as well
-as run linear and logistic regression tests. We will then test our
-assumptions which include the proportional hazards assumption, the
-linearity between covariates and the log hazard, that each data is
-independent of each other and there is no omitted data. A model will be
-generated using the survival package from R. We will then test the model
-and make adjustments as needed before generating our predictions.
 

### 2025-02-19 index.qmd
Dataset link added

@@
 
-<!-- -->
+We will be using the test_FD002 and train_FD002 datasets available for download at
+[NASA's Open Data Portal](https://data.nasa.gov/Aerospace/CMAPSS-Jet-Engine-Simulated-Data/ff5v-kuh6/about_data) and also avaialble on the home page of our [GitHub repository](https://github.com/JR-87/capstone4ds_reed).
+
+```{r}
+testurl <- 'https://raw.githubusercontent.com/JR-87/capstone4ds_reed/refs/heads/main/test_FD002.txt'
+
+test <- read.table(testurl, header = TRUE, sep = '')
+head(test)
+```
+```{r}
+trainurl <- 'https://raw.githubusercontent.com/JR-87/capstone4ds_reed/refs/heads/main/train_FD002.txt'
+
+test <- read.table(trainurl, header = TRUE, sep = '')
+head(test)
+```
 

### 2025-02-19 index.qmd
Minor updates

--- 2025-02-18 index.qmd
@@
 
-Slides: [slides.html](slides.html){target="_blank"}
-
-Literature Comments: [Literature.html](articleSummaries.html)
+[Slides](slides.html){target="_blank"} \|
+[Literature](articleSummaries.html)
 
@@
 the medical world is the most obvious use of survival predictions, the
-same methods can be applied to the predicting bank failure, the
-remaining useful life of machines, machinery maintenance timelines, and
-insurance likelihood of payouts. There are various survival methods that
-can be examined, including Cox proportional hazards model which will be
+same methods can be applied to predicting bank failure, the remaining
+useful life of machines, machinery maintenance timelines, and insurance
+likelihood of payouts. There are various survival methods that can be
+examined, including Cox proportional hazards model which will be
 examined in this paper.
@@
 An additional limitation of CPH comes from the data in the covariates
-selected and how can they can become biased. Two studies done using UK
+selected and how they can become biased. Two studies done using UK
 health data demonstrated how bias can impact the values for the
@@
 
-While health care is the dominate field in which CPH is used, a study in
+While health care is the dominant field in which CPH is used, a study in
 2021 provides an example of using CPH in insurance fields. An insurance
@@
 
-Continuing with using CPH outside of health data, we will study the
-remaining useful life of a NASA aircraft engine using data provided by
-NASA for the study on propagation modeling [@saxena2008]. From the
-literature review, we have noted that we will need to be concerned with
-the constant hazards ratio within our data and determine how we will
-accommodate censored data when creating our model.
-
-<!-- -->
+CPH has the capability to be applied to many fields, though it poses
+some limitations. The constant hazards ratio and determining how to
+censor data are two concerns that can lead to limitations in using CPH
+for models. In addition, the necessity of a time to event value within
+the data source is something that needs to be taken into consideration
+when choosing to use CPH. Without that value, using CPH as a survival
+model will not be the appropriate method to take with that particular
+data. This value is also what makes CPH a prime modeling candidate with
+health data as health metrics frequently have a time component to them,
+especially when comparing someone over a period of time in a health
+study as is done with medications and the development of cancer or other
+illnesses. However, while the usage of CPH is clear in the medical
+field, this paper will demonstrate using CPH within the field of
+mechanics and remaining useful life of a machine.
 
@@
 
+Though CPH is typically used with health data, we wanted to determine
+how CPH can be used in other fields. As such, we will study the
+remaining useful life of a NASA aircraft engine using data provided by
+NASA for the study on propagation modeling [@saxena2008]. In particular,
+we will be examining their data from what they have deemed engine two.
+From the literature review, we have noted that we will need to be
+concerned with the constant hazards ratio within our data and determine
+how we will accommodate censored data when creating our model.
+
+We will be using the test_FD002 and train_FD002 datasets available for
+download at [NASA's Open Data
+Portal](https://data.nasa.gov/Aerospace/CMAPSS-Jet-Engine-Simulated-Data/ff5v-kuh6/about_data)
+and also available on the home page of our [GitHub
+repository](https://github.com/JR-87/capstone4ds_reed).
+
+```{r}
+testurl <- 'https://raw.githubusercontent.com/JR-87/capstone4ds_reed/refs/heads/main/test_FD002.txt'  
+test <- read.table(testurl, header = TRUE, sep = '')
+head(test)
+```
+
+```{r}
+trainurl <- 'https://raw.githubusercontent.com/JR-87/capstone4ds_reed/refs/heads/main/train_FD002.txt'  
+test <- read.table(trainurl, header = TRUE, sep = '') 
+head(test)
+```
+
+##
+
 -   Describe your data sources and collection process.
@@
 
-A study was conducted to determine how...
-
-```{r, warning=FALSE, echo=T, message=FALSE}
-# loading packages 
-library(tidyverse)
-library(knitr)
-library(ggthemes)
-library(ggrepel)
-library(dslabs)
-```
-
-```{r, warning=FALSE, echo=TRUE}
-# Load Data
-kable(head(murders))
-
-ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total))
-
-  ggplot1 + geom_point(aes(col=region), size = 4) +
-  geom_text_repel(aes(label=abb)) +
-  scale_x_log10() +
-  scale_y_log10() +
-  geom_smooth(formula = "y~x", method=lm,se = F)+
-  xlab("Populations in millions (log10 scale)") + 
-  ylab("Total number of murders (log10 scale)") +
-  ggtitle("US Gun Murders in 2010") +
-  scale_color_discrete(name = "Region")+
-      theme_bw()
-
-```
-
-### Modeling and Results
+A study was conducted to determine how...Modeling and Results
 

### 2025-02-20 index.qmd
Data description very rough draft

--- 2025-02-18 index.qmd
@@
 
-<!-- -->
+From NASA's Open Data Portal CMAPSS Jet Engine Simulated Data we will be using the test_FD002 and train_FD002 datasets available for download at
+[NASA's Open Data Portal](https://data.nasa.gov/Aerospace/CMAPSS-Jet-Engine-Simulated-Data/ff5v-kuh6/about_data) and also available on the home page of our [GitHub repository](https://github.com/JR-87/capstone4ds_reed).
+
+These datasets contain time series data for the survival time of different engines under different conditions.
+The conditions are represented by three operational setting fields and 21 senor measurement fields.
+
+Headers of the train and test data are below.
+
+```{r}
+testurl <- 'https://raw.githubusercontent.com/JR-87/capstone4ds_reed/refs/heads/main/test_FD002.txt'
+
+test <- read.table(testurl, header = TRUE, sep = '')
+head(test)
+```
+
+```{r}
+trainurl <- 'https://raw.githubusercontent.com/JR-87/capstone4ds_reed/refs/heads/main/train_FD002.txt'
+
+train <- read.table(trainurl, header = TRUE, sep = '')
+head(test)
+```
+
+The visual below shows a histogram with distributions of survival times of the 260 unique engines in the training data.
+
+```{r, echo=FALSE}
+library(tidyverse)
+
+distributions <- train %>% 
+  group_by(id) %>%
+  summarise(time = max(time))
+
+ggplot(data = distributions,
+       aes(x = time)) +
+  geom_histogram(binwidth = 5,
+                 fill = "grey")+
+  labs(title = "Survival Time Distribution",
+       x = "Survival Time",
+       y = "Frequency")+
+  theme_bw()+
+  theme(plot.title = element_text(hjust = .5))
+
+```
 

### 2025-02-24 index.qmd
Added equation

--- 2025-02-19 index.qmd
@@
 individual will have one of the illness events occur, a model using CPH
-was generated using gender, age, and region as the
-covariates[@zapletal2021]. A study in 2023 focused on the length of
-productive life of female Floridian goats to determine the factors that
-would impact the longevity of the goats. CPH was used to determine that
-the age of first kidding and the specific herd impacts the length of
+was generated using gender, age, and region as the covariates
+[@zapletal2021]. A study in 2023 focused on the length of productive
+life of female Floridian goats to determine the factors that would
+impact the longevity of the goats. CPH was used to determine that the
+age of first kidding and the specific herd impacts the length of
 productive life of the goats [@ziadi2023]. Both of these studies provide
@@
 
--   Detail the models or algorithms used.
+*Notes:*
 
--   Justify your choices based on the problem and data.
+*How coxph was created and creator what it was made for. it is semi
+parametric regression*
 
-*The common non-parametric regression model is*
-$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
-of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
-unknown and* $\varepsilon_i$ *some errors. With the help of this
-definition, we can create the estimation for local averaging i.e.*
-$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
-*is near to* $x$*. In other words, this means that we are discovering
-the line through the data points with the help of surrounding data
-points. The estimation formula is printed below [@R-base]:*
+*explain mathematically with equation*
+
+*How data needs to be formatted - time + event and covariates*
+
+*how it works*
+
+*assumptions - how to test assumptions*
+
+-   *Detail the models or algorithms used.*
+
+-   *Justify your choices based on the problem and data.*
+
+CPH can be estimated as
 
 $$
-M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
-$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
-Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*
+h(t|\mathbf{Z}) = h_0(t)\text{exp}(\Sigma_{k=1}^{p} \beta_kZ_k)
+$$ {#eq-hazard}
 
-*Another equation:*
+where,
 
-$$
-y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
-$$
+-   $h(t|\mathbf{Z})$ is the hazard rate at time $t$ for an individual
+    with risk vector $\mathbf{Z}$
+-   $h_0(t)$ is an arbitrary baseline hazard rate
+-   $\mathbf{Z}$ is the set of $p$ covariates
+-   $\beta$ is the set of $p$ coefficients that measure the impact of
+    the covariates
 
@@
 
-We will be using the test_FD002 and train_FD002 datasets available for
-download at [NASA's Open Data
-Portal](https://data.nasa.gov/Aerospace/CMAPSS-Jet-Engine-Simulated-Data/ff5v-kuh6/about_data)
-and also available on the home page of our [GitHub
-repository](https://github.com/JR-87/capstone4ds_reed).
+Each engine has an unknown amount of wear, manufacturing variation, and
+sensor noise all of which impact the survival time. In the training
+dataset, each engine is operating normally at the beginning of the time
+series, develops a fault, and experiences failure at the end of the time
+series. The conditions the engines are subjected to are represented by
+three operational setting fields and 21 sensor measurement fields. In
+the testing data, the engines have not experienced failure
+[@NASA_CMAPSS].
+
+```{r, echo = FALSE}
+#packages
+library(tidyverse)
+library(gridExtra)
+library(survival)
+library(survminer)
+
+```
 
@@
 testurl <- 'https://raw.githubusercontent.com/JR-87/capstone4ds_reed/refs/heads/main/test_FD002.txt'
+
 test <- read.table(testurl, header = TRUE, sep = '')
@@
 trainurl <- 'https://raw.githubusercontent.com/JR-87/capstone4ds_reed/refs/heads/main/train_FD002.txt'
-test <- read.table(trainurl, header = TRUE, sep = '') 
+
+train <- read.table(trainurl, header = TRUE, sep = '')
 head(test)
@@
 
-##
+The visual below shows a histogram and a box plot representing
+distributions of survival times of the 260 unique engines in the
+training data.
+
+```{r, echo=FALSE}
+distributions <- train %>% 
+  group_by(id) %>%
+  summarise(time = max(time))
+
+Histogram <- ggplot(data = distributions,
+       aes(x = time)) +
+  geom_histogram(binwidth = 5,
+                 fill = "grey")+
+  labs(x = "Survival Time",
+       y = "Frequency")+
+  theme_bw()
+
+box <- ggplot(data = distributions,
+                  aes(y = time))+
+  geom_boxplot(fill = "gray",
+               color = "black")+
+  theme_bw()
+
+grid.arrange(Histogram,box,ncol = 2, top = "Distribution of Engine Longevity in Training Data")
+
+```
+
+```{r}
+summary(distributions$time)
+
+sd(distributions$time)
+```
+
+The engines have a median survival time of 199 iterations with a
+standard deviation of 46.8. The longest surviving engine lasted 378
+iterations, with the shortest survival time for all engines being 128.
+
+```{r}
+
+```
 
@@
 -   Highlight unexpected patterns or anomalies.
+
+### Modeling
 
@@
 
-```{r}
+### Results
 
-```
-
-### Conclusion
+## Conclusion
 

### 2025-02-25 index.qmd
Updated data visualization with the status column and added the univariate data

@@
 how CPH can be used in other fields. As such, we will study the
-remaining useful life of a NASA aircraft engine using data provided by
-NASA for the study on propagation modeling [@saxena2008]. In particular,
-we will be examining their data from what they have deemed engine two.
-From the literature review, we have noted that we will need to be
-concerned with the constant hazards ratio within our data and determine
-how we will accommodate censored data when creating our model.
+remaining useful life (RUL) of a NASA aircraft engine using data
+provided by NASA for the study on propagation modeling [@saxena2008]. In
+particular, we will be examining their data from what they have deemed
+engine two. From the literature review, we have noted that we will need
+to be concerned with the constant hazards ratio within our data and
+determine how we will accommodate censored data when creating our model.
 
@@
 
+For both test and train datasets, we will need to add a column
+indicating status. We will use 0 for still working and 1 for the last
+cycle when the engine fails. For the test data, we know that none of the
+engines have failed so all values in the test column will be 0 for still
+working. For the training data, we know that the data includes all
+cycles up to the last cycle when the machine will fail. As such, we will
+have 0s in all cycles except the last cycle of the machine.
+
 ```{r}
@@
 
-test <- read.table(testurl, header = TRUE, sep = '')
-head(test)
+test <- read.table(testurl, header = TRUE, sep = '') %>%
+  mutate(status = 0)
+head(test, n = 5)
 ```
@@
 
-train <- read.table(trainurl, header = TRUE, sep = '')
-head(test)
+train <- read.table(trainurl, header = TRUE, sep = '') %>%
+  group_by(id) %>%
+  mutate(status = if_else(row_number() == n(), 1, 0)) %>%
+  ungroup()
+
+head(train, n = 5)
 ```
@@
 
+From the study the data is from, we know that some of the sensors will
+not play a role in the RUL. As such, we will run a univariate test of
+all covariates to determine which ones will actually impact the RUL. We
+will use the p-value provided by the Wald test with a significance level
+of $\alpha = 0.05$.
+
 ```{r}
+covariates <- c("os1", "os2", "os3", "sm1", "sm2", "sm3", "sm4", "sm5", "sm6", "sm7",
+                "sm8", "sm9", "sm10", "sm11", "sm12", "sm13", "sm14", "sm15", "sm16",
+                "sm17", "sm18", "sm19", "sm20", "sm21")
 
+univ_formulas <- sapply(covariates,
+                        function(x) as.formula(paste('Surv(time, status)~', x)))
+
+univ_models <- lapply( univ_formulas, function(x){coxph(x, data = train)})
+# Extract data
+univ_results <- lapply(univ_models,
+                       function(x){
+                          x <- summary(x)
+                          p.value<-signif(x$wald["pvalue"], digits=2)
+                          wald.test<-signif(x$wald["test"], digits=2)
+                          beta<-signif(x$coef[1], digits=2);#coeficient beta
+                          HR <-signif(x$coef[2], digits=2);#exp(beta)
+                          HR.confint.lower <- signif(x$conf.int[,"lower .95"], 2)
+                          HR.confint.upper <- signif(x$conf.int[,"upper .95"],2)
+                          HR <- paste0(HR, " (",
+                                       HR.confint.lower, "-", HR.confint.upper, ")")
+                          res<-c(beta, HR, wald.test, p.value)
+                          names(res)<-c("beta", "HR (95% CI for HR)", "wald.test",
+                                        "p.value")
+                          return(res)
+                         })
+as.data.frame(univ_results)
 ```

### 2025-02-25 index.qmd
added label for test/train
Updated references
Added Meier Kaplan

@@
 the testing data, the engines have not experienced failure
-[@NASA_CMAPSS].
+[@saxena2008].
 
@@
 library(survminer)
+library(ggpubr)
+library(magrittr)
+library(ggfortify)
 
@@
 
+**Testing Data**
+
 ```{r}
@@
 
-test <- read.table(testurl, header = TRUE, sep = '') %>%
+test <- read.delim(testurl, header = TRUE, sep = '') %>%
   mutate(status = 0)
@@
 
+**Training Data**
+
 ```{r}
@@
 
-train <- read.table(trainurl, header = TRUE, sep = '') %>%
+train <- read.delim(trainurl, header = TRUE, sep = '') %>%
   group_by(id) %>%
@@
 
+The Kaplan-Meier survival curve is shown below.
+
+```{r}
+t <- as.matrix(train[, "time"])
+s <- as.matrix(train[, "status"])
+
+km_fit <- survfit(Surv(t, s) ~ 1)
+
+autoplot(km_fit) + theme_bw()
+```
+
 From the study the data is from, we know that some of the sensors will
@@
 univ_models <- lapply( univ_formulas, function(x){coxph(x, data = train)})
+
 # Extract data
@@
                           wald.test<-signif(x$wald["test"], digits=2)
-                          beta<-signif(x$coef[1], digits=2);#coeficient beta
+                          beta<-signif(x$coef[1], digits=2);#coefficient beta
                           HR <-signif(x$coef[2], digits=2);#exp(beta)

### 2025-02-27 index.qmd
2/26 updates

@@
 library(ggfortify)
+library(knitr)
 
@@
   mutate(status = 0)
-head(test, n = 5)
+
+kable(head(test, n = 5), format = "markdown")
 ```
@@
 
-head(train, n = 5)
+kable(head(train, n = 5), format = "markdown")
 ```
@@
 distributions of survival times of the 260 unique engines in the
-training data.
+training data. The survival times have a normal distribution
 
-```{r, echo=FALSE}
+The engines have a median survival time of 199 iterations with a
+standard deviation of 46.8. The longest surviving engine lasted 378
+iterations, with the shortest survival time for all engines being 128.
+
+```{r}
 distributions <- train %>% 
@@
 
+summarymetrics <- summary(Minimum = min(time),
+                          Median = median(time),
+                          StandardDeviation = sd(time),
+                          Maximum = max(time), 
+                          distributions$time)
+
+kable(tibble(summarymetrics), format = "markdown")
+
+```
+
+```{r, echo=FALSE}
 Histogram <- ggplot(data = distributions,
@@
 ```
-
-```{r}
-summary(distributions$time)
-
-sd(distributions$time)
-```
-
-The engines have a median survival time of 199 iterations with a
-standard deviation of 46.8. The longest surviving engine lasted 378
-iterations, with the shortest survival time for all engines being 128.
 
@@
                          })
-as.data.frame(univ_results)
+kable(as.data.frame(univ_results))
 ```
+
+Assumptions
+
+Proportional Hazards Assumption
+
+The proportional hazards assumption states the ratio of hazard rates for any two individuals is constant over time regardless of the point in time of the measurement.
+
+The proportional hazards assumption has been proven to be critical for evaluating prior to using Cox Proportional Hazards to model survival data.
 
@@
 
+```{r}
+cox <- coxph(Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5,
+             data = train)
+
+summary(cox)
+```
+```{r}
+cox.zph(cox)
+
+plot(cox.zph(cox))
+```
+
 A study was conducted to determine how...Modeling and Results
@@
 -   Explain your data preprocessing and cleaning steps.
+transformation applied to data to add event
 

### 2025-03-02 index.qmd
3/1 update to methods

@@
 
+When using Cox proportional hazards models to model survival data, the data must include a value for time, a unique identifier column indicating which individual the time series is for, covariates used for predictions, and an event indicating the status of the individual at each point in time.
+The data used in this analysis contain all of the previously mentioned fields with the exception of a status field.
+
 For both test and train datasets, we will need to add a column
@@
 
-The visual below shows a histogram and a box plot representing
-distributions of survival times of the 260 unique engines in the
-training data. The survival times have a normal distribution
-
+The visuals below describe the distribution of the 260 engines in the training data.
+The histogram shows the survival times are normally distributed with a median value of 199 iterations and the boxplot shows there are few outliers out of the 260 engines.
 The engines have a median survival time of 199 iterations with a
@@
 
-Assumptions
+### Assumptions
 
-Proportional Hazards Assumption
+**Proportional Hazards Assumption**
 
-The proportional hazards assumption states the ratio of hazard rates for any two individuals is constant over time regardless of the point in time of the measurement.
+The proportional hazards assumption has been proven to be critical to evaluate prior to using Cox Proportional Hazards when modeling survival data.
+The proportional hazards assumption states the ratio of hazard rates for any two individuals is constant over time regardless of the point in time of the measurement meaning the effect of the covariates must be constant over time.
+The assumption will be tested by calculating scaled Shoenfeld residuals and plotting them against time.
+The resulting plot for each covariate is interpreted to prove the assumption is met if the plot has Shoenfeld residuals roughly constant over time.
+[@Bustan2018]
 
-The proportional hazards assumption has been proven to be critical for evaluating prior to using Cox Proportional Hazards to model survival data.
+**Independent Censoring Assumption**
 
--   Describe your data sources and collection process.
+This assumption describes how if an event is censored it is unrelated to the model being created.
+Censored data is difficult to accommodate when Cox proportional hazards models are used in other industries, such as the medical field, because observation times for the individuals involved in studies may have decades before an event is observed.
+In this scenario, it would not be appropriate to consider this individual when modeling survival.
+In the dataset being used for this model, the training data does not contain censored data.
+The documentation available at the dataset's source clarifies all engines in the training data have experienced the desired event of failure.
 
--   Present initial findings and insights through visualizations.
+**Independence Assumption**
 
--   Highlight unexpected patterns or anomalies.
+Survival times of observed individuals should be independent of each other. While not clarified in the data documentation, for this model it will be assumed that this assumption is true.
+
+### Model Evaluation
+
+To evaluate the model, the Cox regression concordance index will be used.
+Two observed pairs are said to be concordant if an event that is expected to occur before another does occur first.
+For example, if two engines are selected from the data and the model predicts one of them to be at a higher risk of failure than the other and the one with the higher risk fails first, then the pairs are concordant.
+If the alternative happens where the lower risk engine experiences failure first, then the pairs are discordant.
+
+The concordance index is calculated as follows:
+
+$C = \frac{ConcordantPairs}{ConcordantPairs + DiscordantPairs}$
+
+The concordance index will produce a value between 0 and 1 with 1 meaning all pairs are correctly ordered and 0 being the alternative with no pairs correctly ordered.
+The higher the concordance index for the created model, the more accurately it can model survival for the data.[@Therneau2017]
 
 ### Modeling
+
+Creating model
 
@@
 ```
+
+**Checking Proportional Hazards**
 ```{r}
@@
 plot(cox.zph(cox))
+
 ```

### 2025-03-03 index.qmd
updated methods

@@
 
+CPH relies on data that has a time to event component and covariates for
+that event.
+
+### Assumptions
+
+**Proportional Hazards Assumption**
+
+The proportional hazards assumption has been proven to be critical to
+evaluate prior to using CPH when modeling survival data. The
+proportional hazards assumption states that the ratio of hazard rates
+for any two individuals must be constant over time regardless of the
+point in time of the measurement [@Bustan2018] . This means the effect
+of the covariates must be constant over time. This assumption can be
+tested using the R package `survival`. The package contains a
+proportional hazards test called `cox.zph()` [@R-base]. Using the
+hypothesis,
+
+$H_0$: the predictor's coefficient does not vary with time
+
+$H_1$: the predictor's coefficient does vary with time
+
+the `cox.zph()` calculates the scaled Schoenfeld partial residuals and
+returns the $\chi^2$ value, the degrees of freedom, and p-values for
+each individual covariate along with a global value. Using an
+$\alpha = 0.05$ and the global p-value, the determination of whether to
+reject or fail to reject $H_0$ can be made [@nahhas2025]. Individual
+plots for each covariate can then be generated which provide a visual
+interpretation of the p-values generated from `cox.zph()`.
+
+**Independent Censoring Assumption**
+
+This assumption describes how if an event is censored it is unrelated to
+the model being created. Censored data is difficult to accommodate when
+Cox proportional hazards models are used in other industries, such as
+the medical field, because observation times for the individuals
+involved in studies may have decades before an event is observed. In
+this scenario, it would not be appropriate to consider this individual
+when modeling survival.
+
+**Independence Assumption**
+
+Survival times of observed individuals should be independent of each
+other. While not clarified in the data documentation, for this model it
+will be assumed that this assumption is true.
+
+### Equations
+
+CPH can be estimated as
+
+$$
+h(t|\mathbf{Z}) = h_0(t)\text{exp}(\sum_{k=1}^{p} \beta_kZ_k)
+$$ {#eq-hazard}
+
+where,
+
+-   $h(t|\mathbf{Z})$ is the hazard rate at time $t$ for an individual
+    with risk vector $\mathbf{Z}$
+-   $h_0(t)$ is an arbitrary baseline hazard rate
+-   $\mathbf{Z}$ is the set of $p$ covariates
+-   $\beta$ is the set of $p$ coefficients that measure the impact of
+    the covariates [@klein2005]
+
+CPH is a proportional hazards model because the ratio of hazard rates
+between two individuals is
+
+$$
+\frac{h(t|\mathbf{Z})}{h(t|\mathbf{Z}^*)} = \frac{h_0(t)\text{exp}(\Sigma_{k=1}^{p} \beta_kZ_k)}{h_0(t)\text{exp}(\Sigma_{k=1}^{p} \beta_kZ_k^*)} = \text{exp}[\sum_{k=1}^p \beta_k(Z_k-Z_k^*)]
+$$ {#eq-hazardsratio}
+
+where,
+
+-   $\mathbf{Z}$ and $\mathbf{Z}^*$ are covariate values [@klein2005]
+
+### Model Evaluation
+
+To evaluate the model, the Cox regression concordance index will be
+used. Two observed pairs are said to be concordant if an event that is
+expected to occur before another does occur first. For example, if two
+engines are selected from the data and the model predicts one of them to
+be at a higher risk of failure than the other and the one with the
+higher risk fails first, then the pairs are concordant. If the
+alternative happens where the lower risk engine experiences failure
+first, then the pairs are discordant.
+
+The concordance index is calculated as follows:
+
+$C = \frac{ConcordantPairs}{ConcordantPairs + DiscordantPairs}$
+
+The concordance index will produce a value between 0 and 1 with 1
+meaning all pairs are correctly ordered and 0 being the alternative with
+no pairs correctly ordered. The higher the concordance index for the
+created model, the more accurately it can model survival for the
+data.[@Therneau2017]
+
 *Notes:*
@@
 -   *Justify your choices based on the problem and data.*
-
-CPH can be estimated as
-
-$$
-h(t|\mathbf{Z}) = h_0(t)\text{exp}(\Sigma_{k=1}^{p} \beta_kZ_k)
-$$ {#eq-hazard}
-
-where,
-
--   $h(t|\mathbf{Z})$ is the hazard rate at time $t$ for an individual
-    with risk vector $\mathbf{Z}$
--   $h_0(t)$ is an arbitrary baseline hazard rate
--   $\mathbf{Z}$ is the set of $p$ covariates
--   $\beta$ is the set of $p$ coefficients that measure the impact of
-    the covariates
 
@@
 
-When using Cox proportional hazards models to model survival data, the data must include a value for time, a unique identifier column indicating which individual the time series is for, covariates used for predictions, and an event indicating the status of the individual at each point in time.
-The data used in this analysis contain all of the previously mentioned fields with the exception of a status field.
+When using Cox proportional hazards models to model survival data, the
+data must include a value for time, a unique identifier column
+indicating which individual the time series is for, covariates used for
+predictions, and an event indicating the status of the individual at
+each point in time. The data used in this analysis contain all of the
+previously mentioned fields with the exception of a status field.
 
@@
 
-The visuals below describe the distribution of the 260 engines in the training data.
-The histogram shows the survival times are normally distributed with a median value of 199 iterations and the boxplot shows there are few outliers out of the 260 engines.
-The engines have a median survival time of 199 iterations with a
-standard deviation of 46.8. The longest surviving engine lasted 378
-iterations, with the shortest survival time for all engines being 128.
+The visuals below describe the distribution of the 260 engines in the
+training data. The histogram shows the survival times are normally
+distributed with a median value of 199 iterations and the boxplot shows
+there are few outliers out of the 260 engines. The engines have a median
+survival time of 199 iterations with a standard deviation of 46.8. The
+longest surviving engine lasted 378 iterations, with the shortest
+survival time for all engines being 128.
 
@@
 
-### Assumptions
-
-**Proportional Hazards Assumption**
-
-The proportional hazards assumption has been proven to be critical to evaluate prior to using Cox Proportional Hazards when modeling survival data.
-The proportional hazards assumption states the ratio of hazard rates for any two individuals is constant over time regardless of the point in time of the measurement meaning the effect of the covariates must be constant over time.
-The assumption will be tested by calculating scaled Shoenfeld residuals and plotting them against time.
-The resulting plot for each covariate is interpreted to prove the assumption is met if the plot has Shoenfeld residuals roughly constant over time.
-[@Bustan2018]
-
-**Independent Censoring Assumption**
-
-This assumption describes how if an event is censored it is unrelated to the model being created.
-Censored data is difficult to accommodate when Cox proportional hazards models are used in other industries, such as the medical field, because observation times for the individuals involved in studies may have decades before an event is observed.
-In this scenario, it would not be appropriate to consider this individual when modeling survival.
-In the dataset being used for this model, the training data does not contain censored data.
-The documentation available at the dataset's source clarifies all engines in the training data have experienced the desired event of failure.
-
-**Independence Assumption**
-
-Survival times of observed individuals should be independent of each other. While not clarified in the data documentation, for this model it will be assumed that this assumption is true.
-
-### Model Evaluation
-
-To evaluate the model, the Cox regression concordance index will be used.
-Two observed pairs are said to be concordant if an event that is expected to occur before another does occur first.
-For example, if two engines are selected from the data and the model predicts one of them to be at a higher risk of failure than the other and the one with the higher risk fails first, then the pairs are concordant.
-If the alternative happens where the lower risk engine experiences failure first, then the pairs are discordant.
-
-The concordance index is calculated as follows:
-
-$C = \frac{ConcordantPairs}{ConcordantPairs + DiscordantPairs}$
-
-The concordance index will produce a value between 0 and 1 with 1 meaning all pairs are correctly ordered and 0 being the alternative with no pairs correctly ordered.
-The higher the concordance index for the created model, the more accurately it can model survival for the data.[@Therneau2017]
-
 ### Modeling
@@
 **Checking Proportional Hazards**
+
 ```{r}
@@
 
+**Checking Independent Censoring**
+
+In the dataset being used for this model, the training data does not
+contain censored data. The documentation available at the dataset's
+source clarifies all engines in the training data have experienced the
+desired event of failure.
+
 A study was conducted to determine how...Modeling and Results
 
--   Explain your data preprocessing and cleaning steps.
-transformation applied to data to add event
+-   Explain your data preprocessing and cleaning steps. transformation
+    applied to data to add event
 

### 2025-03-04 index.qmd
Updated methodology

@@
 
-CPH relies on data that has a time to event component and covariates for
-that event.
+To determine whether to use CPH for modeling, the data must contain a
+time to event value and meet various assumptions, including proportional
+hazards. Data that is censored will require a survival analysis that
+type of analysis is deliberately designed for censored data. The data
+must also have a binary indicator variable for the event occurrence and
+no missing values [@nahhas2025].
+
+### Equations
+
+**General Equation**
+
+CPH can be estimated as
+
+$$ h(t|\mathbf{Z}) = h_0(t)\text{exp}(\sum_{k=1}^{p} \beta_kZ_k) $$
+
+where,
+
+-   $h(t|\mathbf{Z})$ is the hazard rate at time $t$ for an individual
+    with risk vector $\mathbf{Z}$
+-   $h_0(t)$ is an arbitrary baseline hazard rate
+-   $\mathbf{Z}$ is the set of $p$ covariates
+-   $\beta$ is the set of $p$ coefficients that measure the impact of
+    the covariates [@klein2005]
+
+This equation can then be adapted for the specific dataset being
+studied.
+
+**Hazards Ratio**
+
+CPH is a proportional hazards model because the ratio of hazard rates
+between two individuals is
+
+$$ \frac{h(t|\mathbf{Z})}{h(t|\mathbf{Z}^*)} = \frac{h_0(t)\text{exp}(\Sigma_{k=1}^{p} \beta_kZ_k)}{h_0(t)\text{exp}(\Sigma_{k=1}^{p} \beta_kZ_k^*)} = \text{exp}[\sum_{k=1}^p \beta_k(Z_k-Z_k^*)] $$
+
+where,
+
+-   $\mathbf{Z}$ and $\mathbf{Z}^*$ are covariate values [@klein2005]
+
+### Modeling
+
+**Kaplan-Meier**
+
+Before generating the CPH model, it is beneficial to look at the
+Kaplan-Meier curve. The Kaplan-Meier estimate ...
+
+**CPH Model**
+
+To generate the CPH model, a survival object must first be created.
+Using the R package `survival`, the suvival formula can be generated by
+running `Surv(time, event) ~ x`. That formula is then applied to the R
+function `coxph(formula, data=)` which will generated the CPH model
+[@R-base].
 
 ### Assumptions
+
+For the CPH model to be valid, the following assumptions must be met. If
+they are not met, the data must either be transformed.
 
@@
 
-$H_0$: the predictor's coefficient does not vary with time
-
-$H_1$: the predictor's coefficient does vary with time
+$$
+\begin{align*}
+H_0&: \text{the predictor's coefficient does vary with time} \\
+H_1&: \text{the predictor's coefficient does not vary with time}
+\end{align*}
+$$
 
@@
 
-### Equations
+### Model Evaluation
 
-CPH can be estimated as
+**Predictors**
 
-$$
-h(t|\mathbf{Z}) = h_0(t)\text{exp}(\sum_{k=1}^{p} \beta_kZ_k)
-$$ {#eq-hazard}
+Hypothesis test and wording for the significance of the predictors.
 
-where,
-
--   $h(t|\mathbf{Z})$ is the hazard rate at time $t$ for an individual
-    with risk vector $\mathbf{Z}$
--   $h_0(t)$ is an arbitrary baseline hazard rate
--   $\mathbf{Z}$ is the set of $p$ covariates
--   $\beta$ is the set of $p$ coefficients that measure the impact of
-    the covariates [@klein2005]
-
-CPH is a proportional hazards model because the ratio of hazard rates
-between two individuals is
-
-$$
-\frac{h(t|\mathbf{Z})}{h(t|\mathbf{Z}^*)} = \frac{h_0(t)\text{exp}(\Sigma_{k=1}^{p} \beta_kZ_k)}{h_0(t)\text{exp}(\Sigma_{k=1}^{p} \beta_kZ_k^*)} = \text{exp}[\sum_{k=1}^p \beta_k(Z_k-Z_k^*)]
-$$ {#eq-hazardsratio}
-
-where,
-
--   $\mathbf{Z}$ and $\mathbf{Z}^*$ are covariate values [@klein2005]
-
-### Model Evaluation
+**Concordance Index**
 
@@
 data.[@Therneau2017]
+
+### Predictions
 

### 2025-03-04 index.qmd
3/3 updates to methods

@@
 
+**Linearity Assumption**
+
+The linearity assumption states that the relationship between covariates and the log of the hazard rate must be linear.
+This assumption can be checked by plotting Martingale residuals against the range of values present in each covariate.
+
+Martingale residuals are a type of residuals used in survival analysis that show the discrepancy between the observed and expected number of events. 
+$MartingaleResiduals = ObservedEvents - ExpectedEvents$
+
+If the resulting residuals are linear and appear to have a slope of zero, this assumption is not violated.[@Amini2015]
+
 **Independent Censoring Assumption**
 
-This assumption describes how if an event is censored it is unrelated to
-the model being created. Censored data is difficult to accommodate when
+This assumption describes how if an event is censored it should not be considered when
+the model is being created. Censored data is difficult to accommodate when
 Cox proportional hazards models are used in other industries, such as
 the medical field, because observation times for the individuals
-involved in studies may have decades before an event is observed. In
+involved in studies may have decades before an event is observed and the event may not have the opportunity to occur during the study period. In
 this scenario, it would not be appropriate to consider this individual
@@
 Survival times of observed individuals should be independent of each
-other. While not clarified in the data documentation, for this model it
-will be assumed that this assumption is true.
+other.
+In an example where survival of plane engines were being evaluated, if each plane has two engines and the failure of one engine caused the accelerated failure of another due to increased workload, it would produce biased conclusions and violate the independence assumption.
 
@@
 
+**Checking Linearity Assumption**
+
+Example for one predictor
+```{r}
+martingale <- resid(cox, type = "martingale")
+
+plot(train$os1, martingale, main = "Martingale Residuals: os1")
+abline(h = 0, col = "red")
+
+```
+
 **Checking Independent Censoring**

### 2025-03-08 index.qmd
updated kaplan meier section

@@
 
-To determine whether to use CPH for modeling, the data must contain a
-time to event value and meet various assumptions, including proportional
-hazards. Data that is censored will require a survival analysis that
-type of analysis is deliberately designed for censored data. The data
-must also have a binary indicator variable for the event occurrence and
-no missing values [@nahhas2025].
+### Data Requirements
+
+CPH requires data to contain a time to event value and have a binary
+indicator variable for the event occurrence. The data must not contain
+any missing values, but can be data that is censored which CPH is
+designed to handle [@nahhas2025]. In addition, the data must meet
+several assumptions for CPH to be a valid model to use on the data which
+are discussed below.
 
@@
 
-This equation can then be adapted for the specific dataset being
-studied.
+This equation can then be adapted for the specific data being studied.
 
@@
 
-**Kaplan-Meier**
+**Kaplan-Meier Estimator and Curve**
 
-Before generating the CPH model, it is beneficial to look at the
-Kaplan-Meier curve. The Kaplan-Meier estimate ...
+The Kaplan-Meier estimator and curve are frequently used in survival
+analysis, especially when dealing with censored data or data where the
+time-to-event value fluctuates per subject [@klein2005]. The
+Kaplan-Meier estimate was proposed by Edward L. Kaplan and Paul Meier in
+1958 to deal with incomplete observations [@rich2010]. It is the
+standard estimator of the survival function and is defined by the
+following equation:
+
+$$
+\hat{S}(t) = \begin{cases}
+1 & \text{if }\; t < t_1, \\
+\prod_{t_i \leq t}[1-\frac{d_i}{Y_i}], & \text{if }\; t_1 \leq t
+\end{cases}
+$$
+
+where,
+
+-   $d_i$ is the number of events at time $t_i$
+-   $Y_i$ is the number of individuals who are at risk at time $t_i$
+
+for all values of $t$ where there is data. The estimator is not
+well-defined for $t$ beyond the largest data value for time
+[@klein2005].
+
+The Kaplan-Meier curve on the X-axis represents the length of survival
+time on the interval from 0 to the largest observation time while the
+Y-axis represents the cumulative probability of surviving at a given
+time. The curve is based on step-wise estimates thus making the curve
+non-continuous [@rich2010]. Generating a Kaplan-Meier curve requires the
+R packages `survival` and `survminer` and uses the survival object and
+survival formula for the data. The below curve was generated using the
+`survival` package and dataset "lung" and demonstrates the general
+survival probability.
+
+```{r}
+library(survival)
+library(survminer)
+ggsurvplot(survfit(Surv(time, status) ~ 1, data = lung))
+
+```
+
+Kaplan-Meier can be used to compare the survival probability of
+different groups, such as the sex of an individual. The log rank test
+can be used to determine the significance of each group. The test
+calculates the $\chi^2$ value for each event time for each group and
+sums the results which are then added together to generate the $\chi^2$
+value for the full curve for each group [@rich2010]. The hypothesis test
+for the log rank test is:
+
+$$
+\begin{align*}
+H_0 &: \text{in terms of survivability, there is no difference between the two groups} \\
+H_1 &: \text{there is a survival difference between the two groups} 
+\end{align*}
+$$
+
+```{r}
+survdiff(Surv(time, status) ~ sex, data = lung)
+```
+
+Using an $\alpha = 0.05$ and the same data as the above Kaplan-Meier
+curve comparing sex, the $H_0$ hypothesis is rejected as there is enough
+evidence to suggest there is a survival difference between males and
+females.
+
+While the Kaplan-Meier curve and estimator are useful, there is an
+assumption of non-informative censoring when using them. This assumption
+means that there is no additional information on the survivial
+likelihood for a subject after they have been censored. If that
+assumption is violated, then Kaplan-Meier curves will not be estimating
+the correct equation [@klein2005].
 
@@
 To generate the CPH model, a survival object must first be created.
-Using the R package `survival`, the suvival formula can be generated by
+Using the R package `survival`, the survival formula can be generated by
 running `Surv(time, event) ~ x`. That formula is then applied to the R
@@
 
-For the CPH model to be valid, the following assumptions must be met. If
-they are not met, the data must either be transformed.
+For a CPH model to be valid, the following assumptions must be met
+either from the raw data or through transformation.
 
@@
 
-The linearity assumption states that the relationship between covariates and the log of the hazard rate must be linear.
-This assumption can be checked by plotting Martingale residuals against the range of values present in each covariate.
+The linearity assumption states that the relationship between covariates
+and the log of the hazard rate must be linear. This assumption can be
+checked by plotting Martingale residuals against the range of values
+present in each covariate.
 
-Martingale residuals are a type of residuals used in survival analysis that show the discrepancy between the observed and expected number of events. 
-$MartingaleResiduals = ObservedEvents - ExpectedEvents$
+Martingale residuals are a type of residuals used in survival analysis
+that show the discrepancy between the observed and expected number of
+events. $MartingaleResiduals = ObservedEvents - ExpectedEvents$
 
-If the resulting residuals are linear and appear to have a slope of zero, this assumption is not violated.[@Amini2015]
+If the resulting residuals are linear and appear to have a slope of
+zero, this assumption is not violated.[@Amini2015]
 
@@
 
-This assumption describes how if an event is censored it should not be considered when
-the model is being created. Censored data is difficult to accommodate when
-Cox proportional hazards models are used in other industries, such as
-the medical field, because observation times for the individuals
-involved in studies may have decades before an event is observed and the event may not have the opportunity to occur during the study period. In
-this scenario, it would not be appropriate to consider this individual
-when modeling survival.
+This assumption describes how if an event is censored it should not be
+considered when the model is being created. Censored data is difficult
+to accommodate when Cox proportional hazards models are used in other
+industries, such as the medical field, because observation times for the
+individuals involved in studies may have decades before an event is
+observed and the event may not have the opportunity to occur during the
+study period. In this scenario, it would not be appropriate to consider
+this individual when modeling survival.
 
@@
 Survival times of observed individuals should be independent of each
-other.
-In an example where survival of plane engines were being evaluated, if each plane has two engines and the failure of one engine caused the accelerated failure of another due to increased workload, it would produce biased conclusions and violate the independence assumption.
+other. In an example where survival of plane engines were being
+evaluated, if each plane has two engines and the failure of one engine
+caused the accelerated failure of another due to increased workload, it
+would produce biased conclusions and violate the independence
+assumption.
 
 ### Model Evaluation
+
+Evaluating a CPH model requires examining the predictors to determine
+which ones are significant and generating the concordance index which
+will determine how accurate the model is at predicting survival.
 
@@
 
-*Notes:*
-
-*How coxph was created and creator what it was made for. it is semi
-parametric regression*
-
-*explain mathematically with equation*
-
-*How data needs to be formatted - time + event and covariates*
-
-*how it works*
-
-*assumptions - how to test assumptions*
-
--   *Detail the models or algorithms used.*
-
--   *Justify your choices based on the problem and data.*
+TBD
 
@@
 
-```{r, echo = FALSE}
+```{r}
 #packages
@@
 Example for one predictor
+
 ```{r}

### 2025-03-08 index.qmd
minor update

@@
 Cox proportional hazards (CPH) models are statistical regression models
-specializing in modeling time-to-event predictions with survival data.
-Survival data is data with a value for time and an event
-[@Abeysekera2009]. For example, in a study investigating the survival
-likelihood of mechanical equipment, the time value would begin when the
-equipment begins functioning and would conclude in the event the
-equipment stops functioning [@Smith2003]. CPH models can use single or
-multiple covariates as a means of creating predictions until failure and
-are capable of dealing with survival data that contains censored data as
-well [@Seung2023]. Censored data is when the information about an
-individual in a study is only known for a certain period of time, such
-as the length of the study [@klein2005, 1]. The models also have the
-advantage of being semi-parametric which allows more flexibility as
-compared to other models [@Ming-Chiang2014].
+specializing in modeling time-to-event predictions with survival data
+introduced by D. R. Cox in 1972 [@nahhas2025]. Survival data is data
+with a value for time and an event [@Abeysekera2009]. For example, in a
+study investigating the survival likelihood of mechanical equipment, the
+time value would begin when the equipment begins functioning and would
+conclude in the event the equipment stops functioning [@Smith2003]. CPH
+models can use single or multiple covariates as a means of creating
+predictions until failure and are capable of dealing with survival data
+that contains censored data as well [@Seung2023]. Censored data is when
+the information about an individual in a study is only known for a
+certain period of time, such as the length of the study [@klein2005].
+The models also have the advantage of being semi-parametric which allows
+more flexibility as compared to other models [@Ming-Chiang2014].
 
@@
 `survival` package and dataset "lung" and demonstrates the general
-survival probability.
+survival probability using the `ggsurvplot` function [@R-base].
 
@@
 running `Surv(time, event) ~ x`. That formula is then applied to the R
-function `coxph(formula, data=)` which will generated the CPH model
-[@R-base].
+function `coxph(formula, data=)` which will generate the CPH model
+[@R-base]. The output of the CPH model includes the coefficients
+(\$\\beta\$ and the confidence intervals for each covariate, and the
+concordance value, likelihood ratio test, Wald test, and the score
+(logrank) test of the overall model [@nahhas2025].
 

### 2025-03-09 index.qmd
3/8 updates

--- 2025-03-04 index.qmd
@@
 
-To determine whether to use CPH for modeling, the data must contain a
-time to event value and meet various assumptions, including proportional
-hazards. Data that is censored will require a survival analysis that
-type of analysis is deliberately designed for censored data. The data
-must also have a binary indicator variable for the event occurrence and
-no missing values [@nahhas2025].
+### Data Structure
 
-### Equations
+In order to use CPH to model time to event survival data, the data must contain a value for time, a binary field indicating the event being modeled, an id field, and one or multiple covariate fields.
+The time value is typically formatted as a time series for each individual in the dataset.
+Individuals will have a time value from the beginning of the observation period, to an arbitrary time that may or may not be the time the event being modeled for occurs.
+Data is considered censored when the individual is not given the opportunity to experience the event in the observation period and is handled differently by the model.
+The covariate fields in the dataset can be categorical or numerical.
+[@nahhas2025].
 
-**General Equation**
+### Creating Cox Proportional Hazards Model
 
-CPH can be estimated as
+#### Choosing covariates
 
-$$ h(t|\mathbf{Z}) = h_0(t)\text{exp}(\sum_{k=1}^{p} \beta_kZ_k) $$
+From the study the data is from, we know that some of the sensors will
+not play a role in the RUL. As such, we will run a univariate test of
+all covariates to determine which ones will actually impact the RUL. We
+will use the p-value provided by the Wald test with a significance level
+of $\alpha = 0.05$.
 
-where,
+#### Generating Model
+To generate the CPH model with properly formatted survival data, a survival object must first be created using the R package `survival`.
+The survival object is created by running `Surv(time, event)` where `time` is the time field and `event` is the event field.
 
--   $h(t|\mathbf{Z})$ is the hazard rate at time $t$ for an individual
-    with risk vector $\mathbf{Z}$
--   $h_0(t)$ is an arbitrary baseline hazard rate
--   $\mathbf{Z}$ is the set of $p$ covariates
--   $\beta$ is the set of $p$ coefficients that measure the impact of
-    the covariates [@klein2005]
+To create the CPH model, the survival object is then applied to the R
+function `coxph(survival object ~ x, data)` where `x` represents one or multiple covariate(s) and `data` is the survival data that is being modeled.[@R-base].
 
-This equation can then be adapted for the specific dataset being
-studied.
-
-**Hazards Ratio**
-
-CPH is a proportional hazards model because the ratio of hazard rates
-between two individuals is
-
-$$ \frac{h(t|\mathbf{Z})}{h(t|\mathbf{Z}^*)} = \frac{h_0(t)\text{exp}(\Sigma_{k=1}^{p} \beta_kZ_k)}{h_0(t)\text{exp}(\Sigma_{k=1}^{p} \beta_kZ_k^*)} = \text{exp}[\sum_{k=1}^p \beta_k(Z_k-Z_k^*)] $$
-
-where,
-
--   $\mathbf{Z}$ and $\mathbf{Z}^*$ are covariate values [@klein2005]
-
-### Modeling
-
-**Kaplan-Meier**
-
-Before generating the CPH model, it is beneficial to look at the
-Kaplan-Meier curve. The Kaplan-Meier estimate ...
-
-**CPH Model**
-
-To generate the CPH model, a survival object must first be created.
-Using the R package `survival`, the suvival formula can be generated by
-running `Surv(time, event) ~ x`. That formula is then applied to the R
-function `coxph(formula, data=)` which will generated the CPH model
-[@R-base].
+During the analysis, multiple models will be created with a different combination of covariates.
 
@@
 
-For the CPH model to be valid, the following assumptions must be met. If
-they are not met, the data must either be transformed.
+When creating a CPH model, to ensure an accurate model is created the following assumptions must be met. If
+they are not met, the data must either be transformed, stratified, or other covariates may need to be selected for the model. The assumptions associated with the CPH model are defined below.
 
-**Proportional Hazards Assumption**
+#### Proportional Hazards Assumption
 
@@
 for any two individuals must be constant over time regardless of the
-point in time of the measurement [@Bustan2018] . This means the effect
-of the covariates must be constant over time. This assumption can be
-tested using the R package `survival`. The package contains a
+point in time of the measurement. This means the effect
+of the covariates must be constant over time. [@Bustan2018]
+
+This assumption can be tested using the R package `survival`. The package contains a
 proportional hazards test called `cox.zph()` [@R-base]. Using the
@@
 $\alpha = 0.05$ and the global p-value, the determination of whether to
-reject or fail to reject $H_0$ can be made [@nahhas2025]. Individual
-plots for each covariate can then be generated which provide a visual
-interpretation of the p-values generated from `cox.zph()`.
+reject or fail to reject $H_0$ can be made [@nahhas2025].
 
-**Linearity Assumption**
+Individual plots for each covariate are generated which provide a visual
+interpretation of the p-values generated from `cox.zph()`. If there is no clear slope in the residuals on the plot, it signifies further confirmation the proportional hazards assumption is met.
+
+#### Linearity Assumption
 
@@
 
+Residuals can first be found by using the `resid(type = "martingale")` function and passing martingale in the type argument creating a vector of residuals that can be plotted against the range of values for each covariate.
+
 If the resulting residuals are linear and appear to have a slope of zero, this assumption is not violated.[@Amini2015]
 
-**Independent Censoring Assumption**
+#### Independent Censoring Assumption
 
@@
 
-**Independence Assumption**
+#### Independence Assumption
 
@@
 
-**Predictors**
-
-Hypothesis test and wording for the significance of the predictors.
-
 **Concordance Index**
 
-To evaluate the model, the Cox regression concordance index will be
+To evaluate the model, the Cox regression concordance index can be
 used. Two observed pairs are said to be concordant if an event that is
@@
 
-### Predictions
+### Model Applications and Predictions
 
-*Notes:*
+Cox proportional hazards models can produce an assortment of useful metrics to interpret the model.
+The model created can be used to estimate risk of the event occurring, rate of occurrence, survival probability, and time to event.
 
-*How coxph was created and creator what it was made for. it is semi
-parametric regression*
+#### Hazard Ratio
 
-*explain mathematically with equation*
+CPH models focus on providing a relative risk of the event happening based on the covariates and value for time.
+The hazards ratio is used to evaluate this level of risk and is calculated by finding the ratio between two individuals where one is the baseline value of a covariate at time $t$, and the other is the value of a covariate with a one unit increase.
+The hazards ratio is represented by the formula below.
 
-*How data needs to be formatted - time + event and covariates*
+$$ \frac{h(t|\mathbf{Z})}{h(t|\mathbf{Z}^*)} = \frac{h_0(t)\text{exp}(\Sigma_{k=1}^{p} \beta_kZ_k)}{h_0(t)\text{exp}(\Sigma_{k=1}^{p} \beta_kZ_k^*)} = \text{exp}[\sum_{k=1}^p \beta_k(Z_k-Z_k^*)] = \text{exp}(\sum_{k=1}^{p} \beta_kZ_k) $$
 
-*how it works*
+where,
 
-*assumptions - how to test assumptions*
+-   $\mathbf{Z}$ and $\mathbf{Z}^*$ are covariate values.
 
--   *Detail the models or algorithms used.*
+A hazards ratio can be interpreted as:
+$h(t|\mathbf{Z})$ < 1 indicates a decreased baseline risk as compared to the baseline.
+$h(t|\mathbf{Z})$ > 1 indicates an increased baseline risk as compared to the baseline.
 
--   *Justify your choices based on the problem and data.*
+Hazard ratios can be interpreted to give an estimation of the event occurring.
+For example, a hazard ratio of 1.3 signifies that for a one unit increase in the covariate, a 30% increase in risk of the event occurring is expected.
+Hazard ratios can also be used to calculate hazard rates, which have theiur own applications and interpretations when modeling survival data.
+[@klein2005]
+
+#### Hazard Rate
+
+The hazard rate is defined as:
+
+$$ h(t|\mathbf{Z}) = h_0(t)\text{exp}({\sum_{k=1}^{p} \beta_kZ_k}) $$
+
+where,
+
+-   $h(t|\mathbf{Z})$ is the hazard rate at time $t$ for an individual
+    with risk vector $\mathbf{Z}$
+
+-   $\text{exp}(\sum_{k=1}^{p} \beta_kZ_k)$ is the hazard ratio as explained above.
+
+-   $h_0(t)$ is a baseline hazard rate indicating the hazard rate when all covariates are zero.
+
+The resulting hazard rate is a measure of the instantaneous risk of the event happening at a specific point in time. Similar to the hazard ratio, hazard rates are interpreted as:
+
+$h(t|\mathbf{Z})$ < 1 indicates a lower risk of the event occurring at the specific point in time.
+$h(t|\mathbf{Z})$ > 1 indicates a higher risk of the event occurring at the specific point in time.
+
+Hazard rates can be used as an additional measure of risk, as well as to calculate the cumulative hazard for an individual. [@klein2005]
+
+The hazard rate can be calculated in R by using `stats::predict()` after creating a CPH model. This function will be used to create a column showing the hazard rate for each row in the testing data.
+
+#### Cumulative Hazard
+
+Cumulative hazard represents the total accumulated risk of an individual for the duration of the observation period at time $t$ and is defined as:
+
+$H(t) = \int_0^t h(t) dt$
+where,
+-   $H(t)$ is the cumulative hazard at time $t$
+
+-   $h(t)$ is the hazard rate at time t
+
+As defined by the formula, cumulative hazard is a summation of the hazard rates for an individual from time 0 to time $t$.
+A higher cumulative hazard indicates a higher risk of event.
+Cumulative hazard is used as an additional measure for risk, but it is mainly used to calculate survival probability to plot against time on survival curves.
+
+The cumulative hazard will be calculated in R by grouping the data by the id field and finding the summation of the hazard rates for the individual through the duration of the observation period.
+A column will be added showing the cumulative hazard for each individual at ach point in time.
+
+#### Calculating Survival Probability
+
+$S(t) = exp(-H(t))$
+
+where,
+
+-   $S(t)$ is the probability of survival for an individual at time $t$
+-   $H(t)$ is the cumulative hazard for an individual at time $t$
+
+The resulting survival probability will be between 0 and 1 with a value close to one indicating a high probability of the individual not experiencing the event.
+Conversely, a value close to 0 indicates a higher probability of the individual experiencing the event.
+A survival probability of .05 indicates that only 5% of individuals under the same conditions would survive.
+Survival probability is an important metric when modeling, visualizing, and making predictions with survival data.
+
+Survival probabilities can be calculated in R by using `stats::predict()` after creating a CPH model.
+This function will be used to create a column showing the survival probability for each row in the testing data.
+
+#### Predicting Time Until Event
+
+### Model Visualization
+
+There are multiple different methods to visualize data when creating a  Cox proportional hazards model.
+Survival curves are a popular way to show survival probabilities over time, but there are other visualizations that are helpful when interpreting the results of the model.
+
+#### Forest Plot
+Forest plots can be used to visualize the effects of each covariate in the model on the hazard ratio.
+The resulting plot contains each covariate on the y axis, with the estimated effect on the hazard ratio on the the x axis.
+The plot contains whiskers demonstrating a 95% confidence interval for the effect.
+A positive effect indicates a positive correlation with the hazard with a higher number signifying a more severe effect on the hazard ratio.
+
+A forest plot will be created using the `plot_model()` function from the sjPlot package. The CPH model can be passed as the sole argument to visualize the hazard ratios.
+
+#### Kaplan-Meier Survival Curve
+
+After generating the CPH model, a Kaplan-Meier survival curve can be generated to view survival probabilities against time.
+These plots are important when modeling survival data because they clearly demonstrate the time where the event being modeled for is expected to occur.
+
+The survival curve will be plotted by using the survival package's `survfit()` function. The object created will be plotted using `plot()`
 
@@
 library(knitr)
+library(sjPlot)
 
@@
                  fill = "grey")+
-  labs(x = "Survival Time",
+  labs(x = "Survival Time (Iterations)",
        y = "Frequency")+
@@
                color = "black")+
+  ylab("Time (Iterations)")+
   theme_bw()
@@
 
-From the study the data is from, we know that some of the sensors will
-not play a role in the RUL. As such, we will run a univariate test of
-all covariates to determine which ones will actually impact the RUL. We
-will use the p-value provided by the Wald test with a significance level
-of $\alpha = 0.05$.
+Below is a univariate test to determine which covariates to use in the model
 
@@
 cox <- coxph(Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5,
-             data = train)
+             data = train,
+             id = id)
 
@@
 ```
+
+Create df with max time for each individual to use in predictions
+
+```{r}
+predictiondata <- test %>%
+  group_by(id) %>%
+  slice_max(time) %>%
+  select(id, time, os1, os3, sm1, sm2, sm5, status)
+
+coef <- data.frame(Variable = c("Z1", "Z2", "Z3", "Z4", "Z5"),
+           CoefficientValue = cox$coefficients)
+kable(coef)
+```
+
+Hazard Rate Calculation: $h(t|Z) = h_0(t) * e^{B_1Z_1 + B_2Z_2 + B_3Z_3+B_4Z_4+B_5Z_5}$
+
+```{r}
+test$hazardrates <- predict(cox,
+        newdata = test,
+        type = "risk")
+
+```
+
+$H(t) = \int_0^t h(t) dt$
+where,
+$H(t)$ is the cumulative hazard from time 0 to time $(t)$
+$h(t)$ is the hazard rate at time $t$
+
+```{r}
+test <- test %>%
+  arrange(id, time) %>%
+  group_by(id) %>%
+  mutate(cumulative_hazard = cumsum(hazardrates))
+```
+
+Predicting survival probability
+
+```{r}
+test$survprob <- predict(cox, type = "survival", newdata = test)
+
+```
+
+Forest Plot
+
+```{r}
+plot_model(cox,dot.size = 4,line.size = 2,colors = "red")+
+  theme_bw()+
+  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
+  ylab(label = "Covariates")+
+  xlab(label = "Estimated Effect on Hazard Ratio")
+```
+survival curve
+
+```{r}
+survfit <- survfit(cox, newdata = test)
+plot(survfit, 
+     col = "blue", 
+     conf.int = TRUE, 
+     mark.time = FALSE,
+     xlab = "Time (Iterations)",
+     ylab = "Survival Probability",
+     main = "Survival Probability Over Time")
+```
+
+Probability Calculation: $S(t) = e^{-H(t)}$
+$H(t)$ is the cumulative hazard up to time $t$
 

### 2025-03-09 index.qmd
3/9 updates

@@
 
+After verifying the survival data is formatted properly, the CPH model can be created in R.
+Important steps to consider are choosing covariates and verifying the model meet the model's assumptions after the model is created.
+
 #### Choosing covariates
@@
 
+### Model Visualization
+
+There are multiple different methods to visualize data when creating a  Cox proportional hazards model.
+Survival curves are a popular way to show survival probabilities over time, but there are other visualizations that are helpful when interpreting the results of the model.
+
+#### Forest Plot
+Forest plots can be used to visualize the effects of each covariate in the model on the hazard ratio.
+The resulting plot contains each covariate on the y axis, with the estimated effect on the hazard ratio on the the x axis.
+The plot contains whiskers demonstrating a 95% confidence interval for the effect.
+A positive effect indicates a positive correlation with the hazard with a higher number signifying a more severe effect on the hazard ratio.
+
+A forest plot will be created using the `plot_model()` function from the sjPlot package. The CPH model can be passed as the sole argument to visualize the hazard ratios.
+
+#### Kaplan-Meier Survival Curve
+
+After generating the CPH model, a Kaplan-Meier survival curve can be generated to view survival probabilities against time.
+These plots are important when modeling survival data because they clearly demonstrate the time where the event being modeled for is expected to occur.
+
+The survival curve will be plotted by using the survival package's `survfit()` function. The object created will be plotted using `plot()`
+
 ### Model Applications and Predictions
@@
 where,
+
 -   $H(t)$ is the cumulative hazard at time $t$
@@
 
-### Model Visualization
+Time until event can be calculated by modeling survival probability for future time values.
+When survival probability drops below 50%, it will be assumed the event occurred.
+Predictions for time until event will be made on the testing data for each individual's maximum time value.
+A column will be added to the table showing the expected time until event.
 
-There are multiple different methods to visualize data when creating a  Cox proportional hazards model.
-Survival curves are a popular way to show survival probabilities over time, but there are other visualizations that are helpful when interpreting the results of the model.
-
-#### Forest Plot
-Forest plots can be used to visualize the effects of each covariate in the model on the hazard ratio.
-The resulting plot contains each covariate on the y axis, with the estimated effect on the hazard ratio on the the x axis.
-The plot contains whiskers demonstrating a 95% confidence interval for the effect.
-A positive effect indicates a positive correlation with the hazard with a higher number signifying a more severe effect on the hazard ratio.
-
-A forest plot will be created using the `plot_model()` function from the sjPlot package. The CPH model can be passed as the sole argument to visualize the hazard ratios.
-
-#### Kaplan-Meier Survival Curve
-
-After generating the CPH model, a Kaplan-Meier survival curve can be generated to view survival probabilities against time.
-These plots are important when modeling survival data because they clearly demonstrate the time where the event being modeled for is expected to occur.
-
-The survival curve will be plotted by using the survival package's `survfit()` function. The object created will be plotted using `plot()`
+Tim until event can be modeled in R by passing the `type = "lp"` argument in the `stats::predict()` function to get a value for linear predictor.
 
@@
 
-### Data Exploration and Visualization
-
-Though CPH is typically used with health data, we wanted to determine
-how CPH can be used in other fields. As such, we will study the
-remaining useful life (RUL) of a NASA aircraft engine using data
-provided by NASA for the study on propagation modeling [@saxena2008]. In
-particular, we will be examining their data from what they have deemed
-engine two. From the literature review, we have noted that we will need
-to be concerned with the constant hazards ratio within our data and
-determine how we will accommodate censored data when creating our model.
-
-Each engine has an unknown amount of wear, manufacturing variation, and
-sensor noise all of which impact the survival time. In the training
-dataset, each engine is operating normally at the beginning of the time
-series, develops a fault, and experiences failure at the end of the time
-series. The conditions the engines are subjected to are represented by
-three operational setting fields and 21 sensor measurement fields. In
-the testing data, the engines have not experienced failure
-[@saxena2008].
-
-```{r, echo = FALSE}
-#packages
+```{r, echo = FALSE, warning=FALSE,include=FALSE}
+#Load packages
 library(tidyverse)
@@
 
-When using Cox proportional hazards models to model survival data, the
-data must include a value for time, a unique identifier column
-indicating which individual the time series is for, covariates used for
-predictions, and an event indicating the status of the individual at
-each point in time. The data used in this analysis contain all of the
-previously mentioned fields with the exception of a status field.
+### Data Structure, Exploration, and Visualization
+
+Though CPH is typically used with health data, we wanted to determine
+how CPH can be used in other fields. As such, we will study the
+remaining useful life (RUL) of a NASA aircraft engine using data
+provided by NASA for the study on propagation modeling [@saxena2008]. In
+particular, we will be examining their data from what they have deemed
+engine two. From the literature review, we have noted that we will need
+to be concerned with the constant hazards ratio within our data and
+determine how we will accommodate censored data when creating our model.
+
+#### Data Structure
+
+Each engine has an unknown amount of wear, manufacturing variation, and
+sensor noise all of which impact the survival time. In the training
+dataset, each engine is operating normally at the beginning of the time
+series, develops a fault, and experiences failure at the end of the time
+series. The conditions the engines are subjected to are represented by
+three operational setting fields and 21 sensor measurement fields. In
+the testing data, the engines have not experienced failure
+[@saxena2008].
 
@@
 
+#### Data Visualization
+
 The visuals below describe the distribution of the 260 engines in the
 training data. The histogram shows the survival times are normally
-distributed with a median value of 199 iterations and the boxplot shows
+distributed with a median value of 199 iterations and the box plot shows
 there are few outliers out of the 260 engines. The engines have a median
@@
 
+### Creating Cox Proportional Hazards Model
+
+#### Choosing Covariates
+
 Below is a univariate test to determine which covariates to use in the model
@@
 
-### Modeling
-
-Creating model
+#### Creating Model
 
@@
 
-**Checking Proportional Hazards**
+#### Checking Assumptions
+
+**Checking Proportional Hazards Assumption**
 
@@
 ```
+
+**Checking Independent Censoring**
+
+In the dataset being used for this model, the training data does not
+contain censored data. The documentation available at the dataset's
+source clarifies all engines in the training data have experienced the
+desired event of failure.
+
+### Model Evaluation
+
+### Model Visualization
+
+Forest Plot
+
+```{r}
+plot_model(cox,dot.size = 4,line.size = 2,colors = "red")+
+  theme_bw()+
+  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
+  ylab(label = "Covariates")+
+  xlab(label = "Estimated Effect on Hazard Ratio")
+```
+survival curve
+
+```{r}
+survfit <- survfit(cox, newdata = test)
+plot(survfit, 
+     col = "blue", 
+     conf.int = TRUE, 
+     mark.time = FALSE,
+     xlab = "Time (Iterations)",
+     ylab = "Survival Probability",
+     main = "Survival Probability Over Time")
+```
+
+### Model Applications and Predictions
 
@@
 
-Forest Plot
-
-```{r}
-plot_model(cox,dot.size = 4,line.size = 2,colors = "red")+
-  theme_bw()+
-  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
-  ylab(label = "Covariates")+
-  xlab(label = "Estimated Effect on Hazard Ratio")
-```
-survival curve
-
-```{r}
-survfit <- survfit(cox, newdata = test)
-plot(survfit, 
-     col = "blue", 
-     conf.int = TRUE, 
-     mark.time = FALSE,
-     xlab = "Time (Iterations)",
-     ylab = "Survival Probability",
-     main = "Survival Probability Over Time")
-```
-
 Probability Calculation: $S(t) = e^{-H(t)}$
 $H(t)$ is the cumulative hazard up to time $t$
-
-**Checking Independent Censoring**
-
-In the dataset being used for this model, the training data does not
-contain censored data. The documentation available at the dataset's
-source clarifies all engines in the training data have experienced the
-desired event of failure.
 

### 2025-03-10 index.qmd
3/10 updats

@@
 [Literature](articleSummaries.html)
+
+```{r, echo = FALSE, warning=FALSE, include=FALSE}
+#Load packages
+library(tidyverse)
+library(gridExtra)
+library(survival)
+library(survminer)
+library(ggpubr)
+library(magrittr)
+library(ggfortify)
+library(knitr)
+library(sjPlot)
+
+```
 
@@
 $\alpha = 0.05$ and the global p-value, the determination of whether to
-reject or fail to reject $H_0$ can be made [@nahhas2025].
+reject or fail to reject $H_0$ can be made.
+In addition to the hypothesis test, these Shoenfeld residuals can be plotted against time for each covariate to visualize the assumption.
+When plotting the residuals against time, if the assumption is not broken it would be expected that the points show roughly a straight horizontal line to indicate constant hazard over time [@nahhas2025].
 
@@
 interpretation of the p-values generated from `cox.zph()`. If there is no clear slope in the residuals on the plot, it signifies further confirmation the proportional hazards assumption is met.
+
+Below is an example plot demonstrating Shoenfeld residuals plotted against time to test the proportional hazards assumption.
+While there is some curvature in the line, the residuals are roughly horizontal indicating proportional hazard over time for this covariate.
+In a model with multiple covariates, each one would need to be visualized to ensure it does not break this assumption and can be used in the model.
+
+```{r, echo=FALSE}
+example_surv_object <- Surv(time = lung$time, event = lung$status)
+
+example_coxmodel <- coxph(example_surv_object ~ meal.cal + age, data = lung)
+
+example_phtest <- cox.zph(example_coxmodel)
+
+plot(example_phtest[1],
+     ylab = "Covariate Schoenfeld Residual",
+     xlab = "Time")
+```
 
@@
 If the resulting residuals are linear and appear to have a slope of zero, this assumption is not violated.[@Amini2015]
+
+Below is an example plot showing Martingale residuals plotted against the range of values in the covariate being tested.
+From the plot, it is clear there is no clear slope in the residuals indicating the linearity assumption is not broken.
+For a model with multiple covariates, multiple plots will need to be made.
+
+```{r,echo=FALSE}
+example_martingale <- resid(example_coxmodel, type = "martingale")
+
+plot(example_martingale, 
+     main = "Martingale Residuals: Example Covariate",
+     ylab = "Martingale Residuals for Example Covariate",
+     xlab = "Value Range for Example Covariate")
+abline(h = 0, col = "red")
+```
 
@@
 
-A forest plot will be created using the `plot_model()` function from the sjPlot package. The CPH model can be passed as the sole argument to visualize the hazard ratios.
+A forest plot will be created using the `plot_model()` function from the sjPlot package.
+The CPH model can be passed as the sole argument to visualize the hazard ratios.
+An example forest plot is shown below plotting the hazard ratio for each covariate in the model.
+The point represents the calculated hazard ratio and the whiskers show the 95% confidence interval.
+
+```{r, echo=FALSE,warning=FALSE}
+plot_model(example_coxmodel,
+           dot.size = 1,
+           line.size = 1,
+           colors = "red")+
+  theme_bw()+
+  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
+  ylab(label = "Covariates")+
+  xlab(label = "Estimated Effect on Hazard Ratio")
+```
 
@@
 These plots are important when modeling survival data because they clearly demonstrate the time where the event being modeled for is expected to occur.
+'
+The survival curve will be plotted by using the survival package's `survfit()` function. The object created will be plotted using `plot()`
 
-The survival curve will be plotted by using the survival package's `survfit()` function. The object created will be plotted using `plot()`
+Below is an example survival curve showing survival probabilities over time.
+The line represents the median survival probability with the red area around it showing the 95% confidence interval for the values.
+The curve can be interpreted as showing the probability of event would be expected to happen around a time of 300 where the survival probability is below 50%.
+
+```{r, echo=FALSE}
+example_curve <- survfit(example_coxmodel)
+
+ggsurvplot(example_curve,
+           data = lung,
+           title = "Example Survival Curve",
+           xlab = "Time",
+           ylab = "Survival Probability",
+           legend = "none")
+
+```
 
@@
 Cumulative hazard is used as an additional measure for risk, but it is mainly used to calculate survival probability to plot against time on survival curves.
+[@klein2005]
 
 The cumulative hazard will be calculated in R by grouping the data by the id field and finding the summation of the hazard rates for the individual through the duration of the observation period.
-A column will be added showing the cumulative hazard for each individual at ach point in time.
+A column will be added showing the cumulative hazard for each individual at each point in time.
 
@@
 A survival probability of .05 indicates that only 5% of individuals under the same conditions would survive.
-Survival probability is an important metric when modeling, visualizing, and making predictions with survival data.
+Survival probability is an important metric when modeling, visualizing, and making predictions with survival data.[@klein2005]
 
@@
 ## Analysis and Results
-
-```{r, echo = FALSE, warning=FALSE,include=FALSE}
-#Load packages
-library(tidyverse)
-library(gridExtra)
-library(survival)
-library(survminer)
-library(ggpubr)
-library(magrittr)
-library(ggfortify)
-library(knitr)
-library(sjPlot)
-
-```
 
@@
 ```{r}
-survfit <- survfit(cox, newdata = test)
-plot(survfit, 
-     col = "blue", 
-     conf.int = TRUE, 
-     mark.time = FALSE,
-     xlab = "Time (Iterations)",
-     ylab = "Survival Probability",
-     main = "Survival Probability Over Time")
+survfit <- survfit(cox)
+
+ggsurvplot(survfit, data = test)
 ```

### 2025-03-19 index.qmd
UPdates

@@
 
+### Mathematical Formulas
+
+There are two functions that are the backbone for CPH: the survival
+function and the hazard function. The survival function when $X$ is a
+continuous random variable is
+
+$$
+S(x) = 1 - F(x) = 1 - Pr(X > x) = \int_x^\infty f(x) dx
+$$ {#eq-survival}
+
+where $S(x)$ is the probability that an individual has survived past
+time $x$ [@klein2005]. undefined
+
+The hazard function, which is also known as the hazard rate function,
+can be defined as
+
+$$
+ h(x) = \lim_{\Delta x \rightarrow 0} \frac{P[x\leq X < x + \Delta x | X \geq x]}{\Delta x}
+$$ {#eq-hazardrate}
+
+The function is used to describe the changes in [@klein2005].
+
+Since CPH is a proportional hazard model, the hazard function is adapted
+to deal with the relationship between time to event and the explanatory
+variables [@klein2005]. It can be defined as
+
+$$
+h(t|\mathbf{Z} ) = h_0(t)\text{exp}(\sum_{k=1}^p \beta_kZ_k)
+$$ {#eq-coxhazard}
+
+where,
+
+-   $h(t|\mathbf{Z})$ is the hazard rate at time $t$ for an subject with
+    risk vector $\mathbf{Z}$
+-   $h_0(t)$ is a baseline hazard rate when all covariates are $0$
+-   $\text{exp}(\sum_{k=1}^p \beta_k Z_k)$ is a semi-parametric model
+-   $\beta_k$ is a parameter vector
+-   $Z_k$ is the risk vector [@klein2005]
+
+CPH is considered a proportional hazards model due to the hazard ratio
+between two individuals being a constant value. The ratio of the hazard
+function for an individual with $\mathbf{Z}$ covariates where
+$Z_k = z+1$ and for an individual with $\mathbf{Z}^*$ covariates where
+$Z_k = z$ with all other predictors are fixed [@nahhas2025] is defined
+as
+
+$$
+\frac{h(t|\mathbf{Z})}{h(t|\mathbf{Z}^*)} = \frac{h_0(t)\text{exp}[\sum_{k=1}^p \beta_k Z_k]}{h_0(t)\text{exp}[\sum_{k=1}^p \beta_k Z^*_k]} = \text{exp}[\sum_{k=1}^p \beta_k(Z_k - Z_k^*)]
+$$ {#eq-hazardratio}[@klein2005]. The result of the hazard ratio (HR) is
+a constant that implies an association between a continuous predictor
+and the outcome. The interpretation of the HR is
+
+-   if HR $> 1$ implies a positive association such that
+    $100\% \times (HR - 1)$ is the \\% of greater risk for $Z = z+1$
+    when compared to $Z = z$.
+-   if HR $< 1$ implies a negatie association such that
+    $100\% \times (HR -1)$ is the \\% of decreased risk for $Z = z+1$
+    when compared to $Z= z$.
+-   if HR $=1$ implies no association. [@nahhas2025]
+
+An additional useful equations is the cumulative hazard function. The
+cumulative hazard function can be defined as
+
+$$
+ H(x) = \int_0^x h(u) du 
+$$ {#eq-cumulativehr}
+
+where $h(u)$ is the hazard rate function [@klein2005]. This function
+provides a better understanding of the survival function for graphing
+purposes.
+
 ### Data Structure
 
-In order to use CPH to model time to event survival data, the data must contain a value for time, a binary field indicating the event being modeled, an id field, and one or multiple covariate fields.
-The time value is typically formatted as a time series for each individual in the dataset.
-Individuals will have a time value from the beginning of the observation period, to an arbitrary time that may or may not be the time the event being modeled for occurs.
-Data is considered censored when the individual is not given the opportunity to experience the event in the observation period and is handled differently by the model.
-The covariate fields in the dataset can be categorical or numerical.
+In order to use CPH to model time to event survival data, the data must
+contain a value for time, a binary field indicating the event being
+modeled, an id field, and one or multiple covariate fields. The time
+value is typically formatted as a time series for each subject in the
+dataset. Subjects will have a time value from the beginning of the
+observation period, to an arbitrary time that may or may not be the time
+the event being modeled for occurs. Data is considered censored when the
+subject is not given the opportunity to experience the event in the
+observation period and is handled differently by the model. The
+covariate fields in the dataset can be categorical or numerical
 [@nahhas2025].
@@
 
-After verifying the survival data is formatted properly, the CPH model can be created in R.
-Important steps to consider are choosing covariates and verifying the model meet the model's assumptions after the model is created.
+After verifying the survival data is formatted properly, the CPH model
+can be created in R. Important steps to consider are choosing covariates
+and verifying the model meet the model's assumptions after the model is
+created.
 
@@
 #### Generating Model
-To generate the CPH model with properly formatted survival data, a survival object must first be created using the R package `survival`.
-The survival object is created by running `Surv(time, event)` where `time` is the time field and `event` is the event field.
+
+To generate the CPH model with properly formatted survival data, a
+survival object must first be created using the R package `survival`.
+The survival object is created by running `Surv(time, event)` where
+`time` is the time field and `event` is the event field.
 
 To create the CPH model, the survival object is then applied to the R
-function `coxph(survival object ~ x, data)` where `x` represents one or multiple covariate(s) and `data` is the survival data that is being modeled.[@R-base].
+function `coxph(survival object ~ x, data)` where `x` represents one or
+multiple covariate(s) and `data` is the survival data that is being
+modeled.[@R-base].
 
-During the analysis, multiple models will be created with a different combination of covariates.
+During the analysis, multiple models will be created with a different
+combination of covariates.
 
@@
 
-When creating a CPH model, to ensure an accurate model is created the following assumptions must be met. If
-they are not met, the data must either be transformed, stratified, or other covariates may need to be selected for the model. The assumptions associated with the CPH model are defined below.
+When creating a CPH model, to ensure an accurate model is created the
+following assumptions must be met. If they are not met, the data must
+either be transformed, stratified, or other covariates may need to be
+selected for the model. The assumptions associated with the CPH model
+are defined below.
 
@@
 proportional hazards assumption states that the ratio of hazard rates
-for any two individuals must be constant over time regardless of the
-point in time of the measurement. This means the effect
-of the covariates must be constant over time. [@Bustan2018]
+for any two subjects must be constant over time regardless of the point
+in time of the measurement. This means the effect of the covariates must
+be constant over time. [@Bustan2018]
 
-This assumption can be tested using the R package `survival`. The package contains a
-proportional hazards test called `cox.zph()` [@R-base]. Using the
-hypothesis,
+This assumption can be tested using the R package `survival`. The
+package contains a proportional hazards test called `cox.zph()`
+[@R-base]. Using the hypothesis,
 
@@
 $\alpha = 0.05$ and the global p-value, the determination of whether to
-reject or fail to reject $H_0$ can be made.
-In addition to the hypothesis test, these Shoenfeld residuals can be plotted against time for each covariate to visualize the assumption.
-When plotting the residuals against time, if the assumption is not broken it would be expected that the points show roughly a straight horizontal line to indicate constant hazard over time [@nahhas2025].
+reject or fail to reject $H_0$ can be made. In addition to the
+hypothesis test, these Shoenfeld residuals can be plotted against time
+for each covariate to visualize the assumption. When plotting the
+residuals against time, if the assumption is not broken it would be
+expected that the points show roughly a straight horizontal line to
+indicate constant hazard over time [@nahhas2025].
 
 Individual plots for each covariate are generated which provide a visual
-interpretation of the p-values generated from `cox.zph()`. If there is no clear slope in the residuals on the plot, it signifies further confirmation the proportional hazards assumption is met.
+interpretation of the p-values generated from `cox.zph()`. If there is
+no clear slope in the residuals on the plot, it signifies further
+confirmation the proportional hazards assumption is met.
 
-Below is an example plot demonstrating Shoenfeld residuals plotted against time to test the proportional hazards assumption.
-While there is some curvature in the line, the residuals are roughly horizontal indicating proportional hazard over time for this covariate.
-In a model with multiple covariates, each one would need to be visualized to ensure it does not break this assumption and can be used in the model.
+Below is an example plot demonstrating Shoenfeld residuals plotted
+against time to test the proportional hazards assumption. While there is
+some curvature in the line, the residuals are roughly horizontal
+indicating proportional hazard over time for this covariate. In a model
+with multiple covariates, each one would need to be visualized to ensure
+it does not break this assumption and can be used in the model.
 
@@
 
-The linearity assumption states that the relationship between covariates and the log of the hazard rate must be linear.
-This assumption can be checked by plotting Martingale residuals against the range of values present in each covariate.
+The linearity assumption states that the relationship between covariates
+and the log of the hazard rate must be linear. This assumption can be
+checked by plotting Martingale residuals against the range of values
+present in each covariate.
 
-Martingale residuals are a type of residuals used in survival analysis that show the discrepancy between the observed and expected number of events. 
-$MartingaleResiduals = ObservedEvents - ExpectedEvents$
+Martingale residuals are a type of residuals used in survival analysis
+that show the discrepancy between the observed and expected number of
+events. $MartingaleResiduals = ObservedEvents - ExpectedEvents$
 
-Residuals can first be found by using the `resid(type = "martingale")` function and passing martingale in the type argument creating a vector of residuals that can be plotted against the range of values for each covariate.
+Residuals can first be found by using the `resid(type = "martingale")`
+function and passing martingale in the type argument creating a vector
+of residuals that can be plotted against the range of values for each
+covariate.
 
-If the resulting residuals are linear and appear to have a slope of zero, this assumption is not violated.[@Amini2015]
+If the resulting residuals are linear and appear to have a slope of
+zero, this assumption is not violated.[@Amini2015]
 
-Below is an example plot showing Martingale residuals plotted against the range of values in the covariate being tested.
-From the plot, it is clear there is no clear slope in the residuals indicating the linearity assumption is not broken.
-For a model with multiple covariates, multiple plots will need to be made.
+Below is an example plot showing Martingale residuals plotted against
+the range of values in the covariate being tested. From the plot, it is
+clear there is no clear slope in the residuals indicating the linearity
+assumption is not broken. For a model with multiple covariates, multiple
+plots will need to be made.
 
@@
 
-This assumption describes how if an event is censored it should not be considered when
-the model is being created. Censored data is difficult to accommodate when
-Cox proportional hazards models are used in other industries, such as
-the medical field, because observation times for the individuals
-involved in studies may have decades before an event is observed and the event may not have the opportunity to occur during the study period. In
-this scenario, it would not be appropriate to consider this individual
-when modeling survival.
+This assumption describes how if an event is censored it should not be
+considered when the model is being created. Censored data is difficult
+to accommodate when Cox proportional hazards models are used in other
+industries, such as the medical field, because observation times for the
+subjects involved in studies may have decades before an event is
+observed and the event may not have the opportunity to occur during the
+study period. In this scenario, it would not be appropriate to consider
+this subject when modeling survival.
 
@@
 
-Survival times of observed individuals should be independent of each
-other.
-In an example where survival of plane engines were being evaluated, if each plane has two engines and the failure of one engine caused the accelerated failure of another due to increased workload, it would produce biased conclusions and violate the independence assumption.
+Survival times of observed subjects should be independent of each other.
+In an example where survival of plane engines were being evaluated, if
+each plane has two engines and the failure of one engine caused the
+accelerated failure of another due to increased workload, it would
+produce biased conclusions and violate the independence assumption.
 
@@
 
-To evaluate the model, the Cox regression concordance index can be
-used. Two observed pairs are said to be concordant if an event that is
+To evaluate the model, the Cox regression concordance index can be used.
+Two observed pairs are said to be concordant if an event that is
 expected to occur before another does occur first. For example, if two
@@
 
-There are multiple different methods to visualize data when creating a  Cox proportional hazards model.
-Survival curves are a popular way to show survival probabilities over time, but there are other visualizations that are helpful when interpreting the results of the model.
+There are multiple different methods to visualize data when creating a
+Cox proportional hazards model. Survival curves are a popular way to
+show survival probabilities over time, but there are other
+visualizations that are helpful when interpreting the results of the
+model.
 
 #### Forest Plot
-Forest plots can be used to visualize the effects of each covariate in the model on the hazard ratio.
-The resulting plot contains each covariate on the y axis, with the estimated effect on the hazard ratio on the the x axis.
-The plot contains whiskers demonstrating a 95% confidence interval for the effect.
-A positive effect indicates a positive correlation with the hazard with a higher number signifying a more severe effect on the hazard ratio.
 
-A forest plot will be created using the `plot_model()` function from the sjPlot package.
-The CPH model can be passed as the sole argument to visualize the hazard ratios.
-An example forest plot is shown below plotting the hazard ratio for each covariate in the model.
-The point represents the calculated hazard ratio and the whiskers show the 95% confidence interval.
+Forest plots can be used to visualize the effects of each covariate in
+the model on the hazard ratio. The resulting plot contains each
+covariate on the y axis, with the estimated effect on the hazard ratio
+on the the x axis. The plot contains whiskers demonstrating a 95%
+confidence interval for the effect. A positive effect indicates a
+positive correlation with the hazard with a higher number signifying a
+more severe effect on the hazard ratio.
+
+A forest plot will be created using the `plot_model()` function from the
+sjPlot package. The CPH model can be passed as the sole argument to
+visualize the hazard ratios. An example forest plot is shown below
+plotting the hazard ratio for each covariate in the model. The point
+represents the calculated hazard ratio and the whiskers show the 95%
+confidence interval.
 
@@
 
-After generating the CPH model, a Kaplan-Meier survival curve can be generated to view survival probabilities against time.
-These plots are important when modeling survival data because they clearly demonstrate the time where the event being modeled for is expected to occur.
-'
-The survival curve will be plotted by using the survival package's `survfit()` function. The object created will be plotted using `plot()`
+After generating the CPH model, a Kaplan-Meier survival curve can be
+generated to view survival probabilities against time. These plots are
+important when modeling survival data because they clearly demonstrate
+the time where the event being modeled for is expected to occur. ' The
+survival curve will be plotted by using the survival package's
+`survfit()` function. The object created will be plotted using `plot()`
 
-Below is an example survival curve showing survival probabilities over time.
-The line represents the median survival probability with the red area around it showing the 95% confidence interval for the values.
-The curve can be interpreted as showing the probability of event would be expected to happen around a time of 300 where the survival probability is below 50%.
+Below is an example survival curve showing survival probabilities over
+time. The line represents the median survival probability with the red
+area around it showing the 95% confidence interval for the values. The
+curve can be interpreted as showing the probability of event would be
+expected to happen around a time of 300 where the survival probability
+is below 50%.
 
@@
 
-Cox proportional hazards models can produce an assortment of useful metrics to interpret the model.
-The model created can be used to estimate risk of the event occurring, rate of occurrence, survival probability, and time to event.
+Cox proportional hazards models can produce an assortment of useful
+metrics to interpret the model. The model created can be used to
+estimate risk of the event occurring, rate of occurrence, survival
+probability, and time to event.
 
-#### Hazard Ratio
-
-CPH models focus on providing a relative risk of the event happening based on the covariates and value for time.
-The hazards ratio is used to evaluate this level of risk and is calculated by finding the ratio between two individuals where one is the baseline value of a covariate at time $t$, and the other is the value of a covariate with a one unit increase.
-The hazards ratio is represented by the formula below.
-
-$$ \frac{h(t|\mathbf{Z})}{h(t|\mathbf{Z}^*)} = \frac{h_0(t)\text{exp}(\Sigma_{k=1}^{p} \beta_kZ_k)}{h_0(t)\text{exp}(\Sigma_{k=1}^{p} \beta_kZ_k^*)} = \text{exp}[\sum_{k=1}^p \beta_k(Z_k-Z_k^*)] = \text{exp}(\sum_{k=1}^{p} \beta_kZ_k) $$
-
-where,
-
--   $\mathbf{Z}$ and $\mathbf{Z}^*$ are covariate values.
-
-A hazards ratio can be interpreted as:
-$h(t|\mathbf{Z})$ < 1 indicates a decreased baseline risk as compared to the baseline.
-$h(t|\mathbf{Z})$ > 1 indicates an increased baseline risk as compared to the baseline.
-
-Hazard ratios can be interpreted to give an estimation of the event occurring.
-For example, a hazard ratio of 1.3 signifies that for a one unit increase in the covariate, a 30% increase in risk of the event occurring is expected.
-Hazard ratios can also be used to calculate hazard rates, which have theiur own applications and interpretations when modeling survival data.
-[@klein2005]
-
-#### Hazard Rate
-
-The hazard rate is defined as:
-
-$$ h(t|\mathbf{Z}) = h_0(t)\text{exp}({\sum_{k=1}^{p} \beta_kZ_k}) $$
-
-where,
-
--   $h(t|\mathbf{Z})$ is the hazard rate at time $t$ for an individual
-    with risk vector $\mathbf{Z}$
-
--   $\text{exp}(\sum_{k=1}^{p} \beta_kZ_k)$ is the hazard ratio as explained above.
-
--   $h_0(t)$ is a baseline hazard rate indicating the hazard rate when all covariates are zero.
-
-The resulting hazard rate is a measure of the instantaneous risk of the event happening at a specific point in time. Similar to the hazard ratio, hazard rates are interpreted as:
-
-$h(t|\mathbf{Z})$ < 1 indicates a lower risk of the event occurring at the specific point in time.
-$h(t|\mathbf{Z})$ > 1 indicates a higher risk of the event occurring at the specific point in time.
-
-Hazard rates can be used as an additional measure of risk, as well as to calculate the cumulative hazard for an individual. [@klein2005]
-
-The hazard rate can be calculated in R by using `stats::predict()` after creating a CPH model. This function will be used to create a column showing the hazard rate for each row in the testing data.
-
-#### Cumulative Hazard
-
-Cumulative hazard represents the total accumulated risk of an individual for the duration of the observation period at time $t$ and is defined as:
-
-$H(t) = \int_0^t h(t) dt$
-where,
-
--   $H(t)$ is the cumulative hazard at time $t$
-
--   $h(t)$ is the hazard rate at time t
-
-As defined by the formula, cumulative hazard is a summation of the hazard rates for an individual from time 0 to time $t$.
-A higher cumulative hazard indicates a higher risk of event.
-Cumulative hazard is used as an additional measure for risk, but it is mainly used to calculate survival probability to plot against time on survival curves.
-[@klein2005]
-
-The cumulative hazard will be calculated in R by grouping the data by the id field and finding the summation of the hazard rates for the individual through the duration of the observation period.
-A column will be added showing the cumulative hazard for each individual at each point in time.
-
-#### Calculating Survival Probability
-
-$S(t) = exp(-H(t))$
-
-where,
-
--   $S(t)$ is the probability of survival for an individual at time $t$
--   $H(t)$ is the cumulative hazard for an individual at time $t$
-
-The resulting survival probability will be between 0 and 1 with a value close to one indicating a high probability of the individual not experiencing the event.
-Conversely, a value close to 0 indicates a higher probability of the individual experiencing the event.
-A survival probability of .05 indicates that only 5% of individuals under the same conditions would survive.
-Survival probability is an important metric when modeling, visualizing, and making predictions with survival data.[@klein2005]
-
-Survival probabilities can be calculated in R by using `stats::predict()` after creating a CPH model.
-This function will be used to create a column showing the survival probability for each row in the testing data.
+Survival probabilities can be calculated in R by using
+`stats::predict()` after creating a CPH model. This function will be
+used to create a column showing the survival probability for each row in
+the testing data.
 
@@
 
-Time until event can be calculated by modeling survival probability for future time values.
-When survival probability drops below 50%, it will be assumed the event occurred.
-Predictions for time until event will be made on the testing data for each individual's maximum time value.
-A column will be added to the table showing the expected time until event.
+Time until event can be calculated by modeling survival probability for
+future time values. When survival probability drops below 50%, it will
+be assumed the event occurred. Predictions for time until event will be
+made on the testing data for each individual's maximum time value. A
+column will be added to the table showing the expected time until event.
 
-Tim until event can be modeled in R by passing the `type = "lp"` argument in the `stats::predict()` function to get a value for linear predictor.
+Tim until event can be modeled in R by passing the `type = "lp"`
+argument in the `stats::predict()` function to get a value for linear
+predictor.
 
@@
 
-Below is a univariate test to determine which covariates to use in the model
+Below is a univariate test to determine which covariates to use in the
+model
 
@@
 ```
+
 survival curve
@@
 
-Hazard Rate Calculation: $h(t|Z) = h_0(t) * e^{B_1Z_1 + B_2Z_2 + B_3Z_3+B_4Z_4+B_5Z_5}$
+Hazard Rate Calculation:
+$h(t|Z) = h_0(t) * e^{B_1Z_1 + B_2Z_2 + B_3Z_3+B_4Z_4+B_5Z_5}$
 
@@
 
-$H(t) = \int_0^t h(t) dt$
-where,
-$H(t)$ is the cumulative hazard from time 0 to time $(t)$
-$h(t)$ is the hazard rate at time $t$
+$H(t) = \int_0^t h(t) dt$ where, $H(t)$ is the cumulative hazard from
+time 0 to time $(t)$ $h(t)$ is the hazard rate at time $t$
 
@@
 
-Probability Calculation: $S(t) = e^{-H(t)}$
-$H(t)$ is the cumulative hazard up to time $t$
+Probability Calculation: $S(t) = e^{-H(t)}$ $H(t)$ is the cumulative
+hazard up to time $t$
 

### 2025-03-21 index.qmd
updates

@@
 individual in a study is only known for a certain period of time, such
-as the length of the study [@klein2005, 1]. The models also have the
+as the length of the study [@klein2005]. The models also have the
 advantage of being semi-parametric which allows more flexibility as
@@
 
-An additional useful equations is the cumulative hazard function. The
+An additional useful equation is the cumulative hazard function. The
 cumulative hazard function can be defined as
@@
 contain a value for time, a binary field indicating the event being
-modeled, an id field, and one or multiple covariate fields. The time
-value is typically formatted as a time series for each subject in the
-dataset. Subjects will have a time value from the beginning of the
-observation period, to an arbitrary time that may or may not be the time
-the event being modeled for occurs. Data is considered censored when the
-subject is not given the opportunity to experience the event in the
-observation period and is handled differently by the model. The
-covariate fields in the dataset can be categorical or numerical
-[@nahhas2025].
+modeled, an id field, and one or multiple covariate fields that can be
+categorical or numerical. The time value is typically formatted as a
+time series for each subject in the dataset. The first entry for
+subjects will be a time value from the beginning of the observation
+period with the final entry time varying by subject based on if the
+event occurred for that individual or if the study ended before the
+event occurred. Subjects who did not have the event occur by the end of
+the observation period are considered censored entries and will be
+handleded different in the model [@nahhas2025].
 
-### Creating Cox Proportional Hazards Model
+### Covariate Selection
 
-After verifying the survival data is formatted properly, the CPH model
-can be created in R. Important steps to consider are choosing covariates
-and verifying the model meet the model's assumptions after the model is
-created.
-
-#### Choosing covariates
-
-From the study the data is from, we know that some of the sensors will
-not play a role in the RUL. As such, we will run a univariate test of
-all covariates to determine which ones will actually impact the RUL. We
-will use the p-value provided by the Wald test with a significance level
-of $\alpha = 0.05$.
-
-#### Generating Model
-
-To generate the CPH model with properly formatted survival data, a
-survival object must first be created using the R package `survival`.
-The survival object is created by running `Surv(time, event)` where
-`time` is the time field and `event` is the event field.
-
-To create the CPH model, the survival object is then applied to the R
-function `coxph(survival object ~ x, data)` where `x` represents one or
-multiple covariate(s) and `data` is the survival data that is being
-modeled.[@R-base].
-
-During the analysis, multiple models will be created with a different
-combination of covariates.
+Depending on the study, there will be covariates that do not impact the
+probability of survival. Statistical methods such as the Wald Test can
+be used to determine which ones to use in the final CPH model.
+Generating a univariate CPH model for each covariate and using the Wald
+test with a significance level of $\alpha = 0.05$ will allow for
+determination of if a covariate is significant. Any covariates that are
+determined to be not significant to the survival probability can be
+removed from the final CPH model.
 
@@
 
-When creating a CPH model, to ensure an accurate model is created the
-following assumptions must be met. If they are not met, the data must
-either be transformed, stratified, or other covariates may need to be
-selected for the model. The assumptions associated with the CPH model
-are defined below.
+```{r, echo = FALSE}
+#generation of the model for sample plots
+example_surv_object <- Surv(time = lung$time, event = lung$status)
 
-#### Proportional Hazards Assumption
+example_coxmodel <- coxph(example_surv_object ~ meal.cal + age, data = lung)
+```
 
-The proportional hazards assumption has been proven to be critical to
-evaluate prior to using CPH when modeling survival data. The
-proportional hazards assumption states that the ratio of hazard rates
-for any two subjects must be constant over time regardless of the point
-in time of the measurement. This means the effect of the covariates must
-be constant over time. [@Bustan2018]
+There are four assumptions that a CPH model must meet in order to be an
+accurate model. For any violation of those assumptions, the data will
+need to be either transformed, stratified, or other covariates may need
+to be selected for the model. The four assumptions are the independence
+assumption, non-informative censoring, linearity assumption, and
+proportional hazards assumption.
 
-This assumption can be tested using the R package `survival`. The
-package contains a proportional hazards test called `cox.zph()`
-[@R-base]. Using the hypothesis,
+**Independence Assumption**
+
+CPH assumes that the survival times of observed subjects are independent
+of each other. This is similar to most regression models, though CPH
+does not assume that the residuals of the model be normally distributed
+or have constant variance [@nahhas2025].
+
+**Non-informative Censoring Assumption**
+
+CPH assumes that censoring is non-informative which means that each
+subject has the same risk of experiencing the event regardless of one of
+them being censored. Thus, the knowledge of censoring does not result in
+any new information being provided [@nahhas2025].
+
+**Linearity Assumption**
+
+CPH assumes the relationship between covariates and the outcome is a
+linear relationship [@nahhas2025]. For CPH, the outcome for that
+relationship is the log of the hazard rate. This assumption is checked
+visually using the Martingale residuals for each covariate
+[@nahhas2025]. Martingale residuals are a type of residauls used in
+survival analysis that show the discrepancy between the observed and the
+expected number of events. The equation is
+$\text{Martingale Residuals} = \text{Observed Events} - \text{Expected Events}$.
+The resulting residuals are examined visually and if they are linear and
+appear to have a slope of zero, the linearity assumption is not violated
+[@Amini2015].
+
+Below is an example plot showing Martingale residuals plotted for a
+covariate being tested. From the plot, it is visually clear that the
+residuals are linear and appear to have a slope of zero. For this
+covariate then, the linearity assumption has not been violated. If a
+model is being generated with multiple covariates, multiple Martingale
+residual plots will need to be generated in order to verify this
+assumption for each covariate.
+
+```{r,echo=FALSE}
+example_martingale <- resid(example_coxmodel, type = "martingale")
+
+plot(example_martingale, 
+     main = "Martingale Residuals: Example Covariate",
+     ylab = "Martingale Residuals for Example Covariate",
+     xlab = "Value Range for Example Covariate")
+abline(h = 0, col = "red")
+```
+
+**Proportional Hazards Assumption**
+
+The major assumption for CPH is the proportional hazards assumption
+which states that the ratio of hazard rates for any two subjects must be
+constant at all times. This means the effect of the covariates must be
+constant over time [@Bustan2018]. To test this assumption, the
+hypothesis test below is used along with Schoenfeld partial residuals.
 
@@
 
-```{r, echo=FALSE}
-example_surv_object <- Surv(time = lung$time, event = lung$status)
-
-example_coxmodel <- coxph(example_surv_object ~ meal.cal + age, data = lung)
-
-example_phtest <- cox.zph(example_coxmodel)
+```{example_phtest <- cox.zph(example_coxmodel)}
 
@@
 ```
-
-#### Linearity Assumption
-
-The linearity assumption states that the relationship between covariates
-and the log of the hazard rate must be linear. This assumption can be
-checked by plotting Martingale residuals against the range of values
-present in each covariate.
-
-Martingale residuals are a type of residuals used in survival analysis
-that show the discrepancy between the observed and expected number of
-events. $MartingaleResiduals = ObservedEvents - ExpectedEvents$
-
-Residuals can first be found by using the `resid(type = "martingale")`
-function and passing martingale in the type argument creating a vector
-of residuals that can be plotted against the range of values for each
-covariate.
-
-If the resulting residuals are linear and appear to have a slope of
-zero, this assumption is not violated.[@Amini2015]
-
-Below is an example plot showing Martingale residuals plotted against
-the range of values in the covariate being tested. From the plot, it is
-clear there is no clear slope in the residuals indicating the linearity
-assumption is not broken. For a model with multiple covariates, multiple
-plots will need to be made.
-
-```{r,echo=FALSE}
-example_martingale <- resid(example_coxmodel, type = "martingale")
-
-plot(example_martingale, 
-     main = "Martingale Residuals: Example Covariate",
-     ylab = "Martingale Residuals for Example Covariate",
-     xlab = "Value Range for Example Covariate")
-abline(h = 0, col = "red")
-```
-
-#### Independent Censoring Assumption
-
-This assumption describes how if an event is censored it should not be
-considered when the model is being created. Censored data is difficult
-to accommodate when Cox proportional hazards models are used in other
-industries, such as the medical field, because observation times for the
-subjects involved in studies may have decades before an event is
-observed and the event may not have the opportunity to occur during the
-study period. In this scenario, it would not be appropriate to consider
-this subject when modeling survival.
-
-#### Independence Assumption
-
-Survival times of observed subjects should be independent of each other.
-In an example where survival of plane engines were being evaluated, if
-each plane has two engines and the failure of one engine caused the
-accelerated failure of another due to increased workload, it would
-produce biased conclusions and violate the independence assumption.
 
@@
 
+To generate the CPH model with properly formatted survival data, a
+survival object must first be created using the R package `survival`.
+The survival object is created by running `Surv(time, event)` where
+`time` is the time field and `event` is the event field.
+
+To create the CPH model, the survival object is then applied to the R
+function `coxph(survival object ~ x, data)` where `x` represents one or
+multiple covariate(s) and `data` is the survival data that is being
+modeled.[@R-base].
+
+During the analysis, multiple models will be created with a different
+combination of covariates.
+
 ```{r}

### 2025-03-23 index.qmd
3/23 updates

--- 2025-03-19 index.qmd
@@
 library(gridExtra)
+library(grid)
 library(survival)
@@
 provides a better understanding of the survival function for graphing
-purposes.
+purposes.[@Amini2015]
 
@@
 
-To generate the CPH model with properly formatted survival data, a
-survival object must first be created using the R package `survival`.
-The survival object is created by running `Surv(time, event)` where
-`time` is the time field and `event` is the event field.
-
-To create the CPH model, the survival object is then applied to the R
-function `coxph(survival object ~ x, data)` where `x` represents one or
-multiple covariate(s) and `data` is the survival data that is being
-modeled.[@R-base].
-
 During the analysis, multiple models will be created with a different
 combination of covariates.
+Covariates will be considered in the models that pass the covariate test.
+After creating the models, the assumptions described below will be tested.
+Out of the multiple models created, models will also not be considered where the assumptions are not true.
+
+The model used through the rest of the analysis to make predictions will be the model with the best evaluation where the assumptions ar true.
 
@@
 
-This assumption can be tested using the R package `survival`. The
-package contains a proportional hazards test called `cox.zph()`
-[@R-base]. Using the hypothesis,
+This assumption can be tested with a hypothesis test using the hypothesis:
 
@@
 
-the `cox.zph()` calculates the scaled Schoenfeld partial residuals and
+The `cox.zph()` function calculates the scaled Schoenfeld partial residuals and
 returns the $\chi^2$ value, the degrees of freedom, and p-values for
@@
 
-Residuals can first be found by using the `resid(type = "martingale")`
-function and passing martingale in the type argument creating a vector
-of residuals that can be plotted against the range of values for each
-covariate.
-
 If the resulting residuals are linear and appear to have a slope of
@@
 study period. In this scenario, it would not be appropriate to consider
-this subject when modeling survival.
+this subject when modeling survival.[@Amini2015]
 
@@
 accelerated failure of another due to increased workload, it would
-produce biased conclusions and violate the independence assumption.
+produce biased conclusions and violate the independence assumption.[@Amini2015]
 
@@
 
-### Model Visualization
+### Model Visualization and Prediction
 
@@
 Cox proportional hazards model. Survival curves are a popular way to
-show survival probabilities over time, but there are other
+visualize survival probabilities over time, but there are other
 visualizations that are helpful when interpreting the results of the
 model.
+
+Cox proportional hazards models can produce an assortment of useful
+metrics to interpret the model as described by the equations above. The model created can be used to
+estimate risk of the event occurring, rate of occurrence, survival
+probability, and time to event.
 
@@
 
-A forest plot will be created using the `plot_model()` function from the
-sjPlot package. The CPH model can be passed as the sole argument to
-visualize the hazard ratios. An example forest plot is shown below
+An example forest plot is shown below
 plotting the hazard ratio for each covariate in the model. The point
@@
 important when modeling survival data because they clearly demonstrate
-the time where the event being modeled for is expected to occur. ' The
-survival curve will be plotted by using the survival package's
-`survfit()` function. The object created will be plotted using `plot()`
+the time where the event being modeled for is expected to occur.
 
@@
 expected to happen around a time of 300 where the survival probability
-is below 50%.
+is below 50%.[@Kuitunen2021]
 
@@
 
-### Model Applications and Predictions
-
-Cox proportional hazards models can produce an assortment of useful
-metrics to interpret the model. The model created can be used to
-estimate risk of the event occurring, rate of occurrence, survival
-probability, and time to event.
-
-Survival probabilities can be calculated in R by using
-`stats::predict()` after creating a CPH model. This function will be
-used to create a column showing the survival probability for each row in
-the testing data.
-
 #### Predicting Time Until Event
@@
 column will be added to the table showing the expected time until event.
-
-Tim until event can be modeled in R by passing the `type = "lp"`
-argument in the `stats::predict()` function to get a value for linear
-predictor.
 
@@
 have 0s in all cycles except the last cycle of the machine.
+
+Below are tables showing the structure of the training and testing data after the transformations describd above.
 
@@
 Below is a univariate test to determine which covariates to use in the
-model
+model.
 
@@
 
-```{r}
-cox <- coxph(Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5,
-             data = train,
-             id = id)
+To generate the CPH model, a
+survival object must first be created using the R package `survival`.
+The survival object is created by running `Surv(time, event)` where
+`time` is the time field and `event` is the event field.
 
-summary(cox)
-```
-
-#### Checking Assumptions
-
-**Checking Proportional Hazards Assumption**
+To create the CPH model, the survival object is then applied to the R
+function `coxph(survival object ~ x, data)` where `x` represents one or
+multiple covariate(s) and `data` is the survival data that is being
+modeled.[@R-base].
 
 ```{r}
-cox.zph(cox)
+cox1 <- coxph(data = train, 
+             Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)
 
-plot(cox.zph(cox))
+cox2 <- coxph(data = train, 
+             Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5)
+
+cox3 <- coxph(data = train, 
+             Surv(time, status) ~ os1 + os3 + sm1 + sm5 + sm6 + sm7 + sm13 + sm14 + sm19 + sm21)
+
+cox4 <- coxph(data = train, 
+             Surv(time, status) ~ sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)
+
+cox5 <- coxph(data = train, 
+             Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5  + sm7 + sm11 + sm12 + sm13 + sm19 + sm20 )
+
+cox6 <- coxph(data = train, 
+             Surv(time, status) ~ os1  + sm1 + sm2 + sm5  + sm7 + sm11 + sm12 + sm13 + sm19  + sm21)
+
+cox7 <- coxph(data = train, 
+             Surv(time, status) ~ os1  + sm13 + sm14 + sm19 + sm20)
+
+cox8 <- coxph(data = train, 
+             Surv(time, status) ~ os1 + os3 + sm21)
+
+cox9 <- coxph(data = train, 
+             Surv(time, status) ~ os1 + os3 + sm2 + sm5 + sm6 + sm12 + sm13  + sm19 + sm20 + sm21)
+
+cox10 <- coxph(data = train, 
+             Surv(time, status) ~  os3 + sm1 + sm2 + sm5 + sm6  + sm11 + sm12 + sm13  + sm19 + sm20 + sm21)
 
@@
 
-**Checking Linearity Assumption**
+The resulting models and their corresponding concordance index values are below.
 
+```{r}
+concordance_results <- data.frame(row.names = c(
+  "Model 1",
+  "Model 2",
+  "Model 3",
+  "Model 4",
+  "Model 5",
+  "Model 6",
+  "Model 7",
+  "Model 8",
+  "Model 9",
+  "Model 10"),
+  Concordance = c(
+  cox1$concordance[6],
+  cox2$concordance[6],
+  cox3$concordance[6],
+  cox4$concordance[6],
+  cox5$concordance[6],
+  cox6$concordance[6],
+  cox7$concordance[6],
+  cox8$concordance[6],
+  cox9$concordance[6],
+  cox10$concordance[6]))
+
+kable(concordance_results)
+
+```
+
+The analysis will proceed using model 1 because of the high concordance index. The full summary of model 1 is below.
+
+```{r}
+summary(cox1)
+
+```
+
+### Checking Assumptions
+
+Despite the high concordance value for model 1, the assumptions will need to be evaluated before using the model to make predictions.
+If any of the assumptions for the model are not true, the model will need to be modified
+
+#### Checking Proportional Hazards Assumption
+The `cox.zph` function can be used to check the proportional hazards assumption.
+As mentioned above, the function calculates p values for each covariate.
+If the p values are < .05, the null hypothesis is rejected indicating the proportional hazards assumption is true.
+If the proportional hazards assumption is not true for any covariate, the model will be recreated without those covariates.
+
+```{r}
+cox.zph(cox1)
+```
+The output above shows the p values for the hypothesis test.
+
+In addition to the hypothesis test, the Shoenfeld residuals calculated by the `cox.zph()` function are plotted against time below.
+
+```{r}
+# Function to put ph_plots on one visual
+create_ph_plots <- function(coxmodel){
+  # Function takes argument that is a cox model from coxph()
+
+  phresults <- cox.zph(coxmodel)
+
+  # Define dimensions of visual
+  par(mfrow = c(3,5),
+      mar = c(2,2,4,1))
+
+  # Loop through each covariate creating a plot and adding it to the phplot list
+  for (i in 1:(length(phresults$table[,1])-1)){
+    phplots <- list()
+    phplots[i] <- plot(phresults[i], # Create plot for each iteration
+                       main = rownames(phresults$table)[i]) # Title of each plot is the rowname from coxmodel$table
+
+  }
+  mtext("Schoenfeld Residuals vs. Time for Covariates", side = 3, line = -1.5, outer = TRUE, cex = 1) # Title for Visual
+  par(mfrow = c(1,1)) # Reset plotting area
+}
+
+create_ph_plots(cox1)
+```
+
+The proportional hazards test shows several covariates do not meet the proportional hazards assumption.
+
+The model is recreated below with another proportional hazards test.
+
+```{r}
+
+cox1.1 <- coxph(data = train, 
+             Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)
+cox.zph(cox1.1)
+
+```
+
+The hypothesis test shows the covariates sm1, sm2, sm6, sm7, sm11, sm12, sm20, and sm21 do not meet the proportional hazards assumption.
+The model is created again below with a second hypothesis test showing all of the covariates pass the hypothesis test.
+
+#### Checking Linearity Assumption
+
+Residuals can first be found by using the `resid(type = "martingale")`
+function and passing martingale in the type argument creating a vector
+of residuals that can be plotted against the range of values for each
+covariate.
 Example for one predictor
@@
 ```{r}
-martingale <- resid(cox, type = "martingale")
+martingale <- resid(cox1.1, type = "martingale")
 
@@
 
-**Checking Independent Censoring**
+#### Checking Independent Censoring
 
@@
 
+A forest plot will be created using the `plot_model()` function from the
+sjPlot package. The CPH model can be passed as the sole argument to
+visualize the hazard ratios.
+
 ```{r}
-plot_model(cox,dot.size = 4,line.size = 2,colors = "red")+
+plot_model(cox1.1,dot.size = 4,line.size = 2,colors = "red")+
   theme_bw()+
@@
 
+ The
+survival curve will be plotted by using the survival package's
+`survfit()` function. The object created will be plotted using `plot()`
 ```{r}
-survfit <- survfit(cox)
+survfit <- survfit(cox1.1)
 
@@
 ```{r}
-predictiondata <- test %>%
-  group_by(id) %>%
-  slice_max(time) %>%
-  select(id, time, os1, os3, sm1, sm2, sm5, status)
+#predictiondata <- test %>%
+  #group_by(id) %>%
+  #slice_max(time) %>%
+  #select(id, time, os1, os3, sm1, sm2, sm5, status)
 
-coef <- data.frame(Variable = c("Z1", "Z2", "Z3", "Z4", "Z5"),
-           CoefficientValue = cox$coefficients)
-kable(coef)
+#coef <- data.frame(Variable = c("Z1", "Z2", "Z3", "Z4", "Z5"),
+           #CoefficientValue = cox1.1$coefficients)
+#kable(coef)
 ```
@@
 ```{r}
-test$hazardrates <- predict(cox,
+test$hazardrates <- predict(cox1.1,
         newdata = test,
@@
 ```{r}
-test$survprob <- predict(cox, type = "survival", newdata = test)
+test$survprob <- predict(cox1.1, type = "survival", newdata = test)
 
@@
 hazard up to time $t$
+
+Tim until event can be modeled in R by passing the `type = "lp"`
+argument in the `stats::predict()` function to get a value for linear
+predictor.
+
+```{r}
+
+```
 

### 2025-03-24 CoxPractice.qmd
Version error fixed.

---
title: "Cox Proportional Hazards Practice"
format: html
editor: visual
self-contained: true
execute: 
  warning: false
  message: false
---

#### General Examination of Data

```{r}
#libraries
library(tidyverse)
library(survival)
library(survminer)
library(ggpubr)
library(magrittr)
library(ggfortify)
```

Dataset is from the Mayo Clinic Primary Biliary Cirrhosis data on Kaggle: <https://www.kaggle.com/datasets/jixing475/mayo-clinic-primary-biliary-cirrhosis-data>

```{r}
pbc <- read.csv('C:/Users/jayre/Desktop/Capstone/pbc.csv')

head(pbc, n = 5)
```

Columns:

-   id: case number
-   time: number of days between registration and the earlier of death, transplantion, or study analysis time in July, 1986
-   status: 0=alive, 1=liver transplant, 2=dead
-   trt: 1= D-penicillamine, 2=placebo
-   age: in days
-   sex: 0=male, 1=female
-   ascites: presence of ascites (0=no 1=yes)
-   hepato: presence of hepatomegaly (0=no 1=yes)
-   spiders: presence of spiders (0=no 1=yes)
-   edema: presence of edema (0=no edema and no diuretic therapy for edema; .5 = edema present without diuretics, or edema resolved by diuretics; 1 = edema despite diuretic therapy)
-   bili: serum bilirubin in mg/dl
-   chol: serum cholesterol in mg/dl
-   albumin: in gm/dl
-   copper: urine copper in ug/day
-   alk.phos: alkaline phosphatase in U/liter
-   ast: SGOT in U/ml
-   trig: triglicerides in mg/dl
-   platelet: per cubic ml / 1000
-   protime: prothrombin time in seconds
-   state: histologic stage of disease

```{r}
#sex needs to be 0,1, need no na's in the treatment column, and don't want the transplant category in status
pbc <- pbc %>% 
        mutate(sex = if_else(sex == "f", 1, 0)) %>%
        na.omit(trt) %>%
        filter(status != 1)

#switch status to be 1 (alive) or 2 (deceased)
pbc <- pbc %>% mutate(status = if_else(status == 0, 1, 2))

head(pbc, n = 5)

```

#### Kaplan-Meier Survival Curves and Log-Rate Tests

```{r}
#Generate the Kaplan-Meier survival curve and plot it with ggplot
#pull out time and status
t <- pbc[, "time"]
s <- pbc[, "status"]

km_fit <- survfit(Surv(t, s) ~ 1)

autoplot(km_fit) + theme_bw()
```

```{r}
#generate the probability of survival at certain times between registration and death 
summary(km_fit, times = c(1:5, 50, 100, 150*(1:10), 300*(5:15)))
```

```{r}
#check km with sex
sex <- pbc[, "sex"]
km_sex <- survfit(Surv(t, s) ~ sex)

autoplot(km_sex) + theme_bw()
```

Can use the log-rank test from the survival test to compare the curves between the two groups.

Hypothesis test:

-   $H_0$: in terms of survivability, there is no difference between the two groups
-   $H_1$: there is a survival differential b/w the two groups

Use $\alpha = 0.05$ as the comparison for the p-value

```{r}
#log rank test to see the impact based on sex
survdiff(Surv(t, s) ~ sex)
```

p-value = $0.03 < \alpha$ so we reject the hull hypothesis as there is enough evidence of a survival difference b/w males and females.

```{r}
pbc %>% summarise(mean = mean(age))
#mean age = 50 so check for <= 50 (0) and > 50 (1) 
pbc <- pbc %>% mutate(age2 = if_else(age > 50, 1, 0))

age <- pbc[, "age2"]

survdiff(Surv(t, s) ~ age)
```

p-value $< 0.005 < \alpha$ so we reject the null hypothesis as there is evidence of a difference in survival between those older than 50 and those younger than 50.

```{r}
#plot curve
km_age <- survfit(Surv(t, s) ~ age)
autoplot(km_age) + theme_bw()
```

```{r}
#check with treatment
trt <- pbc[, "trt"]
survdiff(Surv(t, s) ~ trt)
```

$p = 0.6 > \alpha$ so we fail to reject the null hypothesis as there is insufficient evidence of a difference in survivability b/w treatment groups.

#### Univariate Cox Models

(mostly for practice in understanding the models)

```{r}
#sex
coxph(Surv(time, status) ~ sex, data = pbc) %>%
  summary()
```

log-rank matches the test from above with $p = 0.03$

```{r}
#age
coxph(Surv(time, status) ~ age, data = pbc) %>%
  summary()
```

```{r}
#trt
coxph(Surv(time, status) ~ trt, data = pbc) %>%
  summary()
```

#### Multivariate Models

```{r}
m <- coxph(Surv(time, status) ~ age2 + sex + trt, data = pbc)

m %>% summary()
```

For all three tests (likelihood ratio, Wald, and logrank) $p < .005$ which is $< \alpha = 0.05$ so the overall model is significant and we reject the omnibus hypothesis (all $\beta$s are 0).

In the multivariate world, we can see that trt is not significant as the p-value is $> \alpha = 0.05$. However, age and trt are still significant as their p-values are $< \alpha$.

The HRs are:

-   age: 2.0781
-   sex: 0.6163
-   trt: 0.9650

The HR for age and the lower p-value indicate a strong relationship b/w the patient's age and increased risk of death. Specific wording: holding the other covariates constant, an increased age is associated with a poor survivall.

#### Testing Assumptions

```{r}
#proportional hazards
cox.zph(m)
ggcoxzph(cox.zph(m))
```

All p-values are $> \alpha$ so we do not have any coefficients with that are still time dependent.

```{r}
#non-linearity
m_resid <- residuals(m, type = "martingale")

ggplot(data = pbc, mapping = aes(x = age2, y = m_resid)) + 
  geom_point() + 
  geom_smooth() + 
  theme_bw()
```

```{r}
#use the continuous age for comparison
ggplot(data = pbc, mapping = aes(x = age, y = m_resid)) + 
  geom_point() + 
  geom_smooth() + 
  theme_bw()
```

The continuous version of age would be in violation, but the categorical one is not.

### 2025-03-24 index.qmd
fixed errors

--- 2025-03-21 index.qmd
@@
 where $S(x)$ is the probability that an individual has survived past
-time $x$ [@klein2005]. undefined
+time $x$ [@klein2005].
 
@@
 
-The function is used to describe the changes in [@klein2005].
+The function is used to describe how the chance of experiencing the
+event changes with time and only has the requirement of being
+non-negative [@klein2005].
 
-Since CPH is a proportional hazard model, the hazard function is adapted
-to deal with the relationship between time to event and the explanatory
-variables [@klein2005]. It can be defined as
+CPH then uses both functions to develop it's own model, specifically by
+adapting the hazard function. Since CPH is a proportional hazard model,
+the hazard function is adapted to deal with the relationship between
+time to event and the explanatory variables [@klein2005]. It can be
+defined as
 
@@
 purposes.
-
-### Data Structure
-
-In order to use CPH to model time to event survival data, the data must
-contain a value for time, a binary field indicating the event being
-modeled, an id field, and one or multiple covariate fields that can be
-categorical or numerical. The time value is typically formatted as a
-time series for each subject in the dataset. The first entry for
-subjects will be a time value from the beginning of the observation
-period with the final entry time varying by subject based on if the
-event occurred for that individual or if the study ended before the
-event occurred. Subjects who did not have the event occur by the end of
-the observation period are considered censored entries and will be
-handleded different in the model [@nahhas2025].
 
@@
 survival analysis that show the discrepancy between the observed and the
-expected number of events. The equation is
-$\text{Martingale Residuals} = \text{Observed Events} - \text{Expected Events}$.
-The resulting residuals are examined visually and if they are linear and
-appear to have a slope of zero, the linearity assumption is not violated
-[@Amini2015].
+expected number of events. The equation is $$
+\text{Martingale Residuals} = \text{Observed Events} - \text{Expected Events}
+$$ The resulting residuals are examined visually and if they are linear
+and appear to have a slope of zero, the linearity assumption is not
+violated [@Amini2015]. If a model is being generated with multiple
+covariates, multiple Martingale residual plots will need to be generated
+in order to verify this assumption for each covariate.
 
@@
 residuals are linear and appear to have a slope of zero. For this
-covariate then, the linearity assumption has not been violated. If a
-model is being generated with multiple covariates, multiple Martingale
-residual plots will need to be generated in order to verify this
-assumption for each covariate.
+covariate then, the linearity assumption has not been violated.
 
@@
 
-the `cox.zph()` calculates the scaled Schoenfeld partial residuals and
-returns the $\chi^2$ value, the degrees of freedom, and p-values for
-each individual covariate along with a global value. Using an
-$\alpha = 0.05$ and the global p-value, the determination of whether to
-reject or fail to reject $H_0$ can be made. In addition to the
-hypothesis test, these Shoenfeld residuals can be plotted against time
-for each covariate to visualize the assumption. When plotting the
-residuals against time, if the assumption is not broken it would be
-expected that the points show roughly a straight horizontal line to
-indicate constant hazard over time [@nahhas2025].
+The scaled Schoenfeld partial residuals are calcuated and returns the
+$\chi^2$ value, the degrees of freedom, and p-values for each individual
+covariate along with a global value. Using an $\alpha = 0.05$ and the
+global p-value, the determination of whether to reject or fail to reject
+$H_0$ can be made. In addition to the hypothesis test, these Shoenfeld
+residuals can be plotted against time for each covariate to visualize
+the assumption. When plotting the residuals against time, if the
+assumption is not broken it would be expected that the points show
+roughly a straight horizontal line to indicate constant hazard over time
+[@nahhas2025].
 
@@
 
-### Model Evaluation
+### Evaluation
 
@@
 Cox proportional hazards model. Survival curves are a popular way to
-show survival probabilities over time, but there are other
+visualize survival probabilities over time, but there are other
 visualizations that are helpful when interpreting the results of the
 model.
+
+Cox proportional hazards models can produce an assortment of useful
+metrics to interpret the model as described by the equations above. The
+model created can be used to estimate risk of the event occurring, rate
+of occurrence, survival probability, and time to event.
 
@@
 
-A forest plot will be created using the `plot_model()` function from the
-sjPlot package. The CPH model can be passed as the sole argument to
-visualize the hazard ratios. An example forest plot is shown below
-plotting the hazard ratio for each covariate in the model. The point
-represents the calculated hazard ratio and the whiskers show the 95%
-confidence interval.
+An example forest plot is shown below plotting the hazard ratio for each
+covariate in the model. The point represents the calculated hazard ratio
+and the whiskers show the 95% confidence interval.
 
@@
 have 0s in all cycles except the last cycle of the machine.
+
+Below are tables showing the structure of the training and testing data
+after the transformations describd above.
 
@@
 
-summary(cox)
-```
+cox1 <- coxph(data = train, 
+             Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)
 
-#### Checking Assumptions
+cox2 <- coxph(data = train, 
+             Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5)
 
-**Checking Proportional Hazards Assumption**
+cox3 <- coxph(data = train, 
+             Surv(time, status) ~ os1 + os3 + sm1 + sm5 + sm6 + sm7 + sm13 + sm14 + sm19 + sm21)
 
-```{r}
-cox.zph(cox)
+cox4 <- coxph(data = train, 
+             Surv(time, status) ~ sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)
 
-plot(cox.zph(cox))
+cox5 <- coxph(data = train, 
+             Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5  + sm7 + sm11 + sm12 + sm13 + sm19 + sm20 )
+
+cox6 <- coxph(data = train, 
+             Surv(time, status) ~ os1  + sm1 + sm2 + sm5  + sm7 + sm11 + sm12 + sm13 + sm19  + sm21)
+
+cox7 <- coxph(data = train, 
+             Surv(time, status) ~ os1  + sm13 + sm14 + sm19 + sm20)
+
+cox8 <- coxph(data = train, 
+             Surv(time, status) ~ os1 + os3 + sm21)
+
+cox9 <- coxph(data = train, 
+             Surv(time, status) ~ os1 + os3 + sm2 + sm5 + sm6 + sm12 + sm13  + sm19 + sm20 + sm21)
+
+cox10 <- coxph(data = train, 
+             Surv(time, status) ~  os3 + sm1 + sm2 + sm5 + sm6  + sm11 + sm12 + sm13  + sm19 + sm20 + sm21)
 
@@
 
-**Checking Linearity Assumption**
-
-Example for one predictor
+The resulting models and their corresponding concordance index values
+are below.
 
 ```{r}
-martingale <- resid(cox, type = "martingale")
+concordance_results <- data.frame(row.names = c(
+  "Model 1",
+  "Model 2",
+  "Model 3",
+  "Model 4",
+  "Model 5",
+  "Model 6",
+  "Model 7",
+  "Model 8",
+  "Model 9",
+  "Model 10"),
+  Concordance = c(
+  cox1$concordance[6],
+  cox2$concordance[6],
+  cox3$concordance[6],
+  cox4$concordance[6],
+  cox5$concordance[6],
+  cox6$concordance[6],
+  cox7$concordance[6],
+  cox8$concordance[6],
+  cox9$concordance[6],
+  cox10$concordance[6]))
+
+kable(concordance_results)
+
+```
+
+The analysis will proceed using model 1 because of the high concordance
+index. The full summary of model 1 is below.
+
+```{r}
+summary(cox1)
+
+```
+
+### Checking Assumptions
+
+Despite the high concordance value for model 1, the assumptions will
+need to be evaluated before using the model to make predictions. If any
+of the assumptions for the model are not true, the model will need to be
+modified
+
+#### Checking Proportional Hazards Assumption
+
+The `cox.zph` function can be used to check the proportional hazards
+assumption. As mentioned above, the function calculates p values for
+each covariate. If the p values are \< .05, the null hypothesis is
+rejected indicating the proportional hazards assumption is true. If the
+proportional hazards assumption is not true for any covariate, the model
+will be recreated without those covariates.
+
+```{r}
+cox.zph(cox1)
+```
+
+The output above shows the p values for the hypothesis test.
+
+In addition to the hypothesis test, the Shoenfeld residuals calculated
+by the `cox.zph()` function are plotted against time below.
+
+```{r}
+# Function to put ph_plots on one visual
+create_ph_plots <- function(coxmodel){
+  # Function takes argument that is a cox model from coxph()
+
+  phresults <- cox.zph(coxmodel)
+
+  # Define dimensions of visual
+  par(mfrow = c(3,5),
+      mar = c(2,2,4,1))
+
+  # Loop through each covariate creating a plot and adding it to the phplot list
+  for (i in 1:(length(phresults$table[,1])-1)){
+    phplots <- list()
+    phplots[i] <- plot(phresults[i], # Create plot for each iteration
+                       main = rownames(phresults$table)[i]) # Title of each plot is the rowname from coxmodel$table
+
+  }
+  mtext("Schoenfeld Residuals vs. Time for Covariates", side = 3, line = -1.5, outer = TRUE, cex = 1) # Title for Visual
+  par(mfrow = c(1,1)) # Reset plotting area
+}
+
+create_ph_plots(cox1)
+```
+
+The proportional hazards test shows several covariates do not meet the
+proportional hazards assumption.
+
+The model is recreated below with another proportional hazards test.
+
+```{r}
+
+cox1.1 <- coxph(data = train, 
+             Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)
+cox.zph(cox1.1)
+
+```
+
+The hypothesis test shows the covariates sm1, sm2, sm6, sm7, sm11, sm12,
+sm20, and sm21 do not meet the proportional hazards assumption. The
+model is created again below with a second hypothesis test showing all
+of the covariates pass the hypothesis test.
+
+#### Checking Linearity Assumption
+
+Residuals can first be found by using the `resid(type = "martingale")`
+function and passing martingale in the type argument creating a vector
+of residuals that can be plotted against the range of values for each
+covariate. Example for one predictor
+
+```{r}
+martingale <- resid(cox1.1, type = "martingale")
 
@@
 
+The survival curve will be plotted by using the survival package's
+`survfit()` function. The object created will be plotted using `plot()`
+
 ```{r}
-survfit <- survfit(cox)
+survfit <- survfit(cox1.1)
 

### 2025-03-24 index.qmd
Updates 3/24 (JR)

@@
 library(knitr)
+library(ezcox)
 library(sjPlot)
@@
     when compared to $Z = z$.
--   if HR $< 1$ implies a negatie association such that
+-   if HR $< 1$ implies a negative association such that
     $100\% \times (HR -1)$ is the \\% of decreased risk for $Z = z+1$
@@
 $$
+\begin{align*}
  H(x) = \int_0^x h(u) du 
+\end{align*}
 $$ {#eq-cumulativehr}
@@
 
-example_coxmodel <- coxph(example_surv_object ~ meal.cal + age, data = lung)
+example_coxmodel <- coxph(example_surv_object ~ meal.cal, data = lung)
 ```
@@
 visually using the Martingale residuals for each covariate
-[@nahhas2025]. Martingale residuals are a type of residauls used in
+[@nahhas2025]. Martingale residuals are a type of residuals used in
 survival analysis that show the discrepancy between the observed and the
 expected number of events. The equation is $$
+\begin{align*}
 \text{Martingale Residuals} = \text{Observed Events} - \text{Expected Events}
-$$ The resulting residuals are examined visually and if they are linear
-and appear to have a slope of zero, the linearity assumption is not
-violated [@Amini2015]. If a model is being generated with multiple
-covariates, multiple Martingale residual plots will need to be generated
-in order to verify this assumption for each covariate.
+\end{align*}
+$$
+
+The resulting residuals are examined visually and if they are linear and
+appear to have a slope of zero, the linearity assumption is not violated
+[@Amini2015]. If a model is being generated with multiple covariates,
+multiple Martingale residual plots will need to be generated in order to
+verify this assumption for each covariate.
 
@@
 
-The major assumption for CPH is the proportional hazards assumption
+The major assumption for CPH is the proportional hazards (PH) assumption
 which states that the ratio of hazard rates for any two subjects must be
@@
 constant over time [@Bustan2018]. To test this assumption, the
-hypothesis test below is used along with Schoenfeld partial residuals.
-
-$$
+hypothesis test below is used: $$
 \begin{align*}
@@
 
-The scaled Schoenfeld partial residuals are calcuated and returns the
-$\chi^2$ value, the degrees of freedom, and p-values for each individual
-covariate along with a global value. Using an $\alpha = 0.05$ and the
-global p-value, the determination of whether to reject or fail to reject
-$H_0$ can be made. In addition to the hypothesis test, these Shoenfeld
-residuals can be plotted against time for each covariate to visualize
-the assumption. When plotting the residuals against time, if the
-assumption is not broken it would be expected that the points show
-roughly a straight horizontal line to indicate constant hazard over time
-[@nahhas2025].
+To generate the p-value for the hypothesis test, the scaled Schoenfeld
+partial residuals are calculated for each covariate. The Schoenfeld
+partial residuals are the difference between the value of the covariate
+and the expected value of the covariate at the time of failure
+[@klein2005]. To visually confirm the PH assumption, the plot of the
+scaled Schoenfeld partial residuals for each covariate should be a
+horizontal line [@harrison2021]. The global p–value will provide the
+p-value for all covariates in the model which can be compared to an
+$\alpha = 0.05$ to determine if the PH assumption has been met for all
+covariates [@nahhas2025].
 
-Individual plots for each covariate are generated which provide a visual
-interpretation of the p-values generated from `cox.zph()`. If there is
-no clear slope in the residuals on the plot, it signifies further
-confirmation the proportional hazards assumption is met.
+Violations of the PH assumption can be dealt with in several ways. If
+the covariate is categorical, one way is to stratify the covariate which
+will generate a separate baseline hazard function for each level of the
+covariate [@harrison2021]. This removes the PH assumption for that
+covariate, though it will still be assumed for all other covariates in
+each level of the stratification [@nahhas2025]. If the covariate is
+continuous, one way is to add in a function of time based on the shape
+of the hazard ratio plot. This replaces the model's time coefficient
+with a function of time as it assumes the covariate varies either
+linearly or non-linearly with time [@nahhas2025].
 
-Below is an example plot demonstrating Shoenfeld residuals plotted
-against time to test the proportional hazards assumption. While there is
-some curvature in the line, the residuals are roughly horizontal
-indicating proportional hazard over time for this covariate. In a model
-with multiple covariates, each one would need to be visualized to ensure
-it does not break this assumption and can be used in the model.
+Below is an example plot demonstrating scaled Shoenfeld partial
+residuals plotted against time to test the proportional hazards
+assumption. While there is some curvature in the line, the residuals are
+roughly horizontal indicating proportional hazard over time for this
+covariate.
 
-```{example_phtest <- cox.zph(example_coxmodel)}
+```{r, echo = FALSE}
+example_phtest <- cox.zph(example_coxmodel)
 
@@
 
-### Evaluation
+Below is an example of the global value that can be generated for the
+whole model. In comparing it with an $\alpha = 0.05$, the covariate does
+not meet the proportional hazards assumption as $p = 0.128 < \alpha$ .
+However, individually, meal.cal does meet the assumption as
+$p = 0.044 < \alpha$ . As such, age would need to be adapted based on
+one of the above methods in order to include it in the model.
 
-**Concordance Index**
+```{r, echo=FALSE}
+multi <- coxph(example_surv_object ~ meal.cal + age, data = lung)
+ph_tibble <- cox.zph(multi) %>%
+  pluck("table") %>%
+  as.data.frame() %>%
+  tibble::rownames_to_column("variable") %>%
+  as_tibble()
+kable(ph_tibble, format = "markdown")
+```
 
-To evaluate the model, the Cox regression concordance index can be used.
-Two observed pairs are said to be concordant if an event that is
-expected to occur before another does occur first. For example, if two
-engines are selected from the data and the model predicts one of them to
-be at a higher risk of failure than the other and the one with the
-higher risk fails first, then the pairs are concordant. If the
-alternative happens where the lower risk engine experiences failure
-first, then the pairs are discordant.
+### Model Visualization
 
-The concordance index is calculated as follows:
+Two common model visualizations used for CPH is the forest plot and the
+Kaplan-Meier survival curve. The Kaplan-Meier curve assists in
+visualizing the initial survival model for the data while the forest
+plot assists in visualizing the hazard ratio after generating a model.
 
-$C = \frac{ConcordantPairs}{ConcordantPairs + DiscordantPairs}$
+**Kaplan Meier Survival Curve**
 
-The concordance index will produce a value between 0 and 1 with 1
-meaning all pairs are correctly ordered and 0 being the alternative with
-no pairs correctly ordered. The higher the concordance index for the
-created model, the more accurately it can model survival for the
-data.[@Therneau2017]
+The Kaplan-Meier (KM) estimator is also known as the Product-Limit
+estimator and was proposed by Kaplan and Meier in 1958. The estimator
+can be defined as
 
-### Model Visualization and Prediction
+$$
+\hat{S}(t) = \begin{cases}
+1 \qquad \qquad \qquad \: \text{if } \; t< t_1,\\
+\prod_{t_\leq t}[1 - \frac{d_j}{Y_i}], \quad\text{if} \; t_1 \leq t
+\end{cases}
+$$ {#eq-km} where,
 
-There are multiple different methods to visualize data when creating a
-Cox proportional hazards model. Survival curves are a popular way to
-visualize survival probabilities over time, but there are other
-visualizations that are helpful when interpreting the results of the
-model.
+-   $d_i$ is the number of events at time $t_i$
+-   $Y_i$ is the number of individuals who are risk at time $t_i$
+    [@klein2005]
 
-Cox proportional hazards models can produce an assortment of useful
-metrics to interpret the model as described by the equations above. The
-model created can be used to estimate risk of the event occurring, rate
-of occurrence, survival probability, and time to event.
+KM is a non-parametric estimate which means it makes no assumption of
+the shape of the base survival function [@nahhas2025] and is considered
+to be well defined up until the largest observed study time $t_{max}$
+[@klein2005]. The survival curve provides a visualization of the KM
+estimator and are important when modeling survival data because it
+demonstrates the time where the event being modeled is expected to
+occur.
 
-#### Forest Plot
+Below is an example survival curve showing survival probabilities over
+time. The line represents the median survival probability with the red
+area around it showing the 95% confidence interval for the values. The
+curve can be interpreted as showing the probability of event would be
+expected to happen around a time of 300 where the survival probability
+is below 50% [@Kuitunen2021].
+
+```{r, echo=FALSE}
+example_curve <- survfit(example_coxmodel)
+
+ggsurvplot(example_curve,
+           data = lung,
+           title = "Example Survival Curve",
+           xlab = "Time",
+           ylab = "Survival Probability",
+           legend = "none")
+```
+
+**Forest Plots**
 
@@
 ```{r, echo=FALSE,warning=FALSE}
-plot_model(example_coxmodel,
+forest <- coxph(example_surv_object ~ age + sex, data = lung)
+
+plot_model(forest,
            dot.size = 1,
@@
   xlab(label = "Estimated Effect on Hazard Ratio")
-```
-
-#### Kaplan-Meier Survival Curve
-
-After generating the CPH model, a Kaplan-Meier survival curve can be
-generated to view survival probabilities against time. These plots are
-important when modeling survival data because they clearly demonstrate
-the time where the event being modeled for is expected to occur.
-
-Below is an example survival curve showing survival probabilities over
-time. The line represents the median survival probability with the red
-area around it showing the 95% confidence interval for the values. The
-curve can be interpreted as showing the probability of event would be
-expected to happen around a time of 300 where the survival probability
-is below 50%.[@Kuitunen2021]
-
-```{r, echo=FALSE}
-example_curve <- survfit(example_coxmodel)
-
-ggsurvplot(example_curve,
-           data = lung,
-           title = "Example Survival Curve",
-           xlab = "Time",
-           ylab = "Survival Probability",
-           legend = "none")
+show_forest(lung, covariates = c('age'))
 
@@
 
-#### Predicting Time Until Event
+### Evaluation
 
-Time until event can be calculated by modeling survival probability for
-future time values. When survival probability drops below 50%, it will
-be assumed the event occurred. Predictions for time until event will be
-made on the testing data for each individual's maximum time value. A
-column will be added to the table showing the expected time until event.
+Evaluation of the accuracy of the CPH model is done using the
+concordance index which is used to measure the amount of agreement
+between two variables. Specifically, the value looks at concordant pairs
+and disconcordant pairs. A concordant pair is when either $x_i < x_j$
+and $y_i < y_j$ or $x_i > x_j$ and $y_i > y_j$. A discordant pair is
+when either $x_i$ or $y_i$ are not in the same place such as $x_i < x_j$
+but $y_j < y_i$ [@Therneau2017]. In terms of survival, this can be
+thought of as examining when the failure event occurs. If two subjects,
+A and B, are compared with subject A having a higher risk, the pair
+would be considered concordant if subject A does have the failure event
+occur before subject B.
+
+The equation for concordance with CPH is
+
+$$
+C =\frac{c + \frac{t_x}{2}}{c + d + t_x}
+$$ {#eq-concordance}
+
+where,
+
+-   $c$ is the count of pairs that are concordant
+-   $d$ is the count of pairs that are discordant
+-   $t_x$ are the pairs tied to the predictor $x$ [@Therneau2017]
+
+The concordance index will produce a value between 0 and 1. A value of 1
+means all pairs are correctly ordered and a value 0 being the
+alternative with no pairs correctly ordered. Any values between 0 and 1
+indicate how accurate the CPH model is with the data [@Therneau2017].
+
+### Prediction
+
+CPH can be used to predict the survival probability at a specific time
+$S(t|X=x)$, and the hazard ratio for an individual when compared to a
+reference individual $\frac{h(t|X=x)}{h(t|X=x_{ref}}$ [@nahhas2025]. If
+the resulting probability is greater than or equal to 50%, it is assumed
+the event has not occurred. If the resulting probability is less than
+50%, it is assumed the event has occurred. It is also possible to
+generate a plot of the estimated survival curves for all values of $t$
+that are desired.
 

### 2025-03-26 index.qmd
3/25 update

@@
 
+The forest plot for model 1 is shown below. Values closer to 1 indicate the hazard ratio for that covariate have less effect on the calculated hazard ratio.
+
+```{r}
+plot_model(cox1,
+           dot.size = 2,
+           line.size = 1,
+           colors = "red")+
+  theme_bw()+
+  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
+  ylab(label = "Covariates")+
+  xlab(label = "Estimated Effect on Hazard Ratio")
+```
+
 ### Checking Assumptions
@@
 # Function to put ph_plots on one visual
-create_ph_plots <- function(coxmodel){
+create_ph_plots <- function(coxmodel, row, col){
   # Function takes argument that is a cox model from coxph()
@@
   # Define dimensions of visual
-  par(mfrow = c(3,5),
+  par(mfrow = c(row, col),
       mar = c(2,2,4,1))
@@
 
-create_ph_plots(cox1)
+create_ph_plots(cox1, 3, 5)
 ```
@@
 The model is recreated below with another proportional hazards test.
+To improve the model and ensure the proportional hazards assumption is met, several covariates that do not meet the assumption are removed.
 
 ```{r}
+cox1.1 <- coxph(data = train,Surv(time, status) ~ sm2 + sm5 + sm7 + sm11 + sm12 + sm20 + sm21)
 
-cox1.1 <- coxph(data = train, 
-             Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)
 cox.zph(cox1.1)
@@
 ```
+The new model has seven covariates with only one not meeting the proportional hazards assumption.
+Although sm5 does not have a p value > .05, the assumption overall holds for the new model and sm5 was not removed from the model because of its large impact on the hazard ratio shown in the forest plot.
+The new model's full summary is below.
 
-The hypothesis test shows the covariates sm1, sm2, sm6, sm7, sm11, sm12,
-sm20, and sm21 do not meet the proportional hazards assumption. The
-model is created again below with a second hypothesis test showing all
-of the covariates pass the hypothesis test.
+```{r}
+summary(cox1.1)
+
+```
 
@@
 ```{r}
+
+par(mfrow = c(3,3),
+      mar = c(2,2,4,1))
+
 martingale <- resid(cox1.1, type = "martingale")
 
-plot(train$os1, martingale, main = "Martingale Residuals: os1")
+plot(train$sm2, martingale, main = "Martingale Residuals: sm2")
+abline(h = 0, col = "red")
+
+plot(train$sm5, martingale, main = "Martingale Residuals: sm5")
+abline(h = 0, col = "red")
+
+plot(train$sm7, martingale, main = "Martingale Residuals: sm7")
+abline(h = 0, col = "red")
+
+plot(train$sm11, martingale, main = "Martingale Residuals: sm11")
+abline(h = 0, col = "red")
+
+plot(train$sm12, martingale, main = "Martingale Residuals: sm12")
+abline(h = 0, col = "red")
+
+plot(train$sm20, martingale, main = "Martingale Residuals: sm20")
+abline(h = 0, col = "red")
+
+plot(train$sm21, martingale, main = "Martingale Residuals: sm21")
 abline(h = 0, col = "red")
@@
 ```
+The plots of the martingale residuals vs the range of values for each covariate is shown above.
+For each covariate, there is no clear slope indicating the linearity assumption is true for the covariates in the model.
 
@@
 source clarifies all engines in the training data have experienced the
-desired event of failure.
+desired event of failure at the end of each time series.
 
@@
 ### Model Visualization
-
-Forest Plot
-
-A forest plot will be created using the `plot_model()` function from the
-sjPlot package. The CPH model can be passed as the sole argument to
-visualize the hazard ratios.
-
-```{r}
-plot_model(cox1.1,dot.size = 4,line.size = 2,colors = "red")+
-  theme_bw()+
-  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
-  ylab(label = "Covariates")+
-  xlab(label = "Estimated Effect on Hazard Ratio")
-```
 

### 2025-03-30 index.qmd
3/30 updates

@@
 
-summarymetrics <- summary(Minimum = min(time),
-                          Median = median(time),
-                          StandardDeviation = sd(time),
-                          Maximum = max(time), 
-                          distributions$time)
+summarymetrics <- data.frame(c("Minimum", "Median", "Mean", "Standard Deviation", "Maximum"),
+                             c(min(distributions$time), 
+                               median(distributions$time),
+                               mean(distributions$time),
+                               sd(distributions$time),
+                               max(distributions$time)))
 
-kable(tibble(summarymetrics), format = "markdown")
+colnames(summarymetrics) <- c("Metric", "Value")
+summarymetrics$Value <- round(summarymetrics$Value, digits = 2)
+kable(summarymetrics)
 
@@
 ```{r}
-cox <- coxph(Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5,
-             data = train,
-             id = id)
-
-cox1 <- coxph(data = train, 
+cox1 <- coxph(data = train, id = id,
              Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)
 
-cox2 <- coxph(data = train, 
+cox2 <- coxph(data = train, id = id,
              Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5)
 
-cox3 <- coxph(data = train, 
+cox3 <- coxph(data = train, id = id,
              Surv(time, status) ~ os1 + os3 + sm1 + sm5 + sm6 + sm7 + sm13 + sm14 + sm19 + sm21)
 
-cox4 <- coxph(data = train, 
+cox4 <- coxph(data = train, id = id,
              Surv(time, status) ~ sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)
 
-cox5 <- coxph(data = train, 
+cox5 <- coxph(data = train, id = id,
              Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5  + sm7 + sm11 + sm12 + sm13 + sm19 + sm20 )
 
-cox6 <- coxph(data = train, 
+cox6 <- coxph(data = train, id = id,
              Surv(time, status) ~ os1  + sm1 + sm2 + sm5  + sm7 + sm11 + sm12 + sm13 + sm19  + sm21)
 
-cox7 <- coxph(data = train, 
+cox7 <- coxph(data = train, id = id,
              Surv(time, status) ~ os1  + sm13 + sm14 + sm19 + sm20)
 
-cox8 <- coxph(data = train, 
+cox8 <- coxph(data = train, id = id,
              Surv(time, status) ~ os1 + os3 + sm21)
 
-cox9 <- coxph(data = train, 
+cox9 <- coxph(data = train, id = id,
              Surv(time, status) ~ os1 + os3 + sm2 + sm5 + sm6 + sm12 + sm13  + sm19 + sm20 + sm21)
 
-cox10 <- coxph(data = train, 
+cox10 <- coxph(data = train, id = id,
              Surv(time, status) ~  os3 + sm1 + sm2 + sm5 + sm6  + sm11 + sm12 + sm13  + sm19 + sm20 + sm21)
@@
 
-### Model Evaluation
-
 ### Model Visualization
@@
 
-ggsurvplot(survfit, data = test)
 ```
@@
 
-Create df with max time for each individual to use in predictions
+The model created can be used to calculate several relevant metrics when predicting remaining time for the engines until expected failure.
+Predictions will be made on the largest value for time for each individual engine.
+
+#### Model Coefficients
+
+The CPH model created produced coefficients for each covariate indicating the quantitative effect on the resulting hazard rate. The coefficients are shown below.
 
 ```{r}
-#predictiondata <- test %>%
-  #group_by(id) %>%
-  #slice_max(time) %>%
-  #select(id, time, os1, os3, sm1, sm2, sm5, status)
-
-#coef <- data.frame(Variable = c("Z1", "Z2", "Z3", "Z4", "Z5"),
-           #CoefficientValue = cox1.1$coefficients)
-#kable(coef)
+coef <- data.frame(CoefficientValue = cox1.1$coefficients)
+kable(coef)
 ```
 
-Hazard Rate Calculation:
+#### Hazard Rate Calculation
+
+The coefficients above can be used to calculate the hazard rate at time $t$. The hazard rate will be calculated for each engine at the engines max time value.
+
+Hazard Rate:
 $h(t|Z) = h_0(t) * e^{B_1Z_1 + B_2Z_2 + B_3Z_3+B_4Z_4+B_5Z_5}$
 
+The engines with the higher hazard rates are at a higher risk of experiencing failure.
+
 ```{r}
-test$hazardrates <- predict(cox1.1,
+test$HazardRate <- predict(cox1.1,
         newdata = test,
@@
 
+predictiondata <- test %>%
+  group_by(id) %>%
+  slice_max(time) %>%
+  mutate(CumulativeHazard = cumsum(HazardRate)) %>%
+  select(id, time, sm2, sm5, sm7, sm11, sm12, sm20, sm21, status, HazardRate)
+
+kable(head(predictiondata %>% select(id, time, sm2, sm5, sm7, sm11, sm12, sm20, sm21, HazardRate)))
+
 ```
+
+#### Cumulative Hazard Calculation
+
+The cumulative hazard represents the total hazard accumulated for the individual over time and like the hazard rate, is another measure that can be used to estimate the risk for the individual at a point in time.
+The cumulative hazard for each individual is shown below.
+Higher values represent a higher chance of failure.
 
@@
 ```{r}
-test <- test %>%
-  arrange(id, time) %>%
-  group_by(id) %>%
-  mutate(cumulative_hazard = cumsum(hazardrates))
-```
-
-Predicting survival probability
-
-```{r}
-test$survprob <- predict(cox1.1, type = "survival", newdata = test)
+kable(head(predictiondata))
 
@@
 
+#### Survival probability
+
+While cumulative hazard can be used by itself to give an idea of an individual's risk, it can also be used to calculate a probability of survival.
+
 Probability Calculation: $S(t) = e^{-H(t)}$ $H(t)$ is the cumulative
-hazard up to time $t$
+hazard up to time $t$.
 
-Tim until event can be modeled in R by passing the `type = "lp"`
+```{r}
+predictiondata$SurvivalProbability <- predict(cox1.1, type = "survival", newdata = predictiondata)
+kable(head(predictiondata))
+
+```
+
+#### Predicting Time Until Failure
+
+Time until event can be modeled in R by passing the `type = "lp"`
 argument in the `stats::predict()` function to get a value for linear
@@
 ```
-
-A study was conducted to determine how...Modeling and Results
-
--   Explain your data preprocessing and cleaning steps. transformation
-    applied to data to add event
-
--   Present your key findings in a clear and concise manner.
-
--   Use visuals to support your claims.
-
--   **Tell a story about what the data reveals.**
-
-### Results
 

### 2025-04-01 index.qmd
3/30 update
Update 3/31

@@
 library(knitr)
-library(ezcox)
 library(sjPlot)
+library(broom)
 
@@
   xlab(label = "Estimated Effect on Hazard Ratio")
-show_forest(lung, covariates = c('age'))
 
@@
 
-### Prediction
+### Survival Prediction
 
@@
 
-Though CPH is typically used with health data, we wanted to determine
-how CPH can be used in other fields. As such, we will study the
-remaining useful life (RUL) of a NASA aircraft engine using data
-provided by NASA for the study on propagation modeling [@saxena2008]. In
-particular, we will be examining their data from what they have deemed
-engine two. From the literature review, we have noted that we will need
-to be concerned with the constant hazards ratio within our data and
-determine how we will accommodate censored data when creating our model.
+Though CPH is typically used with health data, it does have applications
+in other fields. This paper will study the remaining useful life (RUL)
+of a NASA aircraft engine using data provided by NASA for the study on
+propagation modeling [@saxena2008]. In particular, this paper will be
+examining the data from the engine make two which has a training dataset
+and a testing dataset.
 
-#### Data Structure
+**Data Structure**
 
-Each engine has an unknown amount of wear, manufacturing variation, and
-sensor noise all of which impact the survival time. In the training
-dataset, each engine is operating normally at the beginning of the time
-series, develops a fault, and experiences failure at the end of the time
-series. The conditions the engines are subjected to are represented by
-three operational setting fields and 21 sensor measurement fields. In
-the testing data, the engines have not experienced failure
-[@saxena2008].
+Each engine in the NASA data has an unknown amount of wear,
+manufacturing variation, and sensor noise which will impact the survival
+time. In the training dataset, each engine is operating normally at the
+beginning of the time series, will develop a fault, and experience
+failure at some point. All engines will have failed in the training
+dataset. The conditions the engines are subjected to are represented by
+three operational setting fields and twenty-one sensor measurement
+fields. In the testing dataset, none of the engines have experienced
+failure [@saxena2008].
 
-For both test and train datasets, we will need to add a column
-indicating status. We will use 0 for still working and 1 for the last
-cycle when the engine fails. For the test data, we know that none of the
-engines have failed so all values in the test column will be 0 for still
-working. For the training data, we know that the data includes all
-cycles up to the last cycle when the machine will fail. As such, we will
-have 0s in all cycles except the last cycle of the machine.
+A column indicating status is needed for both the training and testing
+datasets. The value 0 will be used to indicate if the machine is still
+working and the value 1 will be used for the last cycle before the
+engine fails. Both values will appear in the training dataset as all
+engines in that dataset will eventually fail. The testing dataset will
+only have the value of 0 as none of the machines have failed in that
+dataset.
 
@@
 after the transformations describd above.
-
-**Testing Data**
 
@@
 
-kable(head(test, n = 5), format = "markdown")
-```
+kable(head(test, n = 5), format = "markdown", caption = "Testing Data")
 
-**Training Data**
-
-```{r}
 trainurl <- 'https://raw.githubusercontent.com/JR-87/capstone4ds_reed/refs/heads/main/train_FD002.txt'
@@
 
-kable(head(train, n = 5), format = "markdown")
+kable(head(train, n = 5), format = "markdown", caption = 'Training Data')
 ```
 
-#### Data Visualization
+**Data Visualization**
 
-The visuals below describe the distribution of the 260 engines in the
-training data. The histogram shows the survival times are normally
-distributed with a median value of 199 iterations and the box plot shows
-there are few outliers out of the 260 engines. The engines have a median
-survival time of 199 iterations with a standard deviation of 46.8. The
-longest surviving engine lasted 378 iterations, with the shortest
-survival time for all engines being 128.
+There are 260 engines in the training data. The histogram shows that the
+survival times of the engines are normally distributed and the box plot
+shows there six outliers in the dataset. The engines have a median
+survival time of 199 cycles with a standard deviation of 46.8. The
+maximum number of cycles an engine lasted was 378 while the minimum
+number of cycles an engined lasted was 128.
 
-```{r}
+```{r, echo = FALSE}
 distributions <- train %>% 
@@
 summarymetrics$Value <- round(summarymetrics$Value, digits = 2)
-kable(summarymetrics)
+kable(summarymetrics, format = "markdown", padding = 0)
 
@@
 
-The Kaplan-Meier survival curve is shown below.
+To determine the intitial survival function, the Kaplan-Meier survival
+curve was generated and is shown below. The graph verifies that the
+median survival time is roughly 200 cycles and shows the maximum cycle.
+After that cycle has been reached, the survival probability plummets to
+0%.
 
@@
 
-#### Choosing Covariates
+**Covariate Selection**
 
-Below is a univariate test to determine which covariates to use in the
-model.
+Based on the information provided by NASA on their data, there are
+various sensors and operational setting fields that do not impact the
+survival rate of an engine. To determine which covariates will impact
+the survival probability, the below univariate test was run. The
+covariates in which the p-value is $< 0.05$ will be considered for the
+CPH models while the other covariates will be excluded.
 
@@
 
-#### Creating Model
+Based on the above test, the covariates that will be considered for the
+CPH models are os1, os3, sm1, sm2, sm5, sm6, sm7, sm11, sm12, sm13,
+sm14, sm19, sm20, and sm21 which is fourteen out of twenty-four possible
+covariates.
 
-To generate the CPH model with properly formatted survival data, a
-survival object must first be created using the R package `survival`.
-The survival object is created by running `Surv(time, event)` where
-`time` is the time field and `event` is the event field.
+**Generating Models**
 
-To create the CPH model, the survival object is then applied to the R
-function `coxph(survival object ~ x, data)` where `x` represents one or
-multiple covariate(s) and `data` is the survival data that is being
-modeled.[@R-base].
+Generation of all CPH models will be done using R, specifically the R
+package `survival`. To generate the CPH model with properly formatted
+survival data, a survival object is created by running
+`Surv(time, event)` where `time` is the time field and `event` is the
+event field. To generate the actual model, the survival object is then
+applied to the function `coxph(survival object ~ x, data)` where `x`
+represents one or multiple covariate(s) and `data` is the survival data
+that is being modeled.[@R-base].
 
-During the analysis, multiple models will be created with a different
-combination of covariates.
+The table below provides the model number, the covariates used, and the
+concordance index. The concordance index generates a value to determine
+which model is the most accurate with the provided covariates. After
+determining the most accurate model, the CPH assumptions will be tested.
 
@@
 
-The resulting models and their corresponding concordance index values
-are below.
+```{r}
+concordance_results <- data.frame(c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5", 
+                                    "Model 6", "Model 7", "Model 8", "Model 9", "Model 10"), 
+                   c("os1, os3, sm1, sm2, sm5, sm6, sm7, sm11, sm12, sm13, sm14, sm19, sm20, sm21", 
+                     "os1, os3, sm1, sm2, sm5", 
+                     "os1, os3, sm1, sm5, sm6, sm7, sm13, sm14, sm19, sm21", 
+                     "sm5, sm6, sm7, sm11, sm12, sm13, sm14, sm19, sm20, sm21", 
+                     "os1, os3, sm1, sm2, sm5, sm7, sm11, sm12, sm13, sm19, sm20", 
+                     "os1, sm1, sm2, sm5, sm7, sm11, sm12, sm13, sm19, sm21",
+                     "os1, sm13, sm14, sm19, sm20", 
+                     "os1, os3, sm21",
+                     "os1, os3, sm2, sm5, sm6, sm12, sm13, sm19, sm20, sm21",
+                     "os3, sm1, sm2, sm5, sm6, sm11, sm12, sm13, sm19, sm20, sm21"),
+                   c(cox1$concordance[6], cox2$concordance[6], cox3$concordance[6], cox4$concordance[6], cox5$concordance[6], cox6$concordance[6], cox7$concordance[6], cox8$concordance[6], cox9$concordance[6], cox10$concordance[6]))
 
-```{r}
-concordance_results <- data.frame(row.names = c(
-  "Model 1",
-  "Model 2",
-  "Model 3",
-  "Model 4",
-  "Model 5",
-  "Model 6",
-  "Model 7",
-  "Model 8",
-  "Model 9",
-  "Model 10"),
-  Concordance = c(
-  cox1$concordance[6],
-  cox2$concordance[6],
-  cox3$concordance[6],
-  cox4$concordance[6],
-  cox5$concordance[6],
-  cox6$concordance[6],
-  cox7$concordance[6],
-  cox8$concordance[6],
-  cox9$concordance[6],
-  cox10$concordance[6]))
+colnames(concordance_results) <- c("Model", "Covariates", "Concordance Index")
 
-kable(concordance_results)
+kable(concordance_results, format = "markdown")
 
@@
 
-The analysis will proceed using model 1 because of the high concordance
-index. The full summary of model 1 is below.
+Model 1 had the highest concordance index and used all of the covariates
+that were determined to be significant to the survival of the engine.
+The proceeding analysis will be done using model 1. The below tables
+provide the summary values from the model.
 
-```{r}
-summary(cox1)
-
+```{r, echo = FALSE}
+kable(tidy(cox1), format = "markdown")
+g <- glance(cox1)
+g <- g %>% select(n, nevent, concordance, std.error.concordance, logLik, statistic.log, p.value.log, statistic.wald, p.value.wald, statistic.sc, p.value.sc)
+kable(g, format = "markdown")
 ```
 
-The forest plot for model 1 is shown below. Values closer to 1 indicate the hazard ratio for that covariate have less effect on the calculated hazard ratio.
+The forest plot for model 1 is shown below. The covariates that are
+closest to the value of one have a hazard rate that has less effect on
+the calculated hazard ratio for the entire model.
 
-```{r}
+```{r, echo = FALSE}
 plot_model(cox1,
@@
 
-Despite the high concordance value for model 1, the assumptions will
-need to be evaluated before using the model to make predictions. If any
-of the assumptions for the model are not true, the model will need to be
-modified
+After determining the model to use with the covariates, the CPH
+assumptions need to be tested to confirm the model is a valid CPH model.
+The model may have a high concordance index, but it may not be valid for
+CPH to use to predict the survival probability. If any of the
+assumptions for the model are not true, the model will need to be
+modified.
 
-#### Checking Proportional Hazards Assumption
+**Independence Assumption**
+
+**Non-informative Censoring Assumption**
+
+**Linearity Assumption**
+
+**Proportional Hazards Assumption**
 
@@
 
-The model is recreated below with another proportional hazards test.
-To improve the model and ensure the proportional hazards assumption is met, several covariates that do not meet the assumption are removed.
+The model is recreated below with another proportional hazards test. To
+improve the model and ensure the proportional hazards assumption is met,
+several covariates that do not meet the assumption are removed.
 
@@
 ```
-The new model has seven covariates with only one not meeting the proportional hazards assumption.
-Although sm5 does not have a p value > .05, the assumption overall holds for the new model and sm5 was not removed from the model because of its large impact on the hazard ratio shown in the forest plot.
-The new model's full summary is below.
+
+The new model has seven covariates with only one not meeting the
+proportional hazards assumption. Although sm5 does not have a p value \>
+.05, the assumption overall holds for the new model and sm5 was not
+removed from the model because of its large impact on the hazard ratio
+shown in the forest plot. The new model's full summary is below.
 
@@
 ```
-The plots of the martingale residuals vs the range of values for each covariate is shown above.
-For each covariate, there is no clear slope indicating the linearity assumption is true for the covariates in the model.
+
+The plots of the martingale residuals vs the range of values for each
+covariate is shown above. For each covariate, there is no clear slope
+indicating the linearity assumption is true for the covariates in the
+model.
 
@@
 desired event of failure at the end of each time series.
+
+### Adjusting the Model
 
@@
 
+```{r}
+ggsurvplot(survfit, data = test, fun = "pct", surv.median.line = "none")
+```
+
+```{r}
+plot(survfit, mark.time=FALSE, xscale=365.24,
+	xlab='Years', ylab='Survival')
+lines(survfit[1], lwd=2, xscale=365.24)
+```
+
 ### Model Applications and Predictions
 
-The model created can be used to calculate several relevant metrics when predicting remaining time for the engines until expected failure.
-Predictions will be made on the largest value for time for each individual engine.
+The model created can be used to calculate several relevant metrics when
+predicting remaining time for the engines until expected failure.
+Predictions will be made on the largest value for time for each
+individual engine.
 
@@
 
-The CPH model created produced coefficients for each covariate indicating the quantitative effect on the resulting hazard rate. The coefficients are shown below.
+The CPH model created produced coefficients for each covariate
+indicating the quantitative effect on the resulting hazard rate. The
+coefficients are shown below.
 
@@
 
-The coefficients above can be used to calculate the hazard rate at time $t$. The hazard rate will be calculated for each engine at the engines max time value.
+The coefficients above can be used to calculate the hazard rate at time
+$t$. The hazard rate will be calculated for each engine at the engines
+max time value.
 
@@
 
-The engines with the higher hazard rates are at a higher risk of experiencing failure.
+The engines with the higher hazard rates are at a higher risk of
+experiencing failure.
 
@@
 
-The cumulative hazard represents the total hazard accumulated for the individual over time and like the hazard rate, is another measure that can be used to estimate the risk for the individual at a point in time.
-The cumulative hazard for each individual is shown below.
-Higher values represent a higher chance of failure.
+The cumulative hazard represents the total hazard accumulated for the
+individual over time and like the hazard rate, is another measure that
+can be used to estimate the risk for the individual at a point in time.
+The cumulative hazard for each individual is shown below. Higher values
+represent a higher chance of failure.
 
@@
 
-While cumulative hazard can be used by itself to give an idea of an individual's risk, it can also be used to calculate a probability of survival.
+While cumulative hazard can be used by itself to give an idea of an
+individual's risk, it can also be used to calculate a probability of
+survival.
 

### 2025-04-01 index.qmd
Update 4/1

@@
 \begin{align*}
-H_0&: \text{the predictor's coefficient does vary with time} \\
-H_1&: \text{the predictor's coefficient does not vary with time}
+H_0&: \text{the log hazard ratio for each covariate remains constant over time.} \\
+H_1&: \text{the log hazard ratio for each covariate does not remain constant over time}
 \end{align*}
@@
 and the expected value of the covariate at the time of failure
-[@klein2005]. To visually confirm the PH assumption, the plot of the
-scaled Schoenfeld partial residuals for each covariate should be a
-horizontal line [@harrison2021]. The global p–value will provide the
-p-value for all covariates in the model which can be compared to an
-$\alpha = 0.05$ to determine if the PH assumption has been met for all
-covariates [@nahhas2025].
+[@klein2005]. If the global p-value is $< 0.05$, than the model has at
+least one covariate that violates the PH assumption and each individual
+covariate p-value will need to be checked. If the global p-value is
+$> 0.05$, then the PH assumption holds for all covariates [@nahhas2025].
+To visually confirm the PH assumption, the plot of the scaled Schoenfeld
+partial residuals for each covariate should be a horizontal line
+[@harrison2021].
 
@@
 summarymetrics$Value <- round(summarymetrics$Value, digits = 2)
-kable(summarymetrics, format = "markdown", padding = 0)
+kable(summarymetrics, format = "markdown")
 
@@
 
-```{r, echo = FALSE}
+```{r}
 kable(tidy(cox1), format = "markdown")
@@
 
-```{r, echo = FALSE}
+```{r}
 plot_model(cox1,
@@
 
+The survival times for each individual engine in the training data are
+independent of each other. The failure of one engine does not impact the
+failure of another so this assumption is met.
+
 **Non-informative Censoring Assumption**
 
+The training dataset does not contain censored data as the documentation
+available at the dataset’s source clarifies all engines in the training
+data have experienced the desired event of failure at the end of each
+time series. As such, this assumption is met.
+
 **Linearity Assumption**
+
+To check the linearity assumption, the residuals of the model are needed
+which can be generated using the `resid` function in R and assigning the
+type to be martingale (`resid(type = "martingale")` ). This will create
+a vector of residuals that can be plotted against the range of values
+for each covariate.
+
+The plots of the Martingale residuals vs the range of values for each
+covariate is shown below. For each covariate, there is no clear slop
+indicating the linearity assumption is true for the covariates in the
+model.
+
+```{r}
+par(mfrow = c(3,3),
+      mar = c(2,2,4,1))
+
+martingale <- resid(cox1, type = "martingale")
+
+plot(train$os1, martingale, main = "Martingale Residuals: os1")
+abline(h = 0, col = "red")
+
+plot(train$os2, martingale, main = "Martingale Residuals: os2")
+abline(h = 0, col = "red")
+
+plot(train$sm1, martingale, main = "Martingale Residuals: sm1")
+abline(h = 0, col = "red")
+
+plot(train$sm2, martingale, main = "Martingale Residuals: sm2")
+abline(h = 0, col = "red")
+
+plot(train$sm5, martingale, main = "Martingale Residuals: sm5")
+abline(h = 0, col = "red")
+
+plot(train$sm6, martingale, main = "Martingale Residuals: sm6")
+abline(h = 0, col = "red")
+
+plot(train$sm7, martingale, main = "Martingale Residuals: sm7")
+abline(h = 0, col = "red")
+
+plot(train$sm11, martingale, main = "Martingale Residuals: sm11")
+abline(h = 0, col = "red")
+
+plot(train$sm12, martingale, main = "Martingale Residuals: sm12")
+abline(h = 0, col = "red")
+
+plot(train$sm13, martingale, main = "Martingale Residuals: sm13")
+abline(h = 0, col = "red")
+
+plot(train$sm14, martingale, main = "Martingale Residuals: sm14")
+abline(h = 0, col = "red")
+
+plot(train$sm20, martingale, main = "Martingale Residuals: sm20")
+abline(h = 0, col = "red")
+
+plot(train$sm21, martingale, main = "Martingale Residuals: sm21")
+abline(h = 0, col = "red")
+```
 
@@
 assumption. As mentioned above, the function calculates p values for
-each covariate. If the p values are \< .05, the null hypothesis is
-rejected indicating the proportional hazards assumption is true. If the
-proportional hazards assumption is not true for any covariate, the model
-will be recreated without those covariates.
-
-```{r}
-cox.zph(cox1)
-```
-
-The output above shows the p values for the hypothesis test.
-
-In addition to the hypothesis test, the Shoenfeld residuals calculated
-by the `cox.zph()` function are plotted against time below.
+each covariate. The global p-value will first be compared to
+$\alpha = 0.05$. If that p-value is $<\alpha$, then at least one
+covariate violates the PH assumption and the model will need to be
+adapted.
+
+The output below shows the p-values for the hypothesis test and
+Schoenfeld residuals. The global p-value is $< \alpha$ so at least one
+covariate violates the PH assumption. Examining each individual
+covariate through both p-values and the plot of the Schoenfeld residuals
+shows that the covariates os1, os3, sm1, sm5, sm13, and sm14 all violate
+the PH assumption.
+
+```{r}
+cph_ph <- cox.zph(cox1) %>%
+  pluck("table") %>%
+  as.data.frame() %>%
+  tibble::rownames_to_column("variable") %>%
+  as_tibble()
+kable(cph_ph, format = "markdown")
+```
 
@@
 
-The proportional hazards test shows several covariates do not meet the
-proportional hazards assumption.
-
-The model is recreated below with another proportional hazards test. To
-improve the model and ensure the proportional hazards assumption is met,
-several covariates that do not meet the assumption are removed.
+### Adjusting the Model
+
+The results of the above assumptions demonstrated that several
+covariates violate the PH assumption. As such, the model will be
+recreated by removing the covariates that violate the PH assumption.
+After recreating the model, the PH assumption and the linearity
+assumption will be tested again.
+
+The table below demonstrates the new model information which has seven
+covariates: sm2, sm5, sm7, sm11, sm12, sm20, and sm21.
 
@@
 
-cox.zph(cox1.1)
-
-```
-
-The new model has seven covariates with only one not meeting the
-proportional hazards assumption. Although sm5 does not have a p value \>
-.05, the assumption overall holds for the new model and sm5 was not
-removed from the model because of its large impact on the hazard ratio
-shown in the forest plot. The new model's full summary is below.
-
-```{r}
-summary(cox1.1)
-
-```
-
-#### Checking Linearity Assumption
-
-Residuals can first be found by using the `resid(type = "martingale")`
-function and passing martingale in the type argument creating a vector
-of residuals that can be plotted against the range of values for each
-covariate. Example for one predictor
+kable(tidy(cox1.1), format = "markdown")
+g <- glance(cox1)
+g <- g %>% select(n, nevent, concordance, std.error.concordance, logLik, statistic.log, p.value.log, statistic.wald, p.value.wald, statistic.sc, p.value.sc)
+kable(g, format = "markdown")
+
+```
+
+The output below shows the p-values for the new model's PH assumption
+hypothesis test and Schoenfeld residuals. The global p-value is
+$> \alpha$ so the PH assumption has been met for the model even though
+the covariate sm5 has an individual p-value of $< \alpha$. As such, this
+new model does not violate the PH assumption and the analysis can
+continue.
+
+```{r}
+m_2 <- cox.zph(cox1.1) %>%
+  pluck("table") %>%
+  as.data.frame() %>%
+  tibble::rownames_to_column("variable") %>%
+  as_tibble()
+
+kable(m_2, format = "markdown")
+
+# Function to put ph_plots on one visual
+create_ph_plots <- function(coxmodel, row, col){
+  # Function takes argument that is a cox model from coxph()
+
+  phresults <- cox.zph(coxmodel)
+
+  # Define dimensions of visual
+  par(mfrow = c(row, col),
+      mar = c(2,2,4,1))
+
+  # Loop through each covariate creating a plot and adding it to the phplot list
+  for (i in 1:(length(phresults$table[,1])-1)){
+    phplots <- list()
+    phplots[i] <- plot(phresults[i], # Create plot for each iteration
+                       main = rownames(phresults$table)[i]) # Title of each plot is the rowname from coxmodel$table
+
+  }
+  mtext("Schoenfeld Residuals vs. Time for Covariates", side = 3, line = -1.5, outer = TRUE, cex = 1) # Title for Visual
+  par(mfrow = c(1,1)) # Reset plotting area
+}
+
+create_ph_plots(cox1.1, 3, 5)
+
+```
+
+Since a new model was generated, the below plots confirm that the
+linearity assumption is still met for this model.
 
@@
 
-```
-
-The plots of the martingale residuals vs the range of values for each
-covariate is shown above. For each covariate, there is no clear slope
-indicating the linearity assumption is true for the covariates in the
-model.
-
-#### Checking Independent Censoring
-
-In the dataset being used for this model, the training data does not
-contain censored data. The documentation available at the dataset's
-source clarifies all engines in the training data have experienced the
-desired event of failure at the end of each time series.
-
-### Adjusting the Model
-
-### Model Visualization
-
-survival curve
-
-The survival curve will be plotted by using the survival package's
-`survfit()` function. The object created will be plotted using `plot()`
-
-```{r}
-survfit <- survfit(cox1.1)
-
-```
-
-```{r}
-ggsurvplot(survfit, data = test, fun = "pct", surv.median.line = "none")
-```
-
-```{r}
-plot(survfit, mark.time=FALSE, xscale=365.24,
-	xlab='Years', ylab='Survival')
-lines(survfit[1], lwd=2, xscale=365.24)
 ```
@@
 max time value.
-
-Hazard Rate:
-$h(t|Z) = h_0(t) * e^{B_1Z_1 + B_2Z_2 + B_3Z_3+B_4Z_4+B_5Z_5}$
 

### 2025-04-07 slides.qmd
Update Intro and references
Updates 4/7

@@
 ---
-title: "Cox Proportional Hazards Model"
-subtitle: "Temp"
-author: "Jayme Reed (Advisor: Dr. Cohen)"
+title: "Predicting Survivability of NASA Aircraft Engines"
+subtitle: "Using Cox Proportional Hazards Model"
+author: "Jayme Reed & Brad Paton (Advisor: Dr. Cohen)"
 date: '`r Sys.Date()`'
 format:
-  revealjs
+  revealjs:
+    theme: moon
 course: Capstone Projects in Data Science
@@
 
-::: callout-important
-**Remember:** Your goal is to make your audience understand and care
-about your findings. By crafting a compelling story, you can effectively
-communicate the value of your data science project.
-
-Carefully read this template since it has instructions and tips to
-writing!
-
-More information about `revealjs`:
-<https://quarto.org/docs/reference/formats/presentations/revealjs.html>
-:::
-
-## Introduction  {.smaller}
-
--   Develop a storyline that captures attention and maintains interest.
-
--   Your audience is your peers
-
--   Clearly state the problem or question you're addressing.
-
--   Introduce why it is relevant needs.
-
--   Provide an overview of your approach.
-
-In kernel estimator, weight function is known as kernel function
-[@efr2008]. Cite this paper [@bro2014principal]. The GEE [@wang2014].
-The PCA [@daffertshofer2004pca]\*
-
 ## Methods
@@
 
--   Describe your data sources and collection process.
-
--   Present initial findings and insights through visualizations.
-
--   Highlight unexpected patterns or anomalies.
-
 ## Data Exploration and Visualization {.smaller}
-
-A study was conducted to determine how...
-
-```{r, warning=FALSE, echo=F, message=FALSE}
-# loading packages 
-library(tidyverse)
-library(knitr)
-library(ggthemes)
-library(ggrepel)
-library(dslabs)
-```
-
-```{r, warning=FALSE, echo=F}
-# Load Data
-#kable(head(murders))
-
-ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total))
-
-  ggplot1 + geom_point(aes(col=region), size = 4) +
-  geom_text_repel(aes(label=abb)) +
-  scale_x_log10() +
-  scale_y_log10() +
-  geom_smooth(formula = "y~x", method=lm,se = F)+
-  xlab("Populations in millions (log10 scale)") + 
-  ylab("Total number of murders (log10 scale)") +
-  ggtitle("US Gun Murders in 2010") +
-  scale_color_discrete(name = "Region")+
-      theme_bw()
-
-```
 
@@
 
--   Explain your data preprocessing and cleaning steps.
-
--   Present your key findings in a clear and concise manner.
-
--   Use visuals to support your claims.
-
--   **Tell a story about what the data reveals.**
-
-```{r}
-
-```
-
 ## Conclusion
 
--   Summarize your key findings.
-
--   Discuss the implications of your results.
-
 ## References

### 2025-04-08 index.qmd
4/8 update

@@
 
-Below is an example plot demonstrating scaled Shoenfeld partial
+Below is an example plot demonstrating scaled Schoenfeld partial
 residuals plotted against time to test the proportional hazards
@@
 Though CPH is typically used with health data, it does have applications
-in other fields. This paper will study the remaining useful life (RUL)
-of a NASA aircraft engine using data provided by NASA for the study on
-propagation modeling [@saxena2008]. In particular, this paper will be
-examining the data from the engine make two which has a training dataset
-and a testing dataset.
+in other fields. This paper will study the survivability of a NASA
+aircraft engine using data provided by NASA for the study on propagation
+modeling [@saxena2008]. In particular, this paper will be examining the
+data from the engine make two which has a training dataset and a testing
+dataset.
 
@@
 Below are tables showing the structure of the training and testing data
-after the transformations describd above.
+after the transformations described above.
 
@@
 test <- read.delim(testurl, header = TRUE, sep = '') %>%
-  mutate(status = 0)
-
-kable(head(test, n = 5), format = "markdown", caption = "Testing Data")
+  mutate(status = 0,
+         id = id + 260)
 
@@
 
-kable(head(train, n = 5), format = "markdown", caption = 'Training Data')
+data_all <- rbind(train, test)
+
+kable(head(data_all, n = 5), format = "markdown", caption = 'NASA Aircraft Engine Data')
 ```
@@
 maximum number of cycles an engine lasted was 378 while the minimum
-number of cycles an engined lasted was 128.
+number of cycles an engine lasted was 128.
 
 ```{r, echo = FALSE}
-distributions <- train %>% 
+train_grouped <- train %>% 
   group_by(id) %>%
-  summarise(time = max(time))
+  slice_max(time)
+
+data_grouped <- data_all %>%
+  group_by(id) %>%
+  slice_max(time)
 
 summarymetrics <- data.frame(c("Minimum", "Median", "Mean", "Standard Deviation", "Maximum"),
-                             c(min(distributions$time), 
-                               median(distributions$time),
-                               mean(distributions$time),
-                               sd(distributions$time),
-                               max(distributions$time)))
+                             c(min(train_grouped$time), 
+                               median(train_grouped$time),
+                               mean(train_grouped$time),
+                               sd(train_grouped$time),
+                               max(train_grouped$time)))
 
@@
 ```{r, echo=FALSE}
-Histogram <- ggplot(data = distributions,
+Histogram <- ggplot(data = train_grouped,
        aes(x = time)) +
@@
 
-box <- ggplot(data = distributions,
+box <- ggplot(data = train_grouped,
                   aes(y = time))+
@@
 
-To determine the intitial survival function, the Kaplan-Meier survival
-curve was generated and is shown below. The graph verifies that the
-median survival time is roughly 200 cycles and shows the maximum cycle.
-After that cycle has been reached, the survival probability plummets to
-0%.
-
-```{r}
-t <- as.matrix(train[, "time"])
-s <- as.matrix(train[, "status"])
+To determine the initial survival function, the Kaplan-Meier survival
+curve was generated and is shown below. The graph visualizes the
+survival curve and the information that was gathered through the
+histogram.
+
+```{r}
+t <- as.matrix(train_grouped[, "time"])
+s <- as.matrix(train_grouped[, "status"])
 
@@
 
-univ_models <- lapply( univ_formulas, function(x){coxph(x, data = train)})
+univ_models <- lapply( univ_formulas, function(x){coxph(x, data = data_grouped)})
 
@@
 Based on the above test, the covariates that will be considered for the
-CPH models are os1, os3, sm1, sm2, sm5, sm6, sm7, sm11, sm12, sm13,
-sm14, sm19, sm20, and sm21 which is fourteen out of twenty-four possible
+CPH models are os1, os3, sm1, sm5, sm6, sm7, sm11, sm12, sm13, sm14,
+sm19, sm20, and sm21 which is fourteen out of twenty-four possible
 covariates.
@@
 ```{r}
-cox1 <- coxph(data = train, id = id,
-             Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)
-
-cox2 <- coxph(data = train, id = id,
-             Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5)
-
-cox3 <- coxph(data = train, id = id,
-             Surv(time, status) ~ os1 + os3 + sm1 + sm5 + sm6 + sm7 + sm13 + sm14 + sm19 + sm21)
-
-cox4 <- coxph(data = train, id = id,
-             Surv(time, status) ~ sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)
-
-cox5 <- coxph(data = train, id = id,
-             Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5  + sm7 + sm11 + sm12 + sm13 + sm19 + sm20 )
-
-cox6 <- coxph(data = train, id = id,
-             Surv(time, status) ~ os1  + sm1 + sm2 + sm5  + sm7 + sm11 + sm12 + sm13 + sm19  + sm21)
-
-cox7 <- coxph(data = train, id = id,
-             Surv(time, status) ~ os1  + sm13 + sm14 + sm19 + sm20)
-
-cox8 <- coxph(data = train, id = id,
-             Surv(time, status) ~ os1 + os3 + sm21)
-
-cox9 <- coxph(data = train, id = id,
-             Surv(time, status) ~ os1 + os3 + sm2 + sm5 + sm6 + sm12 + sm13  + sm19 + sm20 + sm21)
-
-cox10 <- coxph(data = train, id = id,
-             Surv(time, status) ~  os3 + sm1 + sm2 + sm5 + sm6  + sm11 + sm12 + sm13  + sm19 + sm20 + sm21)
+surv_object <- Surv(data_grouped$time, data_grouped$status)
+
+cox1 <- coxph(data = data_grouped, surv_object ~ os1 + os3 + sm1 + sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)
+
+cox2 <- coxph(data = data_grouped, surv_object ~ os1 + os3 + sm1 + sm5)
+
+cox3 <- coxph(data = data_grouped, surv_object ~ os1 + os3 + sm1 + sm5 + sm6 + sm7 + sm13 + sm14 + sm19 + sm21)
+
+cox4 <- coxph(data = data_grouped, surv_object ~ sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)
+
+cox5 <- coxph(data = data_grouped, surv_object ~ os1 + os3 + sm1 + sm5  + sm7 + sm11 + sm12 + sm13 + sm19 + sm20 )
+
+cox6 <- coxph(data = data_grouped, surv_object ~ os1  + sm1 + sm5  + sm7 + sm11 + sm12 + sm13 + sm19  + sm21)
+
+cox7 <- coxph(data = data_grouped, surv_object ~ os1  + sm13 + sm14 + sm19 + sm20)
+
+cox8 <- coxph(data = data_grouped, surv_object ~ os1 + os3 + sm21)
+
+cox9 <- coxph(data = data_grouped, surv_object ~ os1 + os3 + sm5 + sm6 + sm12 + sm13  + sm19 + sm20 + sm21)
+
+cox10 <- coxph(data = data_grouped, surv_object ~  os3 + sm1 + sm5 + sm6  + sm11 + sm12 + sm13  + sm19 + sm20 + sm21)
 
@@
 ```{r}
-par(mfrow = c(3,3),
+par(mfrow = c(4,4),
       mar = c(2,2,4,1))
@@
 
-plot(train$os1, martingale, main = "Martingale Residuals: os1")
-abline(h = 0, col = "red")
-
-plot(train$os2, martingale, main = "Martingale Residuals: os2")
-abline(h = 0, col = "red")
-
-plot(train$sm1, martingale, main = "Martingale Residuals: sm1")
-abline(h = 0, col = "red")
-
-plot(train$sm2, martingale, main = "Martingale Residuals: sm2")
-abline(h = 0, col = "red")
-
-plot(train$sm5, martingale, main = "Martingale Residuals: sm5")
-abline(h = 0, col = "red")
-
-plot(train$sm6, martingale, main = "Martingale Residuals: sm6")
-abline(h = 0, col = "red")
-
-plot(train$sm7, martingale, main = "Martingale Residuals: sm7")
-abline(h = 0, col = "red")
-
-plot(train$sm11, martingale, main = "Martingale Residuals: sm11")
-abline(h = 0, col = "red")
-
-plot(train$sm12, martingale, main = "Martingale Residuals: sm12")
-abline(h = 0, col = "red")
-
-plot(train$sm13, martingale, main = "Martingale Residuals: sm13")
-abline(h = 0, col = "red")
-
-plot(train$sm14, martingale, main = "Martingale Residuals: sm14")
-abline(h = 0, col = "red")
-
-plot(train$sm20, martingale, main = "Martingale Residuals: sm20")
-abline(h = 0, col = "red")
-
-plot(train$sm21, martingale, main = "Martingale Residuals: sm21")
+plot(data_grouped$os1, martingale, main = "Martingale Residuals: os1")
+abline(h = 0, col = "red")
+
+plot(data_grouped$os2, martingale, main = "Martingale Residuals: os2")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm1, martingale, main = "Martingale Residuals: sm1")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm2, martingale, main = "Martingale Residuals: sm2")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm5, martingale, main = "Martingale Residuals: sm5")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm6, martingale, main = "Martingale Residuals: sm6")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm7, martingale, main = "Martingale Residuals: sm7")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm11, martingale, main = "Martingale Residuals: sm11")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm12, martingale, main = "Martingale Residuals: sm12")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm13, martingale, main = "Martingale Residuals: sm13")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm14, martingale, main = "Martingale Residuals: sm14")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm20, martingale, main = "Martingale Residuals: sm20")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm21, martingale, main = "Martingale Residuals: sm21")
 abline(h = 0, col = "red")
@@
 ```{r}
-cox1.1 <- coxph(data = train,Surv(time, status) ~ sm2 + sm5 + sm7 + sm11 + sm12 + sm20 + sm21)
+cox1.1 <- coxph(data = data_grouped,Surv(time, status) ~ sm2 + sm5 + sm7 + sm11 + sm12 + sm20 + sm21)
 
 kable(tidy(cox1.1), format = "markdown")
-g <- glance(cox1)
+g <- glance(cox1.1)
 g <- g %>% select(n, nevent, concordance, std.error.concordance, logLik, statistic.log, p.value.log, statistic.wald, p.value.wald, statistic.sc, p.value.sc)
@@
 ```{r}
-m_2 <- cox.zph(cox1.1) %>%
+ph_2 <- cox.zph(cox1.1) %>%
   pluck("table") %>%
@@
 
-kable(m_2, format = "markdown")
-
-# Function to put ph_plots on one visual
-create_ph_plots <- function(coxmodel, row, col){
-  # Function takes argument that is a cox model from coxph()
-
-  phresults <- cox.zph(coxmodel)
-
-  # Define dimensions of visual
-  par(mfrow = c(row, col),
-      mar = c(2,2,4,1))
-
-  # Loop through each covariate creating a plot and adding it to the phplot list
-  for (i in 1:(length(phresults$table[,1])-1)){
-    phplots <- list()
-    phplots[i] <- plot(phresults[i], # Create plot for each iteration
-                       main = rownames(phresults$table)[i]) # Title of each plot is the rowname from coxmodel$table
-
-  }
-  mtext("Schoenfeld Residuals vs. Time for Covariates", side = 3, line = -1.5, outer = TRUE, cex = 1) # Title for Visual
-  par(mfrow = c(1,1)) # Reset plotting area
-}
+kable(ph_2, format = "markdown")
 
@@
 
-plot(train$sm2, martingale, main = "Martingale Residuals: sm2")
-abline(h = 0, col = "red")
-
-plot(train$sm5, martingale, main = "Martingale Residuals: sm5")
-abline(h = 0, col = "red")
-
-plot(train$sm7, martingale, main = "Martingale Residuals: sm7")
-abline(h = 0, col = "red")
-
-plot(train$sm11, martingale, main = "Martingale Residuals: sm11")
-abline(h = 0, col = "red")
-
-plot(train$sm12, martingale, main = "Martingale Residuals: sm12")
-abline(h = 0, col = "red")
-
-plot(train$sm20, martingale, main = "Martingale Residuals: sm20")
-abline(h = 0, col = "red")
-
-plot(train$sm21, martingale, main = "Martingale Residuals: sm21")
+plot(data_grouped$sm2, martingale, main = "Martingale Residuals: sm2")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm5, martingale, main = "Martingale Residuals: sm5")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm7, martingale, main = "Martingale Residuals: sm7")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm11, martingale, main = "Martingale Residuals: sm11")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm12, martingale, main = "Martingale Residuals: sm12")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm20, martingale, main = "Martingale Residuals: sm20")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm21, martingale, main = "Martingale Residuals: sm21")
 abline(h = 0, col = "red")
@@
 coef <- data.frame(CoefficientValue = cox1.1$coefficients)
-kable(coef)
+kable(coef, format = "markdown")
 ```
@@
 ```{r}
-test$HazardRate <- predict(cox1.1,
-        newdata = test,
+data_grouped_reduced <- data_grouped %>% select(id, time, sm2, sm5, sm7, sm11, sm12, sm20, sm21, status)
+
+data_grouped_reduced$HazardRate <- predict(cox1.1,
+        newdata = data_grouped_reduced,
         type = "risk")
-
-predictiondata <- test %>%
-  group_by(id) %>%
-  slice_max(time) %>%
-  mutate(CumulativeHazard = cumsum(HazardRate)) %>%
-  select(id, time, sm2, sm5, sm7, sm11, sm12, sm20, sm21, status, HazardRate)
-
-kable(head(predictiondata %>% select(id, time, sm2, sm5, sm7, sm11, sm12, sm20, sm21, HazardRate)))
 
@@
 
-```{r}
-kable(head(predictiondata))
+The cumulative hazard of individuals in the model are shown in the visual below. Higher cumulative hazard indicates a higher level of risk at time `t`.
+
+```{r}
+sf <- survfit(cox1.1)
+
+ggsurvplot(fit = sf,
+           data = data_grouped_reduced, risk.table = TRUE, fun = "cumhaz")$plot +
+  ggtitle('Cumulative Hazard Over Time')
 
@@
 ```{r}
-predictiondata$SurvivalProbability <- predict(cox1.1, type = "survival", newdata = predictiondata)
-kable(head(predictiondata))
-
-```
-
-#### Predicting Time Until Failure
-
-Time until event can be modeled in R by passing the `type = "lp"`
-argument in the `stats::predict()` function to get a value for linear
-predictor.
-
-```{r}
+data_grouped_reduced$SurvivalProbability <- predict(cox1.1, type = "survival", newdata = data_grouped_reduced)
+kable(head(data_grouped_reduced))
+
+```
+
+The resulting survival curve from the model is shown below.
+
+```{r}
+
+survtimes <- data.frame(Percent = c("100% Survival", "< 25% Survival,", "< 50% Survival", "< 75% Survival", "<90% Survival"),
+                        Time = c(sf$time[sf$lower < 1][1], 
+                                    sf$time[sf$lower < .75][1], 
+                                    sf$time[sf$lower < .5][1], 
+                                    sf$time[sf$lower < .25][1], 
+                                    sf$time[sf$lower < .1][1]))
+
+ggsurvplot(fit = sf, data = data_grouped_reduced)$plot +
+  ggtitle('Survival Probability Over Time')
 
@@
 
--   Summarize your key findings.
-
--   Discuss the implications of your results.
+The implications of an engine failure are severe, with a possible result including an aircraft crash with a loss of human life and millions of dollars.
+Based on these implications, there is no acceptable room for failure for aircraft engines.
+The model created a confidence interval for estimating probabilities at each value of time.
+The values for time used will be the value where the probability of survival at the lower end of the confidence interval to exercise caution.
+
+```{r}
+kable(survtimes)
+
+```
+
+The lower end of the confidence interval for survival probability is shown to hold stable at 100% until `t = 128` with a gradual decrease afterwards.
+Half of the engines are expected to survive until `t = 221` and 90% of the engines are expected to fail at `t = 316`.
+These times are reflected by the survival curve shown above.
+The cumulative hazard over time plot confirms the hazard over time increases.
+The plot shows a low hazard until `t = 128` with a gradual increase.
+
+Based on these findings, the model can be applied by recommending for NASA not to use engines after 128 iterations have passed.
+The engines should undergo routine maintenance beforehand and should be evaluated to ensure the probability of survival calculated by the model remains at 100%.
+All engines should be equipped with sensors for the covariates used and readings should be applied to the model after each use of the engine.
+The engine should be removed from service if the resulting survival probability is less than 100%.
 

### 2025-04-11 index.qmd
Updates to index 4/11

@@
 ---
-title: "Predicting Remaining Useful Life of NASA Aircraft Engines"
+title: "Predicting Survival Probability of NASA Aircraft Engines"
 subtitle: "Using Cox Proportional Hazards Model"
@@
 $S(t|X=x)$, and the hazard ratio for an individual when compared to a
-reference individual $\frac{h(t|X=x)}{h(t|X=x_{ref}}$ [@nahhas2025]. If
-the resulting probability is greater than or equal to 50%, it is assumed
-the event has not occurred. If the resulting probability is less than
-50%, it is assumed the event has occurred. It is also possible to
+reference individual $\frac{h(t|X=x)}{h(t|X=x_{ref}}$ [@nahhas2025]. The
+equation is
+
+$$
+S(t) = e^{-H(t)}
+$$ {#eq-survprob}
+
+where, $H(t)$ is the cumulative hazard up to time $t$.
+
+If the resulting probability is greater than or equal to 50%, it is
+assumed the event has not occurred. If the resulting probability is less
+than 50%, it is assumed the event has occurred. It is also possible to
 generate a plot of the estimated survival curves for all values of $t$
-that are desired.
+that are desired. However, CPH is not able to predict the specific time
+the event is going to occur which is considered a limitation of the
+model.
 
@@
 Though CPH is typically used with health data, it does have applications
-in other fields. This paper will study the survivability of a NASA
-aircraft engine using data provided by NASA for the study on propagation
-modeling [@saxena2008]. In particular, this paper will be examining the
-data from the engine make two which has a training dataset and a testing
-dataset.
+in other fields. This paper will study the survival probability of a
+NASA aircraft engine using data provided by NASA for the study on
+propagation modeling [@saxena2008]. In particular, this paper will be
+examining the data from the engine design two which has a training
+dataset and a testing dataset.
 
@@
 
-Below are tables showing the structure of the training and testing data
-after the transformations described above.
+The table below shows the structure for the combined training and
+testing data after the transformations described above.
 
@@
 
-There are 260 engines in the training data. The histogram shows that the
-survival times of the engines are normally distributed and the box plot
-shows there six outliers in the dataset. The engines have a median
-survival time of 199 cycles with a standard deviation of 46.8. The
-maximum number of cycles an engine lasted was 378 while the minimum
-number of cycles an engine lasted was 128.
-
-```{r, echo = FALSE}
+There are 519 engines in the combined data with 260 engines in the
+training data and 259 engines in the testing data. The below metrics are
+generated using the training data as that data contains no censored
+data. The histogram shows that the survival times of the engines are
+normally distributed and the box plot shows there six outliers in the
+dataset. The engines have a median survival time of 199 cycles with a
+standard deviation of 46.78. The maximum number of cycles an engine
+lasted was 378 while the minimum number of cycles an engine lasted was
+128. Since the testing data contains all censored data, the basic
+metrics were not generated.
+
+```{r}
 train_grouped <- train %>% 
@@
 summarymetrics$Value <- round(summarymetrics$Value, digits = 2)
-kable(summarymetrics, format = "markdown")
+kable(summarymetrics, format = "markdown", align = "c")
 
@@
 
-To determine the initial survival function, the Kaplan-Meier survival
-curve was generated and is shown below. The graph visualizes the
-survival curve and the information that was gathered through the
-histogram.
+To determine the initial survival function, using the training data, the
+Kaplan-Meier survival curve was generated and is shown below. The graph
+visualizes the survival curve and the information that was gathered
+through the histogram.
 
@@
 
-**Covariate Selection**
+**Generating Models**
+
+Generation of the CPH models will be done using R, specifically the R
+package `survival`. To generate the CPH model with properly formatted
+survival data, a survival object is created by running
+`Surv(time, event)` where `time` is the time field and `event` is the
+event field. To generate the actual model, the survival object is then
+applied to the function `coxph(survival object ~ x, data)` where `x`
+represents the covariate(s) and `data` is the survival data that is
+being modeled [@R-base].
 
@@
 various sensors and operational setting fields that do not impact the
-survival rate of an engine. To determine which covariates will impact
-the survival probability, the below univariate test was run. The
-covariates in which the p-value is $< 0.05$ will be considered for the
-CPH models while the other covariates will be excluded.
+survival rate of an engine. To determine which covariates are
+statistically significant, the below univariate test was run. The test
+uses the Wald test to determine which covariates will significantly
+impact the model individually. The covariates in which the p-value is
+$< 0.05$ will be considered statistically significant.
 
@@
 
-Based on the above test, the covariates that will be considered for the
-CPH models are os1, os3, sm1, sm5, sm6, sm7, sm11, sm12, sm13, sm14,
-sm19, sm20, and sm21 which is fourteen out of twenty-four possible
-covariates.
-
-**Generating Models**
-
-Generation of all CPH models will be done using R, specifically the R
-package `survival`. To generate the CPH model with properly formatted
-survival data, a survival object is created by running
-`Surv(time, event)` where `time` is the time field and `event` is the
-event field. To generate the actual model, the survival object is then
-applied to the function `coxph(survival object ~ x, data)` where `x`
-represents one or multiple covariate(s) and `data` is the survival data
-that is being modeled.[@R-base].
-
-The table below provides the model number, the covariates used, and the
-concordance index. The concordance index generates a value to determine
-which model is the most accurate with the provided covariates. After
-determining the most accurate model, the CPH assumptions will be tested.
+Based on the above test, none of the covariates are statistically
+significant individually with all the data. As such, to determine the
+best model to use for this analysis, a forward and backward step-wise
+regression will be done to generate the models with the best Akaike
+Information Criterion (AIC) and Bayesian Information Criterion (BIC)
+values and compare the concordance values of those models with models
+that have covariates that were statistically significant when used in a
+CPH model with the training data. The model with the highest concordance
+value, and therefore, the most accurate will be used for the analysis.
+
+The table below provides the model number, the covariates used, the AIC
+and BIC from the step-wise regression if applicable, and the concordance
+index. The concordance index generates a value to determine which model
+is the most accurate with the provided covariates. After determining the
+most accurate model, the CPH assumptions will be tested.
 
@@
 
-cox1 <- coxph(data = data_grouped, surv_object ~ os1 + os3 + sm1 + sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)
-
-cox2 <- coxph(data = data_grouped, surv_object ~ os1 + os3 + sm1 + sm5)
-
-cox3 <- coxph(data = data_grouped, surv_object ~ os1 + os3 + sm1 + sm5 + sm6 + sm7 + sm13 + sm14 + sm19 + sm21)
-
-cox4 <- coxph(data = data_grouped, surv_object ~ sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)
-
-cox5 <- coxph(data = data_grouped, surv_object ~ os1 + os3 + sm1 + sm5  + sm7 + sm11 + sm12 + sm13 + sm19 + sm20 )
-
-cox6 <- coxph(data = data_grouped, surv_object ~ os1  + sm1 + sm5  + sm7 + sm11 + sm12 + sm13 + sm19  + sm21)
-
-cox7 <- coxph(data = data_grouped, surv_object ~ os1  + sm13 + sm14 + sm19 + sm20)
-
-cox8 <- coxph(data = data_grouped, surv_object ~ os1 + os3 + sm21)
-
-cox9 <- coxph(data = data_grouped, surv_object ~ os1 + os3 + sm5 + sm6 + sm12 + sm13  + sm19 + sm20 + sm21)
-
-cox10 <- coxph(data = data_grouped, surv_object ~  os3 + sm1 + sm5 + sm6  + sm11 + sm12 + sm13  + sm19 + sm20 + sm21)
-
-```
-
-```{r}
-concordance_results <- data.frame(c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5", 
-                                    "Model 6", "Model 7", "Model 8", "Model 9", "Model 10"), 
-                   c("os1, os3, sm1, sm2, sm5, sm6, sm7, sm11, sm12, sm13, sm14, sm19, sm20, sm21", 
-                     "os1, os3, sm1, sm2, sm5", 
-                     "os1, os3, sm1, sm5, sm6, sm7, sm13, sm14, sm19, sm21", 
-                     "sm5, sm6, sm7, sm11, sm12, sm13, sm14, sm19, sm20, sm21", 
-                     "os1, os3, sm1, sm2, sm5, sm7, sm11, sm12, sm13, sm19, sm20", 
-                     "os1, sm1, sm2, sm5, sm7, sm11, sm12, sm13, sm19, sm21",
-                     "os1, sm13, sm14, sm19, sm20", 
-                     "os1, os3, sm21",
-                     "os1, os3, sm2, sm5, sm6, sm12, sm13, sm19, sm20, sm21",
-                     "os3, sm1, sm2, sm5, sm6, sm11, sm12, sm13, sm19, sm20, sm21"),
-                   c(cox1$concordance[6], cox2$concordance[6], cox3$concordance[6], cox4$concordance[6], cox5$concordance[6], cox6$concordance[6], cox7$concordance[6], cox8$concordance[6], cox9$concordance[6], cox10$concordance[6]))
-
-colnames(concordance_results) <- c("Model", "Covariates", "Concordance Index")
-
-kable(concordance_results, format = "markdown")
-
-```
-
-Model 1 had the highest concordance index and used all of the covariates
-that were determined to be significant to the survival of the engine.
-The proceeding analysis will be done using model 1. The below tables
-provide the summary values from the model.
-
-```{r}
-kable(tidy(cox1), format = "markdown")
-g <- glance(cox1)
+all_cov <- coxph(data = data_grouped, surv_object ~ os1 + os2 + os3 + sm1 + sm2 + sm3 + sm4 + sm5 + sm6 + sm7 + sm8 + sm9 + sm10 + sm11 + sm12 + sm13 + sm14 + sm15 + sm16 + sm17 + sm18 + sm19 + sm20 + sm21)
+
+aic <- MASS::stepAIC(all_cov, trace = 0)
+n <- nrow(data_grouped)
+bic <- step(all_cov, direction = "both", k = log(n), trace = 0)
+
+train_sig <- lapply( univ_formulas, function(x){coxph(x, data = train_grouped)})
+
+cox1 <- coxph(data = data_grouped, surv_object ~ os1 + os2 + os3 + sm1 + sm2 + sm3 + sm4 + sm5 + sm6 + sm7 + sm8 + sm9 + sm10 + sm11 + sm12 + sm13 + sm14 + sm15 + sm16 + sm17 + sm18 + sm19 + sm20 + sm21)
+
+cox2 <- coxph(data = data_grouped, surv_object ~ os3 + sm3 + sm4 + sm5 + sm8 + sm9 + sm15 + sm16 + sm17 + sm18 + sm20)
+
+cox3 <- coxph(data = data_grouped, surv_object ~ os3 + sm3 + sm4 + sm8 + sm9 + sm15 + sm18)
+
+cox4 <- coxph(data = data_grouped, surv_object ~ os3 + sm13 + sm14 + sm19)
+
+```
+
+```{r, echo = FALSE}
+stepwise_results <- data.frame(c("model 1", "model 2", "model 3", "model 4"), 
+                               c("All covariates", "os3, sm3 - sm5, sm8 - sm9, sm15 - sm18, sm20", "os3, sm3 - sm4, sm8-sm9, sm15, sm18", "os3, sm13 - sm14, sm19"),
+                               c("2449.92", "2431.88", "N/A", "N/A"),
+                               c("2547.71", "N/A", "2461.31", "N/A"),
+                               c(cox1$concordance[6], cox2$concordance[6], cox3$concordance[6], cox4$concordance[6]))
+
+colnames(stepwise_results) <- c("Model", "Covariates", "AIC", "BIC", "Concordance")
+
+kable(stepwise_results, format = "markdown")
+
+```
+
+Based on the above model results, the model that has the highest
+concordance value is model 1 which has all of the covariates. However,
+before proceeding with that model, a test for linear dependency will be
+run to determine all covariates are linearly independent.
+
+```{r}
+kable(alias(cox1), format = "markdown")
+```
+
+The above result indicates that sm19 needs to be removed from the model
+as it linearly dependent on another value. In addition, sm16 resulted in
+a possible infinite coefficient so that covariate has also been removed.
+The proceeding analysis will be done using model 1 without sm16 and
+sm19. The below tables provide the summary values from the model.
+
+```{r}
+cox1.1 <- coxph(data = data_grouped, surv_object ~ os1 + os2 + os3 + sm1 + sm2 + sm3 + sm4 + sm5 + sm6 + sm7 + sm8 + sm9 + sm10 + sm11 + sm12 + sm13 + sm14 + sm15 + sm17 + sm18 + sm20 + sm21)
+
+ci <- cox1.1$concordance[6]
+```
+
+```{r, echo = FALSE}
+kable(tidy(cox1.1), format = "markdown")
+
+g <- glance(cox1.1)
 g <- g %>% select(n, nevent, concordance, std.error.concordance, logLik, statistic.log, p.value.log, statistic.wald, p.value.wald, statistic.sc, p.value.sc)
+
 kable(g, format = "markdown")
@@
 
-```{r}
-plot_model(cox1,
+```{r, echo = FALSE}
+plot_model(cox1.1,
            dot.size = 2,
@@
 The plots of the Martingale residuals vs the range of values for each
-covariate is shown below. For each covariate, there is no clear slop
+covariate is shown below. For each covariate, there is no clear slope
 indicating the linearity assumption is true for the covariates in the
@@
 ```{r}
+martingale <- resid(cox1.1, type = "martingale")
+```
+
+```{r, echo = FALSE}
 par(mfrow = c(4,4),
@@
 
-martingale <- resid(cox1, type = "martingale")
-
 plot(data_grouped$os1, martingale, main = "Martingale Residuals: os1")
@@
 
+plot(data_grouped$os3, martingale, main = "Martingale Residuals: os3")
+abline(h = 0, col = "red")
+
 plot(data_grouped$sm1, martingale, main = "Martingale Residuals: sm1")
@@
 
+plot(data_grouped$sm3, martingale, main = "Martingale Residuals: sm3")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm4, martingale, main = "Martingale Residuals: sm4")
+abline(h = 0, col = "red")
+
 plot(data_grouped$sm5, martingale, main = "Martingale Residuals: sm5")
@@
 
+plot(data_grouped$sm8, martingale, main = "Martingale Residuals: sm8")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm9, martingale, main = "Martingale Residuals: sm9")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm10, martingale, main = "Martingale Residuals: sm10")
+abline(h = 0, col = "red")
+
 plot(data_grouped$sm11, martingale, main = "Martingale Residuals: sm11")
@@
 
+plot(data_grouped$sm15, martingale, main = "Martingale Residuals: sm15")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm17, martingale, main = "Martingale Residuals: sm17")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm18, martingale, main = "Martingale Residuals: sm18")
+abline(h = 0, col = "red")
+
 plot(data_grouped$sm20, martingale, main = "Martingale Residuals: sm20")
@@
 The `cox.zph` function can be used to check the proportional hazards
-assumption. As mentioned above, the function calculates p values for
-each covariate. The global p-value will first be compared to
-$\alpha = 0.05$. If that p-value is $<\alpha$, then at least one
-covariate violates the PH assumption and the model will need to be
-adapted.
+assumption. As mentioned, the function calculates p values for each
+covariate. The global p-value will first be compared to $\alpha = 0.05$.
+If that p-value is $<\alpha$, then at least one covariate violates the
+PH assumption and the model will need to be adapted.
 
 The output below shows the p-values for the hypothesis test and
-Schoenfeld residuals. The global p-value is $< \alpha$ so at least one
-covariate violates the PH assumption. Examining each individual
-covariate through both p-values and the plot of the Schoenfeld residuals
-shows that the covariates os1, os3, sm1, sm5, sm13, and sm14 all violate
-the PH assumption.
-
-```{r}
-cph_ph <- cox.zph(cox1) %>%
+Schoenfeld residuals. The global p-value is $> \alpha$ so none of the
+covariates violates the PH assumption on a global level. Examining each
+individual covariate through both p-values and the plot of the
+Schoenfeld residuals does show that os1, os3, sm13, and sm14 violate the
+PH assumption. To ensure the model is as accurate as possible, the model
+will addapted without those covariates and the global p-values will be
+compared.
+
+```{r}
+cph_ph <- cox.zph(cox1.1) %>%
   pluck("table") %>%
@@
 
-```{r}
+```{r, echo = FALSE}
 # Function to put ph_plots on one visual
@@
 
-create_ph_plots(cox1, 3, 5)
+create_ph_plots(cox1.1, 3, 5)
 ```
@@
 The results of the above assumptions demonstrated that several
-covariates violate the PH assumption. As such, the model will be
-recreated by removing the covariates that violate the PH assumption.
-After recreating the model, the PH assumption and the linearity
-assumption will be tested again.
-
-The table below demonstrates the new model information which has seven
-covariates: sm2, sm5, sm7, sm11, sm12, sm20, and sm21.
-
-```{r}
-cox1.1 <- coxph(data = data_grouped,Surv(time, status) ~ sm2 + sm5 + sm7 + sm11 + sm12 + sm20 + sm21)
-
-kable(tidy(cox1.1), format = "markdown")
-g <- glance(cox1.1)
+covariates violate the PH assumption on a local level. As such, the
+model will be recreated by removing the covariates that violate the PH
+assumption. After recreating the model, the PH assumption and the
+linearity assumption will be tested again with the new PH global p-value
+being compared to the original model's global p-value.
+
+The table below demonstrates the new model information which has
+eighteen covariates: os2, sm1 - sm12, sm15, sm17, sm18, sm20, sm21.
+
+```{r}
+cox1.2 <- coxph(data = data_grouped, surv_object ~ os2 + sm1 + sm2 + sm3 + sm4 + sm5 + sm6 + sm7 + sm8 + sm9 + sm10 + sm11 + sm12 + sm15 + sm17 + sm18 + sm20 + sm21)
+
+kable(tidy(cox1.2), format = "markdown")
+g <- glance(cox1.2)
 g <- g %>% select(n, nevent, concordance, std.error.concordance, logLik, statistic.log, p.value.log, statistic.wald, p.value.wald, statistic.sc, p.value.sc)
@@
 The output below shows the p-values for the new model's PH assumption
-hypothesis test and Schoenfeld residuals. The global p-value is
-$> \alpha$ so the PH assumption has been met for the model even though
-the covariate sm5 has an individual p-value of $< \alpha$. As such, this
-new model does not violate the PH assumption and the analysis can
-continue.
-
-```{r}
-ph_2 <- cox.zph(cox1.1) %>%
+hypothesis test and Schoenfeld residuals. The global p-value is now
+$< \alpha$ so the PH assumption has not been met for the new model. As
+such,, the remaining analysis will be done using the first model that
+had a global p-value $>\alpha$ which has all but two covariates.
+
+```{r, echo = FALSE}
+ph_2 <- cox.zph(cox1.2) %>%
   pluck("table") %>%
@@
 
-create_ph_plots(cox1.1, 3, 5)
-
-```
-
-Since a new model was generated, the below plots confirm that the
-linearity assumption is still met for this model.
-
-```{r}
-
-par(mfrow = c(3,3),
-      mar = c(2,2,4,1))
-
-martingale <- resid(cox1.1, type = "martingale")
-
-plot(data_grouped$sm2, martingale, main = "Martingale Residuals: sm2")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm5, martingale, main = "Martingale Residuals: sm5")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm7, martingale, main = "Martingale Residuals: sm7")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm11, martingale, main = "Martingale Residuals: sm11")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm12, martingale, main = "Martingale Residuals: sm12")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm20, martingale, main = "Martingale Residuals: sm20")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm21, martingale, main = "Martingale Residuals: sm21")
-abline(h = 0, col = "red")
-
-```
-
-### Model Applications and Predictions
-
-The model created can be used to calculate several relevant metrics when
-predicting remaining time for the engines until expected failure.
-Predictions will be made on the largest value for time for each
-individual engine.
-
-#### Model Coefficients
+create_ph_plots(cox1.2, 3, 5)
+
+```
+
+If the new model had passed the PH assumption, the linearity assumption
+would have been checked again.
+
+### Model Results
+
+**Evaluation**
+
+The CPH model that was selected has a concordance value of `{r} ci`
+which indicates that it is more than 50% accurate in predicting the
+survival of the machines. However, this model is not completely accurate
+so there is an expectation of some error in the model survival
+probabilities which will lead to caution in providing results in
+application settings.
+
+**Calculations**
 
@@
 
-#### Hazard Rate Calculation
-
-The coefficients above can be used to calculate the hazard rate at time
-$t$. The hazard rate will be calculated for each engine at the engines
-max time value.
-
-The engines with the higher hazard rates are at a higher risk of
-experiencing failure.
-
-```{r}
-data_grouped_reduced <- data_grouped %>% select(id, time, sm2, sm5, sm7, sm11, sm12, sm20, sm21, status)
+The coefficients above can be used to calculate the hazard rate
+[@eq-hazardrate], the cumulative hazard value[@eq-cumulativehr], and the
+survival probability [@eq-survprob]. The hazard rate at time $t$ will be
+calculated for each engine at the engines max time value. The cumulative
+hazard value is the total hazard accumulated for the subject over time
+and is used to estimate the risk for the subject at a point in time. The
+engines that have higher hazard rates or higher cumulative hazard values
+are at a greater risk of experiencing failure. The below table provides
+the hazard rate and suvivial probability for the first 10 engines in the
+dataset.
+
+```{r}
+data_grouped_reduced <- data_grouped %>% select(id, time, os1, os2, os3, sm1, sm2, sm3, sm4, sm5, sm6, sm7, sm8, sm9, sm10, sm11, sm12, sm13, sm14, sm15, sm17, sm18, sm20, sm21, status)
 
@@
 
-```
-
-#### Cumulative Hazard Calculation
-
-The cumulative hazard represents the total hazard accumulated for the
-individual over time and like the hazard rate, is another measure that
-can be used to estimate the risk for the individual at a point in time.
-The cumulative hazard for each individual is shown below. Higher values
-represent a higher chance of failure.
-
-$H(t) = \int_0^t h(t) dt$ where, $H(t)$ is the cumulative hazard from
-time 0 to time $(t)$ $h(t)$ is the hazard rate at time $t$
-
-The cumulative hazard of individuals in the model are shown in the visual below. Higher cumulative hazard indicates a higher level of risk at time `t`.
-
-```{r}
+data_grouped_reduced$SurvivalProbability <- predict(cox1.1, type = "survival", newdata = data_grouped_reduced)
+
+print_data <- data_grouped_reduced %>% select(id, time, status, HazardRate, SurvivalProbability)
+
+kable(head(print_data, n = 10), format = "markdown")
+```
+
+The below graphs demonstrate the cumulative hazard over time and the
+survival curve from the prediction model.
+
+```{r, echo = FALSE}
 sf <- survfit(cox1.1)
@@
 
-#### Survival probability
-
-While cumulative hazard can be used by itself to give an idea of an
-individual's risk, it can also be used to calculate a probability of
-survival.
-
-Probability Calculation: $S(t) = e^{-H(t)}$ $H(t)$ is the cumulative
-hazard up to time $t$.
-
-```{r}
-data_grouped_reduced$SurvivalProbability <- predict(cox1.1, type = "survival", newdata = data_grouped_reduced)
-kable(head(data_grouped_reduced))
-
-```
-
-The resulting survival curve from the model is shown below.
-
-```{r}
-
-survtimes <- data.frame(Percent = c("100% Survival", "< 25% Survival,", "< 50% Survival", "< 75% Survival", "<90% Survival"),
+```{r, echo = FALSE}
+survtimes <- data.frame(Percent = c("100% Survival", "75% Survival,", "50% Survival", "25% Survival", "10% Survival"),
                         Time = c(sf$time[sf$lower < 1][1], 
@@
 
-The implications of an engine failure are severe, with a possible result including an aircraft crash with a loss of human life and millions of dollars.
-Based on these implications, there is no acceptable room for failure for aircraft engines.
-The model created a confidence interval for estimating probabilities at each value of time.
-The values for time used will be the value where the probability of survival at the lower end of the confidence interval to exercise caution.
-
-```{r}
-kable(survtimes)
-
-```
-
-The lower end of the confidence interval for survival probability is shown to hold stable at 100% until `t = 128` with a gradual decrease afterwards.
-Half of the engines are expected to survive until `t = 221` and 90% of the engines are expected to fail at `t = 316`.
-These times are reflected by the survival curve shown above.
-The cumulative hazard over time plot confirms the hazard over time increases.
-The plot shows a low hazard until `t = 128` with a gradual increase.
-
-Based on these findings, the model can be applied by recommending for NASA not to use engines after 128 iterations have passed.
-The engines should undergo routine maintenance beforehand and should be evaluated to ensure the probability of survival calculated by the model remains at 100%.
-All engines should be equipped with sensors for the covariates used and readings should be applied to the model after each use of the engine.
-The engine should be removed from service if the resulting survival probability is less than 100%.
+The implications of an engine failure are severe, as failure can lead to
+injury or death and the loss of millions of dollars. As such,
+understanding when an engine will fail allows for maintenance to occur
+before negative impacts occur. While CPH does not allow for the
+prediction of specific failure times, it does allow for the creation
+confidence intervals which can be used to gain an understanding of when
+engines are most likely to fail. The table below displays the percent of
+survival at the lower end of the confidence interval to exercise
+caution.
+
+```{r}
+kable(survtimes, format = "markdown")
+
+```
+
+The lower end of the confidence interval for survival probability is
+shown to hold stable at 100% until `t = 128` with a gradual decrease
+afterwards. Half of the engines are expected to survive until `t = 221`
+and 90% of the engines are expected to fail at `t = 316`. These results
+can be viewed in both the above survival graph and the cumulative hazard
+plot which shows a low hazard until `t = 128` with a gradual increase.
+
+Based on these findings, the model can be applied by recommending for
+NASA not to use engines after 128 iterations have passed. The engines
+should undergo routine maintenance beforehand and should be evaluated to
+ensure the probability of survival calculated by the model remains at
+100%. All engines should be equipped with sensors for the covariates
+used and readings should be applied to the model after each use of the
+engine. The engine should be removed from service if the resulting
+survival probability is less than 100%.
 

### 2025-04-13 index.qmd
Slides Update 4/13

@@
 author: "Jayme Reed & Brad Paton (Advisor: Dr. Cohen)"
-date: '`r Sys.Date()`'
+date: last-modified
+date-format: long
 format:
@@
 probability of survival. Statistical methods such as the Wald Test can
-be used to determine which ones to use in the final CPH model.
-Generating a univariate CPH model for each covariate and using the Wald
-test with a significance level of $\alpha = 0.05$ will allow for
-determination of if a covariate is significant. Any covariates that are
-determined to be not significant to the survival probability can be
+be used to determine which ones are statistically significant
+individually and may have an impact in a multivariate model. Generating
+a univariate CPH model for each covariate and using the Wald test with a
+significance level of $\alpha = 0.05$ will allow for determination of if
+a covariate is significant. Additional tests for models will involve
+using a step-wise regression and comparing the results based on the
+assumptions that are met and the concordance index. Any covariates that
+are determined to be not significant to the survival probability can be
 removed from the final CPH model.

### 2025-04-13 slides.qmd
Slides Update 4/13

@@
 ---
-title: "Predicting Survivability of NASA Aircraft Engines"
+title: "Predicting Survival Probability of NASA Aircraft Engines"
 subtitle: "Using Cox Proportional Hazards Model"
 author: "Jayme Reed & Brad Paton (Advisor: Dr. Cohen)"
-date: '`r Sys.Date()`'
+date: last-modified
+date-format: long
 format:
@@
     theme: moon
+    smaller: true
 course: Capstone Projects in Data Science
@@
   message: false
+  echo: false
 editor: 
@@
 
-## Methods
+```{r, echo = FALSE, warning=FALSE, include=FALSE}
+#Load packages
+library(tidyverse)
+library(gridExtra)
+library(grid)
+library(survival)
+library(survminer)
+library(ggpubr)
+library(magrittr)
+library(ggfortify)
+library(knitr)
+library(sjPlot)
+library(broom)
+```
 
--   Detail the models or algorithms used.
+```{r, echo = FALSE}
+#generation of the model for sample plots
+example_surv_object <- Surv(time = lung$time, event = lung$status)
 
--   Justify your choices based on the problem and data.
+example_coxmodel <- coxph(example_surv_object ~ meal.cal, data = lung)
+```
 
-## Data Exploration and Visualization
+## Cox Proportional Hazards (CPH) Model
 
-## Data Exploration and Visualization {.smaller}
+What is it?
 
-## Modeling and Results
+-   A statistical regression method specializing in modeling
+    time-to-event predictions with survival data [@Abeysekera2009]
+    -   Survival data has a value for time and an indicator column for
+        an event
+-   Is a method that can deal with censored data
+    -   Censored data is when the information about an individual in a
+        study is only known for a certain period of time [@klein2005]
+-   Primarily used in the health field but has applications in
+    predicting bank failure, the survival probability of machines, and
+    insurance likelihood payouts
+
+## Limitations
+
+-   The model assumes that as time goes on, the survival probability
+    will approach zero with no survivors [@asghar2024]
+
+-   The proportional hazards assumption can limit the ability to
+    correctly predict the effect of a variable [@jiang2024]
+
+-   The covariate selection can become biased and may not accurately
+    represent the true data [@wang2025],[@zhang2025]
+
+-   The model cannot provide a specific value for when the event will
+    happen, only the probability of when the event might happen
+
+## Mathematical Formulas
+
+-   Survival Function:
+
+    $S(x) = \int_{x}^{\infty} f(x)dx$
+
+-   Hazard Function:
+
+    $h(x) = \lim\limits_{\Delta x \rightarrow 0} \frac{P[x \leq X < x + \Delta x|X \geq x]}{\Delta x}$
+
+-   CPH Model Hazard Function:
+
+    $h(t|\mathbf Z) = h_0(t)\text{exp}(\sum\limits_{k=1}^{p} \beta_kZ_k)$
+
+-   Proportional Hazards Ratio:
+
+    $\frac{h(t|\mathbf Z)}{h(t|\mathbf Z*)} = \text{exp}[\sum\limits_{k=1}^{p} \beta_k(Z_k - Z_k^*)]$
+
+-   Cumulative Hazard Function:
+
+    $H(x) = \int_0^x h(u) du$
+
+## Assumptions
+
+There are four assumptions for CPH:
+
+::: incremental
+-   Independence assumption
+
+    -   Assumes that the survival times of observed subjects are
+        independent of each other [@nahhas2025]
+
+-   Non-informative Censoring Assumption
+
+    -   Assumes that censoring is non-informative [@nahhas2025]
+:::
+
+## Assumptions, Cont.
+
+-   Linearity Assumption
+
+    ::: incremental
+    -   Assumes the relationship between covariates and the outcome is a
+        linear relationship
+
+    -   This is tested using the Martingale residuals for each covariate
+        with the equation Martingale Residuals = Observed Events -
+        Expected Results [@nahhas2025]
+
+        -   The residuals are examined visually
+
+        -   If linear and appear to have a slop of zero, the assumption
+            is not violated [@Amini2015]
+    :::
+
+```{r,echo=FALSE}
+example_martingale <- resid(example_coxmodel, type = "martingale")
+
+plot(example_martingale, 
+     main = "Martingale Residuals: Example Covariate",
+     ylab = "Martingale Residuals for Example Covariate",
+     xlab = "Value Range for Example Covariate")
+abline(h = 0, col = "red")
+
+```
+
+## Assumptions, Cont.
+
+-   Proportional Hazards Assumption
+
+    ::: incremental
+    -   The ratio of hazard rates for any two subjects must be constant
+        at all times [@Bustan2018]
+
+    -   To test this assumption, the below hypothesis test is used:
+
+        $$
+        \begin{align*}
+        H_0&:\; \text{the log hazard ratio for each covariate remains constant over time} \\
+        H_1&: \; \text{the log hazard ratio for each covariate does not remain constant over time}
+        \end{align*}
+        $$
+
+        -   For this test, the goal is to *not* reject $H_0$
+
+    -   The scale Schoenfeld partial residuals are used to find the
+        p-value for this test
+
+        -   The Schoenfeld residuals are the difference between the
+            value of the covariate and the expected value of the
+            covariate at the time of failure [@klein2005]
+
+        -   There will be multiple p-values generated: global and
+            individual
+
+            -   If the global p-value is $< 0.05$, then the model has at
+                least one covariate that violates the proportional
+                hazards assumption and the individual p-values must be
+                examined
+
+            -   If the global p-value is $> 0.05$, then all covariates
+                meet the proportional hazard assumption [@nahhas2025]
+
+    Example of Schoenfeld plot and results
+
+    ```{r, echo = FALSE}
+    example_phtest <- cox.zph(example_coxmodel)
+
+    plot(example_phtest[1],
+         ylab = "Covariate Schoenfeld Residual",
+         xlab = "Time")
+
+    multi <- coxph(example_surv_object ~ meal.cal + age, data = lung)
+    ph_tibble <- cox.zph(multi) %>%
+      pluck("table") %>%
+      as.data.frame() %>%
+      tibble::rownames_to_column("variable") %>%
+      as_tibble()
+    kable(ph_tibble, format = "markdown")
+
+    ```
+    :::
+
+## Model Visualization - Kaplan Meier
+
+## Model Visualization - Forest Plot
+
+## Evaluation
+
+## Prediction
+
+## Data Structure, Exploration, and Visualization
+
+## Creating CPH Model
+
+## Checking Assumptions
+
+## Adjusting the Model
+
+## Model Results
 

### 2025-04-15 index.qmd
update 4/14
Update 4/15 pt1

@@
 the medical world is the most obvious use of survival predictions, the
-same methods can be applied to predicting bank failure, the remaining
-useful life of machines, machinery maintenance timelines, and insurance
+same methods can be applied to predicting bank failure, the survival
+probablity of machines, machinery maintenance timelines, and insurance
 likelihood of payouts. There are various survival methods that can be
@@
 confidence interval for the effect. A positive effect indicates a
-positive correlation with the hazard with a higher number signifying a
-more severe effect on the hazard ratio.
+positive correlation with the hazard ratio and the higher the number the
+more severe the effect is on the hazard ratio.
 
@@
 shown to hold stable at 100% until `t = 128` with a gradual decrease
-afterwards. Half of the engines are expected to survive until `t = 221`
-and 90% of the engines are expected to fail at `t = 316`. These results
+afterwards. Half of the engines are expected to survive until `t = 245`
+and 90% of the engines are expected to fail at `t = 347`. These results
 can be viewed in both the above survival graph and the cumulative hazard

### 2025-04-15 slides.qmd
Update 4/15 pt1

--- 2025-04-07 slides.qmd
@@
 
-## Methods
+```{r, warning=FALSE, include=FALSE}
+#Load packages
+library(tidyverse)
+library(gridExtra)
+library(grid)
+library(survival)
+library(survminer)
+library(ggpubr)
+library(magrittr)
+library(ggfortify)
+library(knitr)
+library(sjPlot)
+library(broom)
+```
 
--   Detail the models or algorithms used.
+```{r, echo = FALSE}
+#generation of the model for sample plots
+example_surv_object <- Surv(time = lung$time, event = lung$status)
 
--   Justify your choices based on the problem and data.
+example_coxmodel <- coxph(example_surv_object ~ meal.cal, data = lung)
+```
 
-## Data Exploration and Visualization
+## Cox Proportional Hazards (CPH) Model
 
-## Data Exploration and Visualization {.smaller}
+What is it?
 
-## Modeling and Results
+::: incremental
+-   A statistical regression method specializing in modeling
+    time-to-event predictions with survival data [@Abeysekera2009]
+
+    -   Survival data has a value for time and an indicator column for
+        an event
+
+-   Is a method that can deal with censored data
+
+    -   Censored data is when the information about an individual in a
+        study is only known for a certain period of time [@klein2005]
+
+-   Primarily used in the health field but has applications in
+    predicting bank failure, the survival probability of machines, and
+    insurance likelihood payouts
+:::
+
+## Limitations
+
+::: incremental
+-   The model assumes that as time goes on, the survival probability
+    will approach zero with no survivors [@asghar2024]
+
+-   The proportional hazards assumption can limit the ability to
+    correctly predict the effect of a variable [@jiang2024]
+
+-   The covariate selection can become biased and may not accurately
+    represent the true data [@wang2025],[@zhang2025]
+
+-   The model cannot provide a specific value for when the event will
+    happen, only the probability of when the event might happen
+:::
+
+## Mathematical Formulas
+
+-   Survival Function: $S(x) = \int_{x}^{\infty} f(x)dx$
+
+-   Hazard Function:
+    $h(x) = \lim\limits_{\Delta x \rightarrow 0} \frac{P[x \leq X < x + \Delta x|X \geq x]}{\Delta x}$
+
+-   CPH Model Hazard Function:
+    $h(t|\mathbf Z) = h_0(t)\text{exp}(\sum\limits_{k=1}^{p} \beta_kZ_k)$
+
+-   Proportional Hazards Ratio:
+    $\frac{h(t|\mathbf Z)}{h(t|\mathbf Z*)} = \text{exp}[\sum\limits_{k=1}^{p} \beta_k(Z_k - Z_k^*)]$
+
+-   Cumulative Hazard Function: $H(x) = \int_0^x h(u) du$
+
+-   Concordance Index: $C = \frac{c + \frac{t_x}{2}}{c + d + t_x}$
+
+-   Survival Probability: $S(t) = e^{-H(t)}$, where $H(t)$ is the above
+    cumulative hazard function
+
+## Assumptions
+
+There are four assumptions for CPH:
+
+::: incremental
+-   Independence assumption
+
+    -   Assumes that the survival times of observed subjects are
+        independent of each other [@nahhas2025]
+
+-   Non-informative Censoring Assumption
+
+    -   Assumes that censoring is non-informative [@nahhas2025]
+
+-   Linearity Assumption
+
+    -   Assumes the relationship between covariates and the outcome is a
+        linear relationship [@nahhas2025]
+
+-   Proportional Hazards Assumption
+
+    -   Assumes the ratio of hazards rates for any two subjects are
+        constant at all times [@Bustan2018]
+:::
+
+## Linearity Assumption
+
+-   Tested using the Martingale residuals for each covariate using the
+    equation Martingale Residuals = Observed Events - Expected Results
+    [@nahhas2025]
+
+-   If the plots are linear and appear to have a slope of zero, the
+    assumption is not violated [@Amini2015]
+
+```{r}
+#| fig-align: center
+example_martingale <- resid(example_coxmodel, type = "martingale")
+
+plot(example_martingale, 
+     main = "Martingale Residuals: Example Covariate",
+     ylab = "Martingale Residuals for Example Covariate",
+     xlab = "Value Range for Example Covariate")
+abline(h = 0, col = "red")
+
+```
+
+## Proportional Hazards Assumption
+
+-   The test is verified using the Schoenfeld partial residuals which
+    are the difference between the value of the covariate and the
+    expected value of the covariate at the time of failure [@klein2005]
+
+::: panel-tabset
+#### Graph
+
+```{r}
+example_phtest <- cox.zph(example_coxmodel)
+
+plot(example_phtest[1],
+     ylab = "Covariate Schoenfeld Residual",
+     xlab = "Time")
+```
+
+#### Table
+
+```{r}
+multi <- coxph(example_surv_object ~ meal.cal + age, data = lung)
+ph_tibble <- cox.zph(multi) %>%
+  pluck("table") %>%
+  as.data.frame() %>%
+  tibble::rownames_to_column("variable") %>%
+  as_tibble()
+kable(ph_tibble, format = "markdown")
+```
+:::
+
+## Kaplan-Meier Survival Curve
+
+::: incremental
+-   Provides visualization for the Kaplan-Meier estimator
+
+-   Is considered to be well defined up until the largest observed study
+    time $t_{max}$
+
+-   Demonstrates the time where the event being modeled is expected to
+    occur
+:::
+
+```{r}
+#| fig-align: center
+example_curve <- survfit(example_coxmodel)
+
+ggsurvplot(example_curve,
+           data = lung,
+           title = "Example Survival Curve",
+           xlab = "Time",
+           ylab = "Survival Probability",
+           legend = "none")
+```
+
+## Forest Plot
+
+-   Provides visualization of the effects of each covariate on the
+    hazard ratio
+
+-   A positive effect indicates a positive correlation with the hazard
+    ratio
+
+```{r, warning=FALSE}
+#| fig-align: center
+forest <- coxph(example_surv_object ~ age + sex, data = lung)
+
+plot_model(forest,
+           dot.size = 1,
+           line.size = 1,
+           colors = "red")+
+  theme_bw()+
+  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
+  ylab(label = "Covariates")+
+  xlab(label = "Estimated Effect on Hazard Ratio")
+
+```
+
+## Evaluation
+
+## Prediction
+
+## Data Structure, Exploration, and Visualization
+
+## Creating CPH Model
+
+## Checking Assumptions
+
+## Adjusting the Model
+
+## Model Results
 

### 2025-04-15 index.qmd
Update 4/15 pt2

@@
 a covariate is significant. Additional tests for models will involve
-using a step-wise regression and comparing the results based on the
+using a stepwise regression and comparing the results based on the
 assumptions that are met and the concordance index. Any covariates that
@@
 Forest plots can be used to visualize the effects of each covariate in
-the model on the hazard ratio. The resulting plot contains each
+the model has on the hazard ratio. The resulting plot contains each
 covariate on the y axis, with the estimated effect on the hazard ratio
@@
 confidence interval for the effect. A positive effect indicates a
-positive correlation with the hazard ratio and the higher the number the
-more severe the effect is on the hazard ratio.
+positive correlation with the hazard ratio where the higher the number
+is the more severe the effect on the hazard ratio.
 
@@
 The concordance index will produce a value between 0 and 1. A value of 1
-means all pairs are correctly ordered and a value 0 being the
+means all pairs are correctly ordered and a value of 0 being the
 alternative with no pairs correctly ordered. Any values between 0 and 1
@@
 
-### Survival Prediction
+### Survival Probability
 
@@
 significant individually with all the data. As such, to determine the
-best model to use for this analysis, a forward and backward step-wise
+best model to use for this analysis, a forward and backward stepwise
 regression will be done to generate the models with the best Akaike
@@
 The table below provides the model number, the covariates used, the AIC
-and BIC from the step-wise regression if applicable, and the concordance
-index. The concordance index generates a value to determine which model
-is the most accurate with the provided covariates. After determining the
-most accurate model, the CPH assumptions will be tested.
+and BIC from the stepwise regression if applicable, and the concordance
+index. After determining the most accurate model, the CPH assumptions
+will be tested.
 
@@
 bic <- step(all_cov, direction = "both", k = log(n), trace = 0)
-
-train_sig <- lapply( univ_formulas, function(x){coxph(x, data = train_grouped)})
 

### 2025-04-15 slides.qmd
Update 4/15 pt2

@@
 library(broom)
-```
-
-```{r, echo = FALSE}
-#generation of the model for sample plots
-example_surv_object <- Surv(time = lung$time, event = lung$status)
-
-example_coxmodel <- coxph(example_surv_object ~ meal.cal, data = lung)
 ```
@@
 
-## Linearity Assumption
+## Visualization
 
--   Tested using the Martingale residuals for each covariate using the
-    equation Martingale Residuals = Observed Events - Expected Results
-    [@nahhas2025]
+The two common model visualizations are:
 
--   If the plots are linear and appear to have a slope of zero, the
-    assumption is not violated [@Amini2015]
+::: incremental
+-   Kaplan-Meier Survival Curve
+
+    -   Demonstrates the time where the event being modeled is expected
+        to occur [@klein2005]
+
+-   Forest Plots
+
+    -   Visualizes the effects of each covariate in the model on the
+        hazard ratio
+:::
+
+## Evaluation and Survival Probability
+
+::: incremental
+-   Model accuracy evaluation is done using the concordance index
+
+    ::: incremental
+    -   The concordance index measures the amount of agreement between
+        two variables
+
+    -   A value of 1 means all the pairs are correctly ordered while a
+        value of 0 means no pairs are correctly ordered
+    :::
+
+-   Survival probability can be predicted at a specific time $t$
+
+    ::: incremental
+    -   If the probability is $\geq 50$%, it is assumed the event has
+        not occurred
+
+    -   If the probability is $< 50$%, it is assumed the event has
+        occurred
+    :::
+:::
+
+## Data Structure
+
+The data selected for this project comes from a study on propagation
+modeling that NASA completed, specifically focusing on the engine two
+testing and training datasets [@saxena2008].
+
+-   Each engine in the NASA data has an unknown amount of wear,
+    manufacturing variation, and sensor noise
+
+-   There are three operation setting fields and twenty-one sensor
+    measurement fields
+
+-   A column indicating status was added to both the testing and
+    training datasets with 0 indicating the machine has not failed and 1
+    indicating the machine has failed
+
+-   The data was then merged together to make one grouped dataset
+
+```{r}
+testurl <- 'https://raw.githubusercontent.com/JR-87/capstone4ds_reed/refs/heads/main/test_FD002.txt'
+
+test <- read.delim(testurl, header = TRUE, sep = '') %>%
+  mutate(status = 0,
+         id = id + 260)
+
+trainurl <- 'https://raw.githubusercontent.com/JR-87/capstone4ds_reed/refs/heads/main/train_FD002.txt'
+
+train <- read.delim(trainurl, header = TRUE, sep = '') %>%
+  group_by(id) %>%
+  mutate(status = if_else(row_number() == n(), 1, 0)) %>%
+  ungroup()
+
+data_all <- rbind(train, test)
+
+train_grouped <- train %>% 
+  group_by(id) %>%
+  slice_max(time)
+
+data_grouped <- data_all %>%
+  group_by(id) %>%
+  slice_max(time)
+
+temp <- data_grouped %>% select(id, time, status, os1, os2, os3, sm1, sm2, sm3)
+
+kable(head(temp, n = 5), format = "markdown", caption = 'NASA Aircraft Engine Data')
+```
+
+## Exploration
+
+-   519 engines in the combined data
+
+    -   260 engines in training data
+
+    -   259 engines in testing data
+
+```{r}
+summarymetrics <- data.frame(c("Minimum", "Median", "Mean", "Standard Deviation", "Maximum"),
+                             c(min(train_grouped$time), 
+                               median(train_grouped$time),
+                               mean(train_grouped$time),
+                               sd(train_grouped$time),
+                               max(train_grouped$time)))
+
+colnames(summarymetrics) <- c("Metric", "Value")
+summarymetrics$Value <- round(summarymetrics$Value, digits = 2)
+kable(summarymetrics, format = "markdown", align = "c", caption = "Summary Metrics")
+```
+
+## Visualization
+
+::: panel-tabset
+#### Histogram/Boxplot
 
@@
 #| fig-align: center
-example_martingale <- resid(example_coxmodel, type = "martingale")
+Histogram <- ggplot(data = train_grouped,
+       aes(x = time)) +
+  geom_histogram(binwidth = 5,
+                 fill = "grey")+
+  labs(x = "Survival Time (Iterations)",
+       y = "Frequency")+
+  theme_bw()
 
-plot(example_martingale, 
-     main = "Martingale Residuals: Example Covariate",
-     ylab = "Martingale Residuals for Example Covariate",
-     xlab = "Value Range for Example Covariate")
-abline(h = 0, col = "red")
+box <- ggplot(data = train_grouped,
+                  aes(y = time))+
+  geom_boxplot(fill = "gray",
+               color = "black")+
+  ylab("Time (Iterations)")+
+  theme_bw()
 
+grid.arrange(Histogram,box,ncol = 2, top = "Distribution of Engine Longevity in Training Data")
 ```
 
-## Proportional Hazards Assumption
-
--   The test is verified using the Schoenfeld partial residuals which
-    are the difference between the value of the covariate and the
-    expected value of the covariate at the time of failure [@klein2005]
-
-::: panel-tabset
-#### Graph
+#### Survival Curve
 
 ```{r}
-example_phtest <- cox.zph(example_coxmodel)
+#| fig-align: center
+t <- as.matrix(train_grouped[, "time"])
+s <- as.matrix(train_grouped[, "status"])
 
-plot(example_phtest[1],
-     ylab = "Covariate Schoenfeld Residual",
-     xlab = "Time")
-```
+km_fit <- survfit(Surv(t, s) ~ 1)
 
-#### Table
-
-```{r}
-multi <- coxph(example_surv_object ~ meal.cal + age, data = lung)
-ph_tibble <- cox.zph(multi) %>%
-  pluck("table") %>%
-  as.data.frame() %>%
-  tibble::rownames_to_column("variable") %>%
-  as_tibble()
-kable(ph_tibble, format = "markdown")
+autoplot(km_fit) + theme_bw()
 ```
@@
 
-## Kaplan-Meier Survival Curve
+## Creating CPH Model
 
-::: incremental
--   Provides visualization for the Kaplan-Meier estimator
+The table provides the model number, the covariates used, the AIC and
+BIC from the stepwise regression if applicable, and the concordance
+index.
 
--   Is considered to be well defined up until the largest observed study
-    time $t_{max}$
+```{r}
+surv_object <- Surv(data_grouped$time, data_grouped$status)
 
--   Demonstrates the time where the event being modeled is expected to
-    occur
-:::
+all_cov <- coxph(data = data_grouped, surv_object ~ os1 + os2 + os3 + sm1 + sm2 + sm3 + sm4 + sm5 + sm6 + sm7 + sm8 + sm9 + sm10 + sm11 + sm12 + sm13 + sm14 + sm15 + sm16 + sm17 + sm18 + sm19 + sm20 + sm21)
+
+aic <- MASS::stepAIC(all_cov, trace = 0)
+n <- nrow(data_grouped)
+bic <- step(all_cov, direction = "both", k = log(n), trace = 0)
+
+cox1 <- coxph(data = data_grouped, surv_object ~ os1 + os2 + os3 + sm1 + sm2 + sm3 + sm4 + sm5 + sm6 + sm7 + sm8 + sm9 + sm10 + sm11 + sm12 + sm13 + sm14 + sm15 + sm16 + sm17 + sm18 + sm19 + sm20 + sm21)
+
+cox2 <- coxph(data = data_grouped, surv_object ~ os3 + sm3 + sm4 + sm5 + sm8 + sm9 + sm15 + sm16 + sm17 + sm18 + sm20)
+
+cox3 <- coxph(data = data_grouped, surv_object ~ os3 + sm3 + sm4 + sm8 + sm9 + sm15 + sm18)
+
+cox4 <- coxph(data = data_grouped, surv_object ~ os3 + sm13 + sm14 + sm19)
+
+stepwise_results <- data.frame(c("model 1", "model 2", "model 3", "model 4"), 
+                               c("All covariates", "os3, sm3 - sm5, sm8 - sm9, sm15 - sm18, sm20", "os3, sm3 - sm4, sm8-sm9, sm15, sm18", "os3, sm13 - sm14, sm19"),
+                               c("2449.92", "2431.88", "N/A", "N/A"),
+                               c("2547.71", "N/A", "2461.31", "N/A"),
+                               c(cox1$concordance[6], cox2$concordance[6], cox3$concordance[6], cox4$concordance[6]))
+
+colnames(stepwise_results) <- c("Model", "Covariates", "AIC", "BIC", "Concordance")
+
+kable(stepwise_results, format = "markdown")
+```
+
+Model for the continuing analysis will have all covariates except sm16
+and sm19.
+
+## CPH Model Forest Plot
 
@@
 #| fig-align: center
-example_curve <- survfit(example_coxmodel)
+cox1.1 <- coxph(data = data_grouped, surv_object ~ os1 + os2 + os3 + sm1 + sm2 + sm3 + sm4 + sm5 + sm6 + sm7 + sm8 + sm9 + sm10 + sm11 + sm12 + sm13 + sm14 + sm15 + sm17 + sm18 + sm20 + sm21)
 
-ggsurvplot(example_curve,
-           data = lung,
-           title = "Example Survival Curve",
-           xlab = "Time",
-           ylab = "Survival Probability",
-           legend = "none")
-```
+ci <- cox1.1$concordance[6]
 
-## Forest Plot
-
--   Provides visualization of the effects of each covariate on the
-    hazard ratio
-
--   A positive effect indicates a positive correlation with the hazard
-    ratio
-
-```{r, warning=FALSE}
-#| fig-align: center
-forest <- coxph(example_surv_object ~ age + sex, data = lung)
-
-plot_model(forest,
-           dot.size = 1,
+plot_model(cox1.1,
+           dot.size = 2,
            line.size = 1,
@@
   xlab(label = "Estimated Effect on Hazard Ratio")
-
 ```
 
-## Evaluation
-
-## Prediction
-
-## Data Structure, Exploration, and Visualization
-
-## Creating CPH Model
-
 ## Checking Assumptions
-
-## Adjusting the Model
 

### 2025-04-16 index.qmd
Update 4/15 pt3

@@
 CPH to use to predict the survival probability. If any of the
-assumptions for the model are not true, the model will need to be
+assumptions for the model are not met, the model will need to be
 modified.
@@
 
-The survival times for each individual engine in the training data are
+The survival times for each individual engine in the data are
 independent of each other. The failure of one engine does not impact the
@@
 
-The training dataset does not contain censored data as the documentation
-available at the dataset’s source clarifies all engines in the training
-data have experienced the desired event of failure at the end of each
-time series. As such, this assumption is met.
+While the data that was pulled from the training set does not include
+censored data, all of the data from the testing dataset is censored as
+those engines had not experienced failure in the study time period.
+However, it is known that those engines will experience failure at some
+point and knowing those engines are censored does not provide any new
+information. As such, this assumption is met.
 
@@
 are at a greater risk of experiencing failure. The below table provides
-the hazard rate and suvivial probability for the first 10 engines in the
+the hazard rate and survival probability for the first 10 engines in the
 dataset.
@@
 before negative impacts occur. While CPH does not allow for the
-prediction of specific failure times, it does allow for the creation
+prediction of specific failure times, it does allow for the creation of
 confidence intervals which can be used to gain an understanding of when

### 2025-04-16 slides.qmd
Update 4/15 pt3

@@
         constant at all times [@Bustan2018]
-:::
-
-## Visualization
-
-The two common model visualizations are:
-
-::: incremental
--   Kaplan-Meier Survival Curve
-
-    -   Demonstrates the time where the event being modeled is expected
-        to occur [@klein2005]
-
--   Forest Plots
-
-    -   Visualizes the effects of each covariate in the model on the
-        hazard ratio
 :::
@@
 
-## Checking Assumptions
+## Checking Assumptions {.scrollable}
 
-## Model Results
+::: panel-tabset
+#### Independence
+
+-   This assumption is met as the survival times for each individual
+    engine in the data is independent of each other
+
+#### Non-informative Censoring
+
+-   This assumption is met as knowing there is censored data does not
+    provide any new information about the survival probability
+
+#### Linearity Assumption
+
+-   This assumption is tested by looking at the Martingale residuals
+    (Martingale Residuals = Observed Events - Expected Events) and
+    determining if there is a slope of zero
+
+-   Based on the below plots, this assumption is met.
+
+```{r}
+martingale <- resid(cox1.1, type = "martingale")
+
+par(mfrow = c(4,4),
+      mar = c(2,2,4,1))
+
+plot(data_grouped$os1, martingale, main = "Martingale Residuals: os1")
+abline(h = 0, col = "red")
+
+plot(data_grouped$os2, martingale, main = "Martingale Residuals: os2")
+abline(h = 0, col = "red")
+
+plot(data_grouped$os3, martingale, main = "Martingale Residuals: os3")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm1, martingale, main = "Martingale Residuals: sm1")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm2, martingale, main = "Martingale Residuals: sm2")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm3, martingale, main = "Martingale Residuals: sm3")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm4, martingale, main = "Martingale Residuals: sm4")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm5, martingale, main = "Martingale Residuals: sm5")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm6, martingale, main = "Martingale Residuals: sm6")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm7, martingale, main = "Martingale Residuals: sm7")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm8, martingale, main = "Martingale Residuals: sm8")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm9, martingale, main = "Martingale Residuals: sm9")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm10, martingale, main = "Martingale Residuals: sm10")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm11, martingale, main = "Martingale Residuals: sm11")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm12, martingale, main = "Martingale Residuals: sm12")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm13, martingale, main = "Martingale Residuals: sm13")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm14, martingale, main = "Martingale Residuals: sm14")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm15, martingale, main = "Martingale Residuals: sm15")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm17, martingale, main = "Martingale Residuals: sm17")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm18, martingale, main = "Martingale Residuals: sm18")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm20, martingale, main = "Martingale Residuals: sm20")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm21, martingale, main = "Martingale Residuals: sm21")
+abline(h = 0, col = "red")
+```
+
+#### Proportional Hazards
+
+-   This assumption is tested by looking at the scaled Schoenfeld
+    partial residuals (the value of the covariate - the expected value
+    of the covariate at the time of failure) and examining the p-value
+
+-   Based on the below table and plots, this assumption is met
+
+```{r}
+cph_ph <- cox.zph(cox1.1) %>%
+  pluck("table") %>%
+  as.data.frame() %>%
+  tibble::rownames_to_column("variable") %>%
+  as_tibble()
+temp <- cph_ph %>% filter(variable %in% c("os1", "os2", "sm1", "sm2", "GLOBAL"))
+kable(temp, format = "markdown")
+
+create_ph_plots <- function(coxmodel, row, col){
+  # Function takes argument that is a cox model from coxph()
+
+  phresults <- cox.zph(coxmodel)
+
+  # Define dimensions of visual
+  par(mfrow = c(row, col),
+      mar = c(2,2,4,1))
+
+  # Loop through each covariate creating a plot and adding it to the phplot list
+  for (i in 1:(length(phresults$table[,1])-1)){
+    phplots <- list()
+    phplots[i] <- plot(phresults[i], # Create plot for each iteration
+                       main = rownames(phresults$table)[i]) # Title of each plot is the rowname from coxmodel$table
+
+  }
+  mtext("Schoenfeld Residuals vs. Time for Covariates", side = 3, line = -1.5, outer = TRUE, cex = 1) # Title for Visual
+  par(mfrow = c(1,1)) # Reset plotting area
+}
+
+create_ph_plots(cox1.1, 3, 5)
+
+```
+:::
+
+## Model Results {.scrollable}
+
+::: panel-tabset
+#### Evaluation
+
+-   The CPH model was determined to be a valid model and has a
+    concordance index of `{r} ci`
+
+#### Calculations
+
+-   Using the model, we can generate the coefficients and calculate the
+    hazard rate, the cumulative hazard value, and the survival
+    probability
+
+```{r}
+#| fig-align: center
+data_grouped_reduced <- data_grouped %>% select(id, time, os1, os2, os3, sm1, sm2, sm3, sm4, sm5, sm6, sm7, sm8, sm9, sm10, sm11, sm12, sm13, sm14, sm15, sm17, sm18, sm20, sm21, status)
+
+data_grouped_reduced$HazardRate <- predict(cox1.1,
+        newdata = data_grouped_reduced,
+        type = "risk")
+
+data_grouped_reduced$SurvivalProbability <- predict(cox1.1, type = "survival", newdata = data_grouped_reduced)
+
+print_data <- data_grouped_reduced %>% select(id, time, status, HazardRate, SurvivalProbability)
+
+kable(head(print_data, n = 5), format = "markdown")
+
+sf <- survfit(cox1.1)
+
+ggsurvplot(fit = sf,
+           data = data_grouped_reduced, risk.table = TRUE, fun = "cumhaz")$plot +
+  ggtitle('Cumulative Hazard Over Time')
+
+survtimes <- data.frame(Percent = c("100% Survival", "75% Survival,", "50% Survival", "25% Survival", "10% Survival"),
+                        Time = c(sf$time[sf$lower < 1][1], 
+                                    sf$time[sf$lower < .75][1], 
+                                    sf$time[sf$lower < .5][1], 
+                                    sf$time[sf$lower < .25][1], 
+                                    sf$time[sf$lower < .1][1]))
+
+ggsurvplot(fit = sf, data = data_grouped_reduced)$plot +
+  ggtitle('Survival Probability Over Time')
+
+```
+:::
 
@@
 
+-   Based on the model, NASA should not use their engines after 128
+    iterations have passed
+
+```{r}
+kable(survtimes, format = "markdown")
+```
+
 ## References

### 2025-04-17 index.qmd
Update 4/17

@@
 same methods can be applied to predicting bank failure, the survival
-probablity of machines, machinery maintenance timelines, and insurance
+probability of machines, machinery maintenance timelines, and insurance
 likelihood of payouts. There are various survival methods that can be
-examined, including Cox proportional hazards model which will be
+examined, including the Cox proportional hazards model, which will be
 examined in this paper.
@@
 as the length of the study [@klein2005]. The models also have the
-advantage of being semi-parametric which allows more flexibility as
+advantage of being semi-parametric, which allows more flexibility as
 compared to other models [@Ming-Chiang2014].
@@
 COVID-19 study done using data from a Pakistani hospital noted that the
-model was limited due to the recovery possibility of COVID-19 patients
+model was limited due to the recovery possibility of COVID-19 patients,
 as CPH assumes as time goes on, the survival probability will approach
@@
 between chest pain occurring and mortality of the person experiencing
-chest pain using CPH. A response to that study was written that provided
-the limitations of CPH with the biggest concern being the assumption of
-constant hazards ratio that CPH relies on for the model. In the heart
-study, the effect of Triglyceride-Glucose was assumed to be a constant,
-but is actually a value that fluctuates over time along with an
-individual's body weight and blood glucose levels. As such, the model
-would struggle to predict correctly the effect of Triglyceride-Glucose
-given the violation of the constant hazards ratio [@jiang2024].
+chest pain using CPH. A response to that study demonstrated the various
+limitations of CPH, with the primary concern being the constant hazards
+ratio assumption that CPH relies on. In the heart study, the effect of
+triglyceride-glucose was assumed to be a constant, but it is actually a
+value that fluctuates over time along with an individual's body weight
+and blood glucose levels. As such, the model would struggle to predict
+correctly the effect of triglyceride-glucose given the violation of the
+constant hazards ratio [@jiang2024].
 
@@
 health data demonstrated how bias can impact the values for the
-covariates when relying on patient submitted information. The first was
+covariates when relying on patient-submitted information. The first was
 a study on the connection between salt intake and the development of
-anxiety or depression after 14.5 years [@wang2025] and the second was a
+anxiety or depression after 14.5 years [@wang2025], and the second was a
 study on the association between postpartum depression and the
 development of two or more chronic diseases [@zhang2025]. Both of those
-studies relied on patient submitted health data and survey responses
-that were supplemented with death records and other accessible health
-records thus creating a possible bias in their data which could
-negatively influence their model as their covariates may not have
-accurately reflected the true data. These articles provide examples of
-the care that needs to be taken when working with CPH models.
+studies relied on patient-submitted health data and survey responses
+supplemented with death records and other accessible health records.
+This created a possible bias in their data that may have negatively
+influenced their model as their collected data may not accurately
+reflect the true data. These articles provide examples of the care that
+needs to be taken when working with CPH models.
 
@@
 company in the Czech Republic provides policies for certain illness
-events until the event occurs when the insured will receive money and
-their contract is complete. To assist the company with predicting if an
+events, with the insured receiving money when the event occurs, which
+concludes the contract. To assist the company with predicting if an
 individual will have one of the illness events occur, a model using CPH
@@
 life of female Floridian goats to determine the factors that would
-impact the longevity of the goats. CPH was used to determine that the
-age of first kidding and the specific herd impacts the length of
-productive life of the goats [@ziadi2023]. Both of these studies provide
-examples of the reaches of CPH models outside the health industry.
-
-CPH has the capability to be applied to many fields, though it poses
-some limitations. The constant hazards ratio and determining how to
-censor data are two concerns that can lead to limitations in using CPH
-for models. In addition, the necessity of a time to event value within
-the data source is something that needs to be taken into consideration
-when choosing to use CPH. Without that value, using CPH as a survival
-model will not be the appropriate method to take with that particular
-data. This value is also what makes CPH a prime modeling candidate with
-health data as health metrics frequently have a time component to them,
-especially when comparing someone over a period of time in a health
-study as is done with medications and the development of cancer or other
+impact the longevity of the goats. CPH was used to determine that both
+the age at first birth and the specific herd the female goat belonged to
+impacts the length of productive life [@ziadi2023]. Both of these
+studies provide examples of the reaches of CPH models outside the health
+industry.
+
+While CPH can be applied across various fields for survival probability
+modeling, the method does have elements that may limit its accuracy that
+must be addressed. Two such elements are the constant hazards ratio and
+the censoring of data, as both can lead to limited accuracy. In
+addition, the necessity of a time-to-event value within the data source
+is something that needs to be considered when choosing to use CPH.
+Without that value, using CPH as a survival model will not be the
+appropriate method to take with that particular data. This value is also
+what makes CPH a prime modeling candidate with health data, as health
+metrics frequently have a time component to them. This is frequently
+seen when comparing someone over a period of time in a health study, as
+is done with medications and the development of cancer or other
 illnesses. However, while the usage of CPH is clear in the medical
 field, this paper will demonstrate using CPH within the field of
-mechanics and remaining useful life of a machine.
+mechanics and survival probability metrics of a machine.
 
@@
 
-CPH then uses both functions to develop it's own model, specifically by
+CPH then uses both functions to develop its own model, specifically by
 adapting the hazard function. Since CPH is a proportional hazard model,
@@
 $Z_k = z+1$ and for an individual with $\mathbf{Z}^*$ covariates where
-$Z_k = z$ with all other predictors are fixed [@nahhas2025] is defined
-as
+$Z_k = z$ with all other predictors fixed [@nahhas2025] is defined as
 
@@
 Depending on the study, there will be covariates that do not impact the
-probability of survival. Statistical methods such as the Wald Test can
+probability of survival. Statistical methods such as the Wald test can
 be used to determine which ones are statistically significant
@@
 of each other. This is similar to most regression models, though CPH
-does not assume that the residuals of the model be normally distributed
+does not assume that the residuals of the model are normally distributed
 or have constant variance [@nahhas2025].
@@
 
-CPH assumes that censoring is non-informative which means that each
+CPH assumes that censoring is non-informative, which means that each
 subject has the same risk of experiencing the event regardless of one of
@@
 visually using the Martingale residuals for each covariate
-[@nahhas2025]. Martingale residuals are a type of residuals used in
-survival analysis that show the discrepancy between the observed and the
-expected number of events. The equation is $$
+[@nahhas2025]. Martingale residuals are a type of residual used in
+survival analysis that shows the discrepancy between the observed and
+the expected number of events. The equation is $$
 \begin{align*}
@@
 
-The resulting residuals are examined visually and if they are linear and
-appear to have a slope of zero, the linearity assumption is not violated
-[@Amini2015]. If a model is being generated with multiple covariates,
-multiple Martingale residual plots will need to be generated in order to
-verify this assumption for each covariate.
+The resulting residuals are examined visually, and if they are linear
+and appear to have a slope of zero, the linearity assumption is not
+violated [@Amini2015]. If a model is being generated with multiple
+covariates, multiple Martingale residual plots will need to be generated
+in order to verify this assumption for each covariate.
 
@@
 residuals are linear and appear to have a slope of zero. For this
-covariate then, the linearity assumption has not been violated.
+covariate, the linearity assumption has not been violated.
 
@@
 
-The major assumption for CPH is the proportional hazards (PH) assumption
-which states that the ratio of hazard rates for any two subjects must be
-constant at all times. This means the effect of the covariates must be
-constant over time [@Bustan2018]. To test this assumption, the
-hypothesis test below is used: $$
+The major assumption for CPH is the proportional hazards (PH)
+assumption, which states that the ratio of hazard rates for any two
+subjects must be constant at all times. This means the effect of the
+covariates must be constant over time [@Bustan2018]. To test this
+assumption, the hypothesis test below is used: $$
 \begin{align*}
@@
 and the expected value of the covariate at the time of failure
-[@klein2005]. If the global p-value is $< 0.05$, than the model has at
-least one covariate that violates the PH assumption and each individual
+[@klein2005]. If the global p-value is $< 0.05$, then the model has at
+least one covariate that violates the PH assumption, and each individual
 covariate p-value will need to be checked. If the global p-value is
@@
 Violations of the PH assumption can be dealt with in several ways. If
-the covariate is categorical, one way is to stratify the covariate which
-will generate a separate baseline hazard function for each level of the
-covariate [@harrison2021]. This removes the PH assumption for that
-covariate, though it will still be assumed for all other covariates in
-each level of the stratification [@nahhas2025]. If the covariate is
+the covariate is categorical, one way is to stratify the covariate,
+which will generate a separate baseline hazard function for each level
+of the covariate [@harrison2021]. This removes the PH assumption for
+that covariate, though it will still be assumed for all other covariates
+in each level of the stratification [@nahhas2025]. If the covariate is
 continuous, one way is to add in a function of time based on the shape
@@
 assumption. While there is some curvature in the line, the residuals are
-roughly horizontal indicating proportional hazard over time for this
+roughly horizontal, indicating proportional hazard over time for this
 covariate.
@@
 
-Two common model visualizations used for CPH is the forest plot and the
+Two common model visualizations used for CPH are the forest plot and the
 Kaplan-Meier survival curve. The Kaplan-Meier curve assists in
-visualizing the initial survival model for the data while the forest
+visualizing the initial survival model for the data, while the forest
 plot assists in visualizing the hazard ratio after generating a model.
@@
 
-KM is a non-parametric estimate which means it makes no assumption of
+KM is a non-parametric estimate, which means it makes no assumption of
 the shape of the base survival function [@nahhas2025] and is considered
-to be well defined up until the largest observed study time $t_{max}$
+to be well defined up until the largest observed study time, $t_{max}$
 [@klein2005]. The survival curve provides a visualization of the KM
-estimator and are important when modeling survival data because it
-demonstrates the time where the event being modeled is expected to
-occur.
+estimator and is important when modeling survival data because it
+demonstrates the time when the event being modeled is expected to occur.
 
 Below is an example survival curve showing survival probabilities over
-time. The line represents the median survival probability with the red
+time. The line represents the median survival probability, with the red
 area around it showing the 95% confidence interval for the values. The
-curve can be interpreted as showing the probability of event would be
-expected to happen around a time of 300 where the survival probability
-is below 50% [@Kuitunen2021].
+curve can be interpreted as showing the probability of an event that
+would be expected to happen around a time of 300, where the survival
+probability is below 50% [@Kuitunen2021].
 
@@
 
-Forest plots can be used to visualize the effects of each covariate in
-the model has on the hazard ratio. The resulting plot contains each
-covariate on the y axis, with the estimated effect on the hazard ratio
-on the the x axis. The plot contains whiskers demonstrating a 95%
+Forest plots can be used to visualize the effects each covariate in the
+model has on the hazard ratio. The resulting plot contains each
+covariate on the y-axis, with the estimated effect on the hazard ratio
+on the the x-axis. The plot contains whiskers demonstrating a 95%
 confidence interval for the effect. A positive effect indicates a
-positive correlation with the hazard ratio where the higher the number
-is the more severe the effect on the hazard ratio.
-
-An example forest plot is shown below plotting the hazard ratio for each
-covariate in the model. The point represents the calculated hazard ratio
-and the whiskers show the 95% confidence interval.
+positive correlation with the hazard ratio, where the higher the number
+is, the more severe the effect on the hazard ratio.
+
+An example forest plot is shown below, plotting the hazard ratio for
+each covariate in the model. The point represents the calculated hazard
+ratio, and the whiskers show the 95% confidence interval.
 
@@
 Evaluation of the accuracy of the CPH model is done using the
-concordance index which is used to measure the amount of agreement
+concordance index, which is used to measure the amount of agreement
 between two variables. Specifically, the value looks at concordant pairs
-and disconcordant pairs. A concordant pair is when either $x_i < x_j$
-and $y_i < y_j$ or $x_i > x_j$ and $y_i > y_j$. A discordant pair is
-when either $x_i$ or $y_i$ are not in the same place such as $x_i < x_j$
-but $y_j < y_i$ [@Therneau2017]. In terms of survival, this can be
-thought of as examining when the failure event occurs. If two subjects,
-A and B, are compared with subject A having a higher risk, the pair
-would be considered concordant if subject A does have the failure event
-occur before subject B.
+and discordant pairs. A concordant pair is when either $x_i < x_j$ and
+$y_i < y_j$ or $x_i > x_j$ and $y_i > y_j$. A discordant pair is when
+either $x_i$ or $y_i$ are not in the same place, such as $x_i < x_j$ but
+$y_j < y_i$ [@Therneau2017]. In terms of survival, this can be thought
+of as examining when the failure event occurs. If two subjects, A and B,
+are compared, with subject A having a higher risk, the pair would be
+considered concordant if subject A does have the failure event occur
+before subject B.
 
@@
 The concordance index will produce a value between 0 and 1. A value of 1
-means all pairs are correctly ordered and a value of 0 being the
-alternative with no pairs correctly ordered. Any values between 0 and 1
-indicate how accurate the CPH model is with the data [@Therneau2017].
+means all pairs are correctly ordered, and a value of 0 means none of
+the pairs are correctly ordered. Any values between 0 and 1 indicate how
+accurate the CPH model is with the data [@Therneau2017].
 
@@
 that are desired. However, CPH is not able to predict the specific time
-the event is going to occur which is considered a limitation of the
+the event is going to occur, which is considered a limitation of the
 model.
@@
 propagation modeling [@saxena2008]. In particular, this paper will be
-examining the data from the engine design two which has a training
+examining the data from the engine design two, which has a training
 dataset and a testing dataset.
@@
 Each engine in the NASA data has an unknown amount of wear,
-manufacturing variation, and sensor noise which will impact the survival
-time. In the training dataset, each engine is operating normally at the
-beginning of the time series, will develop a fault, and experience
-failure at some point. All engines will have failed in the training
-dataset. The conditions the engines are subjected to are represented by
-three operational setting fields and twenty-one sensor measurement
-fields. In the testing dataset, none of the engines have experienced
-failure [@saxena2008].
+manufacturing variation, and sensor noise, which will impact the
+survival time. In the training dataset, each engine is operating
+normally at the beginning of the time series, will develop a fault, and
+experience failure at some point. All engines will have failed in the
+training dataset. The conditions the engines are subjected to are
+represented by three operational setting fields and twenty-one sensor
+measurement fields. In the testing dataset, none of the engines have
+experienced failure [@saxena2008].
 
@@
 engines in that dataset will eventually fail. The testing dataset will
-only have the value of 0 as none of the machines have failed in that
+only have the value of 0, as none of the machines have failed in that
 dataset.
@@
 
-There are 519 engines in the combined data with 260 engines in the
+There are 519 engines in the combined data, with 260 engines in the
 training data and 259 engines in the testing data. The below metrics are
-generated using the training data as that data contains no censored
+generated using the training data, as that data contains no censored
 data. The histogram shows that the survival times of the engines are
-normally distributed and the box plot shows there six outliers in the
-dataset. The engines have a median survival time of 199 cycles with a
-standard deviation of 46.78. The maximum number of cycles an engine
-lasted was 378 while the minimum number of cycles an engine lasted was
+normally distributed, and the box plot shows there are six outliers in
+the dataset. The engines have a median survival time of 199 cycles with
+a standard deviation of 46.78. The maximum number of cycles an engine
+lasted was 378, while the minimum number of cycles an engine lasted was
 128. Since the testing data contains all censored data, the basic
@@
 survival data, a survival object is created by running
-`Surv(time, event)` where `time` is the time field and `event` is the
+`Surv(time, event)`, where `time` is the time field and `event` is the
 event field. To generate the actual model, the survival object is then
-applied to the function `coxph(survival object ~ x, data)` where `x`
+applied to the function `coxph(survival object ~ x, data)`, where `x`
 represents the covariate(s) and `data` is the survival data that is
@@
 Information Criterion (AIC) and Bayesian Information Criterion (BIC)
-values and compare the concordance values of those models with models
-that have covariates that were statistically significant when used in a
-CPH model with the training data. The model with the highest concordance
-value, and therefore, the most accurate will be used for the analysis.
+values. This method produced three possible models. The fourth model in
+the below table used covariates that were determined to significant with
+the training data only.
 
@@
 
-Based on the above model results, the model that has the highest
-concordance value is model 1 which has all of the covariates. However,
-before proceeding with that model, a test for linear dependency will be
-run to determine all covariates are linearly independent.
-
-```{r}
-kable(alias(cox1), format = "markdown")
-```
-
-The above result indicates that sm19 needs to be removed from the model
-as it linearly dependent on another value. In addition, sm16 resulted in
-a possible infinite coefficient so that covariate has also been removed.
-The proceeding analysis will be done using model 1 without sm16 and
-sm19. The below tables provide the summary values from the model.
+In determining the best model for this data, it was determined that CPH
+is not the survival method that should be used on this data. By stepwise
+regression, the models with the lowest BIC and AIC should have been used
+in the following analysis; however, both models continuously failed the
+CPH assumptions. As such, the model for the following analysis will be
+model 1, which contained all covariates and had the highest concordance
+value.
+
+While model 1 does have initially contain all 24 covariates, there are
+two that will need to be removed before proceeding with the analysis due
+to infinite coefficients and linearly dependence. The covariate sm19 is
+linearly dependent on another value, and the covariate sm16 resulted in
+a possible infinite coefficient. The following analysis will be done
+using model 1 without sm16 and sm19. The below tables provide the
+summary values from the model.
 
@@
 independent of each other. The failure of one engine does not impact the
-failure of another so this assumption is met.
+failure of another, so this assumption is met.
 
@@
 While the data that was pulled from the training set does not include
-censored data, all of the data from the testing dataset is censored as
+censored data, all of the data from the testing dataset is censored, as
 those engines had not experienced failure in the study time period.
 However, it is known that those engines will experience failure at some
-point and knowing those engines are censored does not provide any new
+point, and knowing those engines are censored does not provide any new
 information. As such, this assumption is met.
@@
 
-To check the linearity assumption, the residuals of the model are needed
-which can be generated using the `resid` function in R and assigning the
-type to be martingale (`resid(type = "martingale")` ). This will create
-a vector of residuals that can be plotted against the range of values
-for each covariate.
-
-The plots of the Martingale residuals vs the range of values for each
-covariate is shown below. For each covariate, there is no clear slope
-indicating the linearity assumption is true for the covariates in the
-model.
+To check the linearity assumption, the residuals of the model are
+needed, which can be generated using the `resid` function in R and
+assigning the type to be martingale (`resid(type = "martingale")` ).
+This will create a vector of residuals that can be plotted against the
+range of values for each covariate.
+
+The plots of the Martingale residuals versus the range of values for
+each covariate are shown below. For each covariate, there is no clear
+slope indicating the linearity assumption is true for the covariates in
+the model.
 
@@
 The `cox.zph` function can be used to check the proportional hazards
-assumption. As mentioned, the function calculates p values for each
+assumption. As mentioned, the function calculates p-values for each
 covariate. The global p-value will first be compared to $\alpha = 0.05$.
 If that p-value is $<\alpha$, then at least one covariate violates the
-PH assumption and the model will need to be adapted.
+PH assumption, and the model will need to be adapted.
 
@@
 Schoenfeld residuals does show that os1, os3, sm13, and sm14 violate the
-PH assumption. To ensure the model is as accurate as possible, the model
-will addapted without those covariates and the global p-values will be
-compared.
+PH assumption. However, since the global p-value does show that the
+model itself meets the PH assumption, the analysis will proceed with
+those covariates. If the global p-value had not met the assumption,
+further analysis of those four covariates would need to be completed in
+order to met the PH assumption.
 
@@
 
-### Adjusting the Model
-
-The results of the above assumptions demonstrated that several
-covariates violate the PH assumption on a local level. As such, the
-model will be recreated by removing the covariates that violate the PH
-assumption. After recreating the model, the PH assumption and the
-linearity assumption will be tested again with the new PH global p-value
-being compared to the original model's global p-value.
-
-The table below demonstrates the new model information which has
-eighteen covariates: os2, sm1 - sm12, sm15, sm17, sm18, sm20, sm21.
-
-```{r}
-cox1.2 <- coxph(data = data_grouped, surv_object ~ os2 + sm1 + sm2 + sm3 + sm4 + sm5 + sm6 + sm7 + sm8 + sm9 + sm10 + sm11 + sm12 + sm15 + sm17 + sm18 + sm20 + sm21)
-
-kable(tidy(cox1.2), format = "markdown")
-g <- glance(cox1.2)
-g <- g %>% select(n, nevent, concordance, std.error.concordance, logLik, statistic.log, p.value.log, statistic.wald, p.value.wald, statistic.sc, p.value.sc)
-kable(g, format = "markdown")
-
-```
-
-The output below shows the p-values for the new model's PH assumption
-hypothesis test and Schoenfeld residuals. The global p-value is now
-$< \alpha$ so the PH assumption has not been met for the new model. As
-such,, the remaining analysis will be done using the first model that
-had a global p-value $>\alpha$ which has all but two covariates.
-
-```{r, echo = FALSE}
-ph_2 <- cox.zph(cox1.2) %>%
-  pluck("table") %>%
-  as.data.frame() %>%
-  tibble::rownames_to_column("variable") %>%
-  as_tibble()
-
-kable(ph_2, format = "markdown")
-
-create_ph_plots(cox1.2, 3, 5)
-
-```
-
-If the new model had passed the PH assumption, the linearity assumption
-would have been checked again.
-
 ### Model Results
@@
 
-The CPH model that was selected has a concordance value of `{r} ci`
-which indicates that it is more than 50% accurate in predicting the
-survival of the machines. However, this model is not completely accurate
-so there is an expectation of some error in the model survival
-probabilities which will lead to caution in providing results in
-application settings.
+The CPH model that was selected has a concordance value of
+`{r} round(ci, 2)`, which indicates that it is more than 50% accurate in
+predicting the survival of the machines. However, this model is not
+completely accurate, so there is an expectation of some error in the
+model survival probabilities which will lead to caution in providing
+results in application settings.
 
@@
 The coefficients above can be used to calculate the hazard rate
-[@eq-hazardrate], the cumulative hazard value[@eq-cumulativehr], and the
-survival probability [@eq-survprob]. The hazard rate at time $t$ will be
-calculated for each engine at the engines max time value. The cumulative
-hazard value is the total hazard accumulated for the subject over time
-and is used to estimate the risk for the subject at a point in time. The
-engines that have higher hazard rates or higher cumulative hazard values
-are at a greater risk of experiencing failure. The below table provides
-the hazard rate and survival probability for the first 10 engines in the
-dataset.
+([@eq-hazardrate]), the cumulative hazard value ([@eq-cumulativehr]),
+and the survival probability ([@eq-survprob]). The hazard rate at time
+$t$ will be calculated for each engine at the engine's max time value.
+The cumulative hazard value is the total hazard accumulated for the
+subject over time and is used to estimate the risk for the subject at a
+point in time. The engines that have higher hazard rates or higher
+cumulative hazard values are at a greater risk of experiencing failure.
+The below table provides the hazard rate and survival probability for
+the first 10 engines in the dataset.
 
@@
 prediction of specific failure times, it does allow for the creation of
-confidence intervals which can be used to gain an understanding of when
+confidence intervals, which can be used to gain an understanding of when
 engines are most likely to fail. The table below displays the percent of
@@
 The lower end of the confidence interval for survival probability is
-shown to hold stable at 100% until `t = 128` with a gradual decrease
-afterwards. Half of the engines are expected to survive until `t = 245`
+shown to hold stable at 100% until `t = 128`, with a gradual decrease
+afterwards. Half of the engines are expected to survive until `t = 245`,
 and 90% of the engines are expected to fail at `t = 347`. These results
@@
 
-Based on these findings, the model can be applied by recommending for
-NASA not to use engines after 128 iterations have passed. The engines
+Based on these findings, the model can be applied by recommending that
+NASA not use engines after 128 iterations have passed. The engines
 should undergo routine maintenance beforehand and should be evaluated to
@@
 100%. All engines should be equipped with sensors for the covariates
-used and readings should be applied to the model after each use of the
+used, and readings should be applied to the model after each use of the
 engine. The engine should be removed from service if the resulting

### 2025-04-22 slides.qmd
Slide updates 4/22

@@
 
--   Survival Function: $S(x) = \int_{x}^{\infty} f(x)dx$
-
--   Hazard Function:
-    $h(x) = \lim\limits_{\Delta x \rightarrow 0} \frac{P[x \leq X < x + \Delta x|X \geq x]}{\Delta x}$
+-   Concordance Index: $C = \frac{c + \frac{t_x}{2}}{c + d + t_x}$
 
@@
 -   Cumulative Hazard Function: $H(x) = \int_0^x h(u) du$
-
--   Concordance Index: $C = \frac{c + \frac{t_x}{2}}{c + d + t_x}$
 
@@
 
--   The data was then merged together to make one grouped dataset
-
 ```{r}
@@
 
-kable(head(temp, n = 5), format = "markdown", caption = 'NASA Aircraft Engine Data')
+kable(head(temp, n = 3), format = "markdown", caption = 'NASA Aircraft Engine Data')
 ```
@@
 
-autoplot(km_fit) + theme_bw()
+autoplot(km_fit) + theme_bw() + labs(y = "Survival Probability (%)", x = 'Time (Iterations)')
 ```
@@
   theme_bw()+
-  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
-  ylab(label = "Covariates")+
-  xlab(label = "Estimated Effect on Hazard Ratio")
+  labs(title = "Estimated Effects of Covariates on Hazard Ratios",
+       y = "Estimated Effect on Hazard Ratio",
+       x = "Covariates")
 ```
@@
 
-par(mfrow = c(4,4),
+par(mfrow = c(4,6),
       mar = c(2,2,4,1))
 
-plot(data_grouped$os1, martingale, main = "Martingale Residuals: os1")
+plot(data_grouped$os1, martingale, main = "Martingale: os1")
 abline(h = 0, col = "red")
 
-plot(data_grouped$os2, martingale, main = "Martingale Residuals: os2")
+plot(data_grouped$os2, martingale, main = "Martingale: os2")
 abline(h = 0, col = "red")
 
-plot(data_grouped$os3, martingale, main = "Martingale Residuals: os3")
+plot(data_grouped$os3, martingale, main = "Martingale: os3")
 abline(h = 0, col = "red")
 
-plot(data_grouped$sm1, martingale, main = "Martingale Residuals: sm1")
+plot(data_grouped$sm1, martingale, main = "Martingale: sm1")
 abline(h = 0, col = "red")
 
-plot(data_grouped$sm2, martingale, main = "Martingale Residuals: sm2")
+plot(data_grouped$sm2, martingale, main = "Martingale: sm2")
 abline(h = 0, col = "red")
 
-plot(data_grouped$sm3, martingale, main = "Martingale Residuals: sm3")
+plot(data_grouped$sm3, martingale, main = "Martingale: sm3")
 abline(h = 0, col = "red")
 
-plot(data_grouped$sm4, martingale, main = "Martingale Residuals: sm4")
+plot(data_grouped$sm4, martingale, main = "Martingale: sm4")
 abline(h = 0, col = "red")
 
-plot(data_grouped$sm5, martingale, main = "Martingale Residuals: sm5")
+plot(data_grouped$sm5, martingale, main = "Martingale: sm5")
 abline(h = 0, col = "red")
 
-plot(data_grouped$sm6, martingale, main = "Martingale Residuals: sm6")
+plot(data_grouped$sm6, martingale, main = "Martingale: sm6")
 abline(h = 0, col = "red")
 
-plot(data_grouped$sm7, martingale, main = "Martingale Residuals: sm7")
+plot(data_grouped$sm7, martingale, main = "Martingale: sm7")
 abline(h = 0, col = "red")
 
-plot(data_grouped$sm8, martingale, main = "Martingale Residuals: sm8")
+plot(data_grouped$sm8, martingale, main = "Martingale: sm8")
 abline(h = 0, col = "red")
 
-plot(data_grouped$sm9, martingale, main = "Martingale Residuals: sm9")
+plot(data_grouped$sm9, martingale, main = "Martingale: sm9")
 abline(h = 0, col = "red")
 
-plot(data_grouped$sm10, martingale, main = "Martingale Residuals: sm10")
+plot(data_grouped$sm10, martingale, main = "Martingale: sm10")
 abline(h = 0, col = "red")
 
-plot(data_grouped$sm11, martingale, main = "Martingale Residuals: sm11")
+plot(data_grouped$sm11, martingale, main = "Martingale: sm11")
 abline(h = 0, col = "red")
 
-plot(data_grouped$sm12, martingale, main = "Martingale Residuals: sm12")
+plot(data_grouped$sm12, martingale, main = "Martingale: sm12")
 abline(h = 0, col = "red")
 
-plot(data_grouped$sm13, martingale, main = "Martingale Residuals: sm13")
+plot(data_grouped$sm13, martingale, main = "Martingale: sm13")
 abline(h = 0, col = "red")
 
-plot(data_grouped$sm14, martingale, main = "Martingale Residuals: sm14")
+plot(data_grouped$sm14, martingale, main = "Martingale: sm14")
 abline(h = 0, col = "red")
 
-plot(data_grouped$sm15, martingale, main = "Martingale Residuals: sm15")
+plot(data_grouped$sm15, martingale, main = "Martingale: sm15")
 abline(h = 0, col = "red")
 
-plot(data_grouped$sm17, martingale, main = "Martingale Residuals: sm17")
+plot(data_grouped$sm17, martingale, main = "Martingale: sm17")
 abline(h = 0, col = "red")
 
-plot(data_grouped$sm18, martingale, main = "Martingale Residuals: sm18")
+plot(data_grouped$sm18, martingale, main = "Martingale: sm18")
 abline(h = 0, col = "red")
 
-plot(data_grouped$sm20, martingale, main = "Martingale Residuals: sm20")
+plot(data_grouped$sm20, martingale, main = "Martingale: sm20")
 abline(h = 0, col = "red")
 
-plot(data_grouped$sm21, martingale, main = "Martingale Residuals: sm21")
+plot(data_grouped$sm21, martingale, main = "Martingale: sm21")
 abline(h = 0, col = "red")

### 2025-04-22 slides.qmd
Final slides update 4/22

@@
 
-::: incremental
 -   A statistical regression method specializing in modeling
@@
     insurance likelihood payouts
-:::
 
@@
 
-::: incremental
 -   The model assumes that as time goes on, the survival probability
@@
     happen, only the probability of when the event might happen
-:::
 
@@
 
-::: incremental
 -   Independence assumption
@@
         constant at all times [@Bustan2018]
-:::
 
@@
 
-::: incremental
 -   Model accuracy evaluation is done using the concordance index
 
-    ::: incremental
     -   The concordance index measures the amount of agreement between
@@
         value of 0 means no pairs are correctly ordered
-    :::
 
@@
 
-    ::: incremental
     -   If the probability is $\geq 50$%, it is assumed the event has
@@
         occurred
-    :::
-:::
 
@@
 
-The table provides the model number, the covariates used, the AIC and
-BIC from the stepwise regression if applicable, and the concordance
-index.
+-   After an analysis of the covariates provided with the data, we
+    determined that sm16 and sm19 should not be included in the model
+
+    -   sm16 produces an infinite coefficient when taking the $log$
+
+    -   sm19 is linearly dependent on another covariate
+
+$$
+\begin{split}
+h(t) = h_0(t)\text{exp}(&-3.97\text{ os1}+26.06\text{ os2}+121.38\text{ os3}+30.47\text{ sm1} \\ &+0.06\text{ sm2}+0.04\text{ sm3}+0.04\text{ sm4} + 48.62\text{ sm5}\\ &-39.93\text{ sm6}-0.10\text{ sm7}+2.26\text{ sm8}-0.03\text{ sm9} \\ &+15.22\text{ sm10} +0.43\text{ sm11}+0.01\text{ sm12}-1.38\text{ sm13}\\ &+0.02\text{ sm14}+8.75\text{ sm15}+0.11\text{ sm17} -14.86\text{ sm18} \\ &-1.02\text{ sm20} -0.34\text{ sm21})
+\end{split}
+$$
 
@@
 
-all_cov <- coxph(data = data_grouped, surv_object ~ os1 + os2 + os3 + sm1 + sm2 + sm3 + sm4 + sm5 + sm6 + sm7 + sm8 + sm9 + sm10 + sm11 + sm12 + sm13 + sm14 + sm15 + sm16 + sm17 + sm18 + sm19 + sm20 + sm21)
+cox1.1 <- coxph(data = data_grouped, surv_object ~ os1 + os2 + os3 + sm1 + sm2 + sm3 + sm4 + sm5 + sm6 + sm7 + sm8 + sm9 + sm10 + sm11 + sm12 + sm13 + sm14 + sm15 + sm17 + sm18 + sm20 + sm21)
 
-aic <- MASS::stepAIC(all_cov, trace = 0)
-n <- nrow(data_grouped)
-bic <- step(all_cov, direction = "both", k = log(n), trace = 0)
-
-cox1 <- coxph(data = data_grouped, surv_object ~ os1 + os2 + os3 + sm1 + sm2 + sm3 + sm4 + sm5 + sm6 + sm7 + sm8 + sm9 + sm10 + sm11 + sm12 + sm13 + sm14 + sm15 + sm16 + sm17 + sm18 + sm19 + sm20 + sm21)
-
-cox2 <- coxph(data = data_grouped, surv_object ~ os3 + sm3 + sm4 + sm5 + sm8 + sm9 + sm15 + sm16 + sm17 + sm18 + sm20)
-
-cox3 <- coxph(data = data_grouped, surv_object ~ os3 + sm3 + sm4 + sm8 + sm9 + sm15 + sm18)
-
-cox4 <- coxph(data = data_grouped, surv_object ~ os3 + sm13 + sm14 + sm19)
-
-stepwise_results <- data.frame(c("model 1", "model 2", "model 3", "model 4"), 
-                               c("All covariates", "os3, sm3 - sm5, sm8 - sm9, sm15 - sm18, sm20", "os3, sm3 - sm4, sm8-sm9, sm15, sm18", "os3, sm13 - sm14, sm19"),
-                               c("2449.92", "2431.88", "N/A", "N/A"),
-                               c("2547.71", "N/A", "2461.31", "N/A"),
-                               c(cox1$concordance[6], cox2$concordance[6], cox3$concordance[6], cox4$concordance[6]))
-
-colnames(stepwise_results) <- c("Model", "Covariates", "AIC", "BIC", "Concordance")
-
-kable(stepwise_results, format = "markdown")
 ```
-
-Model for the continuing analysis will have all covariates except sm16
-and sm19.
 
@@
     determining if there is a slope of zero
-
--   Based on the below plots, this assumption is met.
 
@@
     of the covariate at the time of failure) and examining the p-value
-
--   Based on the below table and plots, this assumption is met
 

### 2025-04-27 index.qmd
updates

@@
            xlab = "Time",
-           ylab = "Survival Probability",
+           ylab = "Survival Probability (%)",
            legend = "none")
@@
   labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
-  ylab(label = "Covariates")+
-  xlab(label = "Estimated Effect on Hazard Ratio")
+  xlab(label = "Covariates")+
+  ylab(label = "Estimated Effect on Hazard Ratio")
 
@@
 
-autoplot(km_fit) + theme_bw()
+autoplot(km_fit) + theme_bw() + labs(y = "Survival Probability (%)", x = 'Time (Iterations)', title = "Kaplan-Meier Survival Curve")
 ```
@@
   theme_bw()+
-  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
-  ylab(label = "Covariates")+
-  xlab(label = "Estimated Effect on Hazard Ratio")
+  labs(title = "Estimated Effects of Covariates on Hazard Ratio")+
+  xlab(label = "Covariates")+
+  ylab(label = "Estimated Effect on Hazard Ratio")
 ```
@@
 ```{r, echo = FALSE}
-par(mfrow = c(4,4),
+par(mfrow = c(4,6),
       mar = c(2,2,4,1))
 
-plot(data_grouped$os1, martingale, main = "Martingale Residuals: os1")
-abline(h = 0, col = "red")
-
-plot(data_grouped$os2, martingale, main = "Martingale Residuals: os2")
-abline(h = 0, col = "red")
-
-plot(data_grouped$os3, martingale, main = "Martingale Residuals: os3")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm1, martingale, main = "Martingale Residuals: sm1")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm2, martingale, main = "Martingale Residuals: sm2")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm3, martingale, main = "Martingale Residuals: sm3")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm4, martingale, main = "Martingale Residuals: sm4")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm5, martingale, main = "Martingale Residuals: sm5")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm6, martingale, main = "Martingale Residuals: sm6")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm7, martingale, main = "Martingale Residuals: sm7")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm8, martingale, main = "Martingale Residuals: sm8")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm9, martingale, main = "Martingale Residuals: sm9")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm10, martingale, main = "Martingale Residuals: sm10")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm11, martingale, main = "Martingale Residuals: sm11")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm12, martingale, main = "Martingale Residuals: sm12")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm13, martingale, main = "Martingale Residuals: sm13")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm14, martingale, main = "Martingale Residuals: sm14")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm15, martingale, main = "Martingale Residuals: sm15")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm17, martingale, main = "Martingale Residuals: sm17")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm18, martingale, main = "Martingale Residuals: sm18")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm20, martingale, main = "Martingale Residuals: sm20")
-abline(h = 0, col = "red")
-
-plot(data_grouped$sm21, martingale, main = "Martingale Residuals: sm21")
+plot(data_grouped$os1, martingale, main = "Martingale: os1")
+abline(h = 0, col = "red")
+
+plot(data_grouped$os2, martingale, main = "Martingale: os2")
+abline(h = 0, col = "red")
+
+plot(data_grouped$os3, martingale, main = "Martingale: os3")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm1, martingale, main = "Martingale: sm1")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm2, martingale, main = "Martingale: sm2")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm3, martingale, main = "Martingale: sm3")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm4, martingale, main = "Martingale: sm4")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm5, martingale, main = "Martingale: sm5")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm6, martingale, main = "Martingale: sm6")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm7, martingale, main = "Martingale: sm7")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm8, martingale, main = "Martingale: sm8")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm9, martingale, main = "Martingale: sm9")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm10, martingale, main = "Martingale: sm10")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm11, martingale, main = "Martingale: sm11")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm12, martingale, main = "Martingale: sm12")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm13, martingale, main = "Martingale: sm13")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm14, martingale, main = "Martingale: sm14")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm15, martingale, main = "Martingale: sm15")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm17, martingale, main = "Martingale: sm17")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm18, martingale, main = "Martingale: sm18")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm20, martingale, main = "Martingale: sm20")
+abline(h = 0, col = "red")
+
+plot(data_grouped$sm21, martingale, main = "Martingale: sm21")
 abline(h = 0, col = "red")
@@
 
-create_ph_plots(cox1.1, 3, 5)
+create_ph_plots(cox1.1, 4, 6)
 ```
@@
            data = data_grouped_reduced, risk.table = TRUE, fun = "cumhaz")$plot +
-  ggtitle('Cumulative Hazard Over Time')
+  ggtitle('Cumulative Hazard Over Time')+
+  xlab("Time (Iterations)")
 
@@
 ggsurvplot(fit = sf, data = data_grouped_reduced)$plot +
-  ggtitle('Survival Probability Over Time')
+  ggtitle('Survival Probability Over Time')+
+  xlab('Timee (Iterations)')+
+  ylab('Survival Probability (%)')
 
@@
 plot which shows a low hazard until `t = 128` with a gradual increase.
+The model's final concordance index of .696 indicates that the calculated confidence interval is fairly accurate, but it is still subject to a substantial margin for error.
 
@@
 
+While the concordance index of .696 achieved in this analysis demonstrates the model is fairly accurate, there is still room for improvement.
+Survival models resulting in a concordance index of .9 and higher are highly accurate and would provide a tighter, more accurate confidence interval for the longevity of engines.
+After attempting transformations of the data and creating models many different combinations of covariates, the final c-index that was calculated is the highest value achieved where the covariates were able to meet the proportional hazards model.
+
+As discovered during the literature review, a weakness of the cox proportional hazards model is the necessity to meet the proportional hazards assumption.
+Due to the difficulty meeting the assumption during this analysis, to achieve a more accurate model it would be beneficial to explore other survival models for this dataset.
+
 ## References

### 2025-04-29 index.qmd
Final updates 4/29

@@
 library(broom)
+library(glmnet)
 
@@
 industry, but the usage of the model does not come without challenges. A
-COVID-19 study done using data from a Pakistani hospital noted that the
-model was limited due to the recovery possibility of COVID-19 patients,
-as CPH assumes as time goes on, the survival probability will approach
-zero with no survivors. Though CPH was not the best model for this
-study, the results of the model were that individuals who were
-asymptomatic and young had a higher chance of surviving the virus
-[@asghar2024]. In 2024, a study was done in the US on the relationship
-between chest pain occurring and mortality of the person experiencing
-chest pain using CPH. A response to that study demonstrated the various
-limitations of CPH, with the primary concern being the constant hazards
-ratio assumption that CPH relies on. In the heart study, the effect of
-triglyceride-glucose was assumed to be a constant, but it is actually a
-value that fluctuates over time along with an individual's body weight
-and blood glucose levels. As such, the model would struggle to predict
-correctly the effect of triglyceride-glucose given the violation of the
-constant hazards ratio [@jiang2024].
-
-An additional limitation of CPH comes from the data in the covariates
-selected and how they can become biased. Two studies done using UK
-health data demonstrated how bias can impact the values for the
-covariates when relying on patient-submitted information. The first was
-a study on the connection between salt intake and the development of
-anxiety or depression after 14.5 years [@wang2025], and the second was a
-study on the association between postpartum depression and the
-development of two or more chronic diseases [@zhang2025]. Both of those
-studies relied on patient-submitted health data and survey responses
-supplemented with death records and other accessible health records.
-This created a possible bias in their data that may have negatively
-influenced their model as their collected data may not accurately
-reflect the true data. These articles provide examples of the care that
-needs to be taken when working with CPH models.
+COVID-19 study using data from a Pakistani hospital noted that the model
+was limited due to the recovery possibility of COVID-19 patients. CPH
+assumes as time goes on, the survival probability will approach zero
+with no survivors, which was not the case with COVID-19. Though CPH was
+not the best model for this study, the results of the model were that
+individuals who were asymptomatic and young had a higher chance of
+surviving the virus [@asghar2024]. In 2024, a study was done in the US
+on the relationship between chest pain occurring and mortality of the
+person experiencing chest pain using CPH. A response to that study
+demonstrated the various limitations of CPH, with the primary concern
+being the constant hazards ratio assumption that CPH relies on. In the
+heart study, the effect of triglyceride-glucose was assumed to be a
+constant, but it is actually a value that fluctuates over time along
+with an individual's body weight and blood glucose levels. As such, the
+model would struggle to predict correctly the effect of
+triglyceride-glucose given the violation of the constant hazards ratio
+[@jiang2024].
+
+An additional limitation of CPH comes from the data as the selection of
+covariates and the generation of the data can be biased. Two studies
+done using UK health data demonstrated how bias can impact the values
+for the covariates when relying on patient-submitted information. The
+first was a study on the connection between salt intake and the
+development of anxiety or depression after 14.5 years [@wang2025], and
+the second was a study on the association between postpartum depression
+and the development of two or more chronic diseases [@zhang2025]. Both
+of those studies relied on patient-submitted health data and survey
+responses supplemented with death records and other accessible health
+records. The reliance This created a possible bias in their data that
+may have negatively influenced their model as their collected data may
+not accurately reflect the true data. These articles provide examples of
+the care that needs to be taken when working with CPH models.
 
@@
 impacts the length of productive life [@ziadi2023]. Both of these
-studies provide examples of the reaches of CPH models outside the health
-industry.
+studies provide examples of CPH models outside the health industry.
 
@@
 
-### Covariate Selection
-
-Depending on the study, there will be covariates that do not impact the
-probability of survival. Statistical methods such as the Wald test can
-be used to determine which ones are statistically significant
-individually and may have an impact in a multivariate model. Generating
-a univariate CPH model for each covariate and using the Wald test with a
-significance level of $\alpha = 0.05$ will allow for determination of if
-a covariate is significant. Additional tests for models will involve
-using a stepwise regression and comparing the results based on the
-assumptions that are met and the concordance index. Any covariates that
-are determined to be not significant to the survival probability can be
-removed from the final CPH model.
-
 ### Assumptions
@@
 impact the model individually. The covariates in which the p-value is
-$< 0.05$ will be considered statistically significant.
+$< 0.05$ will be considered statistically significant. The table
+displays the covariate, the $\beta$ value, the hazard ratio (HR), the
+95% confidence interval for the hazard ratio, and the p-value generated
+from the Wald test. As can be seen, none of the covariates are
+statistically signification on their own.
 
 ```{r}
+surv_object <- Surv(data_grouped$time, data_grouped$status)
+
 covariates <- c("os1", "os2", "os3", "sm1", "sm2", "sm3", "sm4", "sm5", "sm6", "sm7",
@@
 
-univ_formulas <- sapply(covariates,
+forms <- sapply(covariates,
                         function(x) as.formula(paste('Surv(time, status)~', x)))
 
-univ_models <- lapply( univ_formulas, function(x){coxph(x, data = data_grouped)})
+univ_models <- lapply(forms, function(x){coxph(x, data = data_grouped)})
 
@@
                           p.value<-signif(x$wald["pvalue"], digits=2)
-                          wald.test<-signif(x$wald["test"], digits=2)
                           beta<-signif(x$coef[1], digits=2);#coefficient beta
@@
                           HR.confint.upper <- signif(x$conf.int[,"upper .95"],2)
-                          HR <- paste0(HR, " (",
-                                       HR.confint.lower, "-", HR.confint.upper, ")")
-                          res<-c(beta, HR, wald.test, p.value)
-                          names(res)<-c("beta", "HR (95% CI for HR)", "wald.test",
-                                        "p.value")
+                          CI <- paste0("(", HR.confint.lower, "-", HR.confint.upper, ")")
+                          res<-c(beta, HR, CI, p.value)
+                          names(res)<-c("beta", "HR", "95% CI", 
+                                        "p-value")
                           return(res)
@@
 
-Based on the above test, none of the covariates are statistically
-significant individually with all the data. As such, to determine the
-best model to use for this analysis, a forward and backward stepwise
-regression will be done to generate the models with the best Akaike
-Information Criterion (AIC) and Bayesian Information Criterion (BIC)
-values. This method produced three possible models. The fourth model in
-the below table used covariates that were determined to significant with
-the training data only.
+To determine the best model to use for this analysis, a forward and
+backward stepwise regression was done to generate the models with the
+best Akaike Information Criterion (AIC) and Bayesian Information
+Criterion (BIC) values. The second and third models were generated using
+this regression method. The fourth model was generated using the Lasso
+penalized Cox regression method.
 
@@
 and BIC from the stepwise regression if applicable, and the concordance
-index. After determining the most accurate model, the CPH assumptions
-will be tested.
+index.
 
 ```{r}
-surv_object <- Surv(data_grouped$time, data_grouped$status)
-
 all_cov <- coxph(data = data_grouped, surv_object ~ os1 + os2 + os3 + sm1 + sm2 + sm3 + sm4 + sm5 + sm6 + sm7 + sm8 + sm9 + sm10 + sm11 + sm12 + sm13 + sm14 + sm15 + sm16 + sm17 + sm18 + sm19 + sm20 + sm21)
 
+#stepwise regression models
 aic <- MASS::stepAIC(all_cov, trace = 0)
@@
 
+#input/out for glmnet
+all_mod <- model.matrix(Surv(time, status) ~  + os2 + os3 + sm1 + sm2 + sm3 + sm4 + sm5 + sm6 + sm7 + sm8 + sm9 + sm10 + sm11 + sm12 + sm13 + sm14 + sm15 + sm16 + sm17 + sm18 + sm19 + sm20 + sm21, data = data_grouped)[, -1]
+
+# lasso penalized cox
+cvfit <- cv.glmnet(all_mod, surv_object, family = "cox")
+coefs <- coef(cvfit, s = "lambda.min")
+new_cov <- rownames(coefs)[which(coefs != 0)]
+
+#all Cox models 
+#all covariates
 cox1 <- coxph(data = data_grouped, surv_object ~ os1 + os2 + os3 + sm1 + sm2 + sm3 + sm4 + sm5 + sm6 + sm7 + sm8 + sm9 + sm10 + sm11 + sm12 + sm13 + sm14 + sm15 + sm16 + sm17 + sm18 + sm19 + sm20 + sm21)
 
+#aic
 cox2 <- coxph(data = data_grouped, surv_object ~ os3 + sm3 + sm4 + sm5 + sm8 + sm9 + sm15 + sm16 + sm17 + sm18 + sm20)
 
+#bic
 cox3 <- coxph(data = data_grouped, surv_object ~ os3 + sm3 + sm4 + sm8 + sm9 + sm15 + sm18)
 
-cox4 <- coxph(data = data_grouped, surv_object ~ os3 + sm13 + sm14 + sm19)
+#lasso
+cox4 <- coxph(data = data_grouped, surv_object ~ os2 + sm1 + sm3 + sm4 + sm9 + sm10 + sm11 + sm14 + sm15 + sm16 + sm17 + sm18)
 
@@
 stepwise_results <- data.frame(c("model 1", "model 2", "model 3", "model 4"), 
-                               c("All covariates", "os3, sm3 - sm5, sm8 - sm9, sm15 - sm18, sm20", "os3, sm3 - sm4, sm8-sm9, sm15, sm18", "os3, sm13 - sm14, sm19"),
+                               c("All covariates", "os3, sm3, sm4, sm5, sm8, sm9, sm15, sm16, sm17, sm18, sm20", "os3, sm3, sm4, sm8, sm9, sm15, sm18", "os2, sm1, sm3, sm4, sm9, sm10, sm11, sm14, sm15, sm16, sm17, sm18"),
                                c("2449.92", "2431.88", "N/A", "N/A"),
@@
 
-In determining the best model for this data, it was determined that CPH
-is not the survival method that should be used on this data. By stepwise
-regression, the models with the lowest BIC and AIC should have been used
-in the following analysis; however, both models continuously failed the
-CPH assumptions. As such, the model for the following analysis will be
-model 1, which contained all covariates and had the highest concordance
-value.
-
-While model 1 does have initially contain all 24 covariates, there are
+All four models were checked for the CPH assumptions. All but the first
+model failed the assumptions, even after adapting the covariate using
+stratification and/or removal. As such, the first model was the one
+selected for the continuing analyses.
+
+While model one does have initially contain all 24 covariates, there are
 two that will need to be removed before proceeding with the analysis due
@@
 a possible infinite coefficient. The following analysis will be done
-using model 1 without sm16 and sm19. The below tables provide the
+using model one without sm16 and sm19. The below tables provide the
 summary values from the model.
@@
 
-The forest plot for model 1 is shown below. The covariates that are
+The forest plot for model one is shown below. The covariates that are
 closest to the value of one have a hazard rate that has less effect on
@@
 plot which shows a low hazard until `t = 128` with a gradual increase.
-The model's final concordance index of .696 indicates that the calculated confidence interval is fairly accurate, but it is still subject to a substantial margin for error.
+The model's final concordance index of .696 indicates that the
+calculated confidence interval is fairly accurate, but it is still
+subject to a substantial margin for error.
 
@@
 
-While the concordance index of .696 achieved in this analysis demonstrates the model is fairly accurate, there is still room for improvement.
-Survival models resulting in a concordance index of .9 and higher are highly accurate and would provide a tighter, more accurate confidence interval for the longevity of engines.
-After attempting transformations of the data and creating models many different combinations of covariates, the final c-index that was calculated is the highest value achieved where the covariates were able to meet the proportional hazards model.
-
-As discovered during the literature review, a weakness of the cox proportional hazards model is the necessity to meet the proportional hazards assumption.
-Due to the difficulty meeting the assumption during this analysis, to achieve a more accurate model it would be beneficial to explore other survival models for this dataset.
+While the concordance index of .696 achieved in this analysis
+demonstrates the model is fairly accurate, there is still room for
+improvement. Survival models resulting in a concordance index of .9 or
+higher are highly accurate and would provide a tighter, more accurate
+confidence interval for the longevity of engines. After attempting
+transformations of the data and creating models with many different
+combinations of covariates, the final c-index that was calculated is the
+highest value achieved where the covariates were able to meet the
+proportional hazards model.
+
+As discussed during earlier, a weakness of the Cox proportional hazards
+model is the necessity to meet the proportional hazards assumption.
+Given the difficulty of meeting the PH assumption, a different survival
+analysis would have better suited this dataset.
 
